### 2016-09-08 Project1/article/article.tex
Added project1 folder

\documentclass[10pt, a4paper]{amsart}

\usepackage[]{graphicx}
\usepackage[]{hyperref}

\usepackage[T1]{fontenc}

\title{Solving the Poisson-equation in one dimension. \\
  \hrulefill\large{FYS3150: Computational Physics}\hrulefill}

\author[Sebastian G. Winther-Larsen]{Sebastian G. Winther-Larsen \\
  \href{https://github.com/gregwinther/FYS3150/Project1}{\texttt{github.com/gregwinther}}}

\begin{document}

\begin{titlepage}
\begin{abstract}
Lorem ipsum dolor sit amet, consectur adipiscing lit. Nullam ut lacus eget lorem...
\end{abstract}
\maketitle
\tableofcontents
\end{titlepage}

\end{document}
ipsum lorem

### 2016-09-11 Project1/article/article.tex
Expanded directory structure
Wrote out most of the theory

@@
 \usepackage[]{hyperref}
+\usepackage[]{physics}
 
@@
 
-\title{Solving the Poisson-equation in one dimension. \\
-  \hrulefill\large{FYS3150: Computational Physics}\hrulefill}
+\title[Solving the Poisson-equation in one dimension]{Solving the Poisson-equation in one dimension \\
+  \hrulefill\small{ FYS3150: Computational Physics }\hrulefill}
 
 \author[Sebastian G. Winther-Larsen]{Sebastian G. Winther-Larsen \\
-  \href{https://github.com/gregwinther/FYS3150/Project1}{\texttt{github.com/gregwinther}}}
+  \href{https://github.com/gregwinther/FYS3150/}{\texttt{github.com/gregwinther}}}
 
@@
 
+\section{Introduction}
+
+\section{Theory}
+\subsection{The Poisson Equation}
+
+The Poisson equation is a classical equation from electromagnetism.
+The electrostatic potential $\Phi$ is generated by a localized charge distribution $\rho(\vb{r})$.
+In three dimensions the equation reads
+\begin{equation}
+\laplacian\Phi=-4\pi \rho(\vb{r})
+\end{equation}
+where $\laplacian$ is the Laplace operator.
+In three dimensions the Laplace operator can be expressed using spherical coordinates, but in this study I am assuming that $\Phi$ and $\rho$ are spherically symmetric, thus reducing the equation to a one-dimensional problem.
+only dependent on radius $r$.
+\begin{equation}
+\laplacian=\frac{1}{r^2}\frac{d}{dr}\left(r^2\frac{d\Phi}{dr} \right)
+\end{equation}
+By substituting $\Phi(r)=\phi(r)/r$ the Poisson equation is reduced to 
+\begin{equation}
+\frac{d^2\phi}{dr^2}=-4\pi r\rho(r)
+\end{equation}
+and by letting $\phi \rightarrow u$ and $r \rightarrow x$ one is left with the very simple equation
+\begin{equation}
+-u''(x)=f(x)
+\end{equation}
+
+The inhomogenous term $f$, or source term, is given by the charge distribution $\rho$ multiplied by $r$ and the constant $-4\pi$.
+In this study, however, the source term will be $f(x)=100e^{-10x}$ and the results can be compared to the analytical solution $u(x)=1-(1-e^{-10})x-e^{-10x}$.
+
+\subsection{Approximation of the Second Derivative}
+In this study the one-dimensional Poisson equation will be solved with Dirichlet boundary conditions by rewrting
+
 \end{document}
-ipsum lorem

### 2016-09-11 Project1/article/article.tex
Theory part almost done

@@
 \begin{equation}
--u''(x)=f(x)
+-u''(x)=f(x) \label{eq:2nd}
 \end{equation}
@@
 \subsection{Approximation of the Second Derivative}
-In this study the one-dimensional Poisson equation will be solved with Dirichlet boundary conditions by rewrting
+In this study the one-dimensional Poisson equation will be solved with Dirichlet boundary conditions by rewriting it as a set of linear equations.
+The discretized approximation of $u$ is defined as $v_i$ with grid points $x_i=ih$, step size of $h=\frac{1}{n+1}$, in the interval $x_0=0$ to $x_{n+1}=1$ and with boundary conditions $v_0=v_n+1=0$.
+The interior solution $v_i \forall i \in {1,...,n}$ is to be found.
+The second order derivative is approximated with the three point formula such that equation \ref{eq:2nd} becomes
+\begin{equation}
+-\frac{v_{i+1}-2v_i+v{i-1}}{h²}=f_i \label{eq:2approx}
+\end{equation}
+
+By defining $\tilde{\vb{b}}=h^2f_i$ one can rewrite equation \ref{eq:2approx} as $-v_{i+1}-2v_i+v{i-1}=h^2f_i$.
+If we ignore the end points, $i=0$ and $i=n+1$, this equation can be represented as a matrix equation.
+\begin{equation}
+\begin{bmatrix}
+2 & -1 & 0 & 0 & \cdots & 0 & 0 & 0 \\
+-1 & 2 & -1 & 0 & \cdots & 0 & 0 & 0 \\
+0 & -1 & 2 & -1 & \cdots & 0 & 0 & 0 \\ 
+& & \vdots &  & \ddots &  & \vdots & \\
+0 & 0 & 0 & 0 & \cdots & -1 & 2 & -1 \\
+0 & 0 & 0 & 0 & \cdots & 0 & -1 & 2 
+\end{bmatrix}
+\begin{bmatrix}
+v_1 \\
+v_2 \\
+v_3 \\
+\vdots \\
+v_{n-1} \\
+v_n
+\end{bmatrix}=
+\begin{bmatrix}
+b_1 \\
+b_2 \\
+b_3 \\
+\vdots \\
+b_{n-1} \\
+b_n
+\end{bmatrix}
+\end{equation}
 

### 2016-09-13 Project1/article/article.tex
Working on algorithms in article

@@
 
-\author[Sebastian G. Winther-Larsen]{Sebastian G. Winther-Larsen \\
+\author[Svalheim \& Winther-Larsen]{Trygve Leithe Svalheim \\
+   Sebastian G. Winther-Larsen \\
   \href{https://github.com/gregwinther/FYS3150/}{\texttt{github.com/gregwinther}}}
@@
 
-By defining $\tilde{\vb{b}}=h^2f_i$ one can rewrite equation \ref{eq:2approx} as $-v_{i+1}-2v_i+v{i-1}=h^2f_i$.
-If we ignore the end points, $i=0$ and $i=n+1$, this equation can be represented as a matrix equation.
+By defining $\vb{f}=h^2f_i$ one can rewrite equation
+\ref{eq:2approx} as $-v_{i+1}-2v_i+v_{i-1}=h^2f_i$. If we ignore the
+end points, $i=0$ and $i=n+1$, this equation can be represented as a
+matrix equation.
+\begin{equation}
+A\vb{v}=\vb{f}
+\end{equation}
 \begin{equation}
@@
 \begin{bmatrix}
-b_1 \\
-b_2 \\
-b_3 \\
+f_1 \\
+f_2 \\
+f_3 \\
 \vdots \\
-b_{n-1} \\
-b_n
+f_{n-1} \\
+f_n
 \end{bmatrix}
@@
 
+\section{Algorithms}
+Two main methods are implemented and compared. The first method is
+gaussian elimination of the tridiagonal matrix $A$, also known as the
+\emph{Thomas Algorithm} \cite{thomasalgo}. This is a simplified form of Gaussian
+elimination that can ve used to solve tridiagonal systems of
+equations. The method is improved upon in order to take into account
+the fact that the matrix we are dealing with has the same numbers
+along the diagonals. The second method is the LU-decomposistion
+method.
+
+\subsection{Tridiagonal Matrix Algorithm}
+Our tridiagonal system can be represented by 
+\begin{equation}
+a_ix_{i-1} + b_ix_i + c_ix_{i+1} = \vb{f}
+\end{equation}
+with $a_i=-1,b_i=2$ and $c_i=-1$, except for $a_1=0$ and $c_n=0$. Row
+reducing a matrix will reveal how the algorithm functions. Limiting
+the problem to four dimensions for easier reading and to save the
+rainforest\footnote{Writing out a general case will take up more paper
+space}.
+
+\begin{equation}
+\left[
+\begin{array}{cccc|c}
+b_1 & c_1 & 0 & 0 & f_1 \\
+a_2 & b_2 & c_2 & 0  & f_2 \\
+0 & a_3 & b_3 & c_3 & f_3 \\
+0 & 0 & a_4 & b_4 & f_4
+\end{array}
+\right] \sim
+\left[
+\begin{array}{cccc|c}
+1 & c_1/b_1 & 0 & 0 & b_1/b_1 \\
+0 & b_2-\frac{c_1}{b_1}a_2 & c_2 & 0  & f_2-\frac{f_1}{b_1}a_2 \\
+0 & a_3 & b_3 & c_3 & f_3 \\
+0 & 0 & a_4 & b_4 & f_4
+\end{array}
+\right]
+\end{equation}
+Now let $\beta_1=b_1$ and $\beta_2=b_2-\frac{c_1}{b_1}a_2$. As new
+elements start to appear in vector $\vb{f}$, right to the vertical bar
+in the augmented matrix, they are also relabeled to $\tilde{f}_i$. For
+example $\tilde{f}_1=f_1/\beta_1$. One more iteration will reveal the
+pattern of the algorithm.
+
+\begin{align}
+\left[
+\begin{array}{cccc|c}
+1 & c_1/\beta_1 & 0 & 0 & \tilde{f}_1 \\
+0 & 1 & c_2/\beta_2 & 0  & (f_2-\tilde{f}a_2)/\beta_2 \\
+0 & a_3 & b_3 & c_3 & f_3 \\
+0 & 0 & a_4 & b_4 & f_4
+\end{array}
+\right] &=
+\left[
+\begin{array}{cccc|c}
+1 & c_1/\beta_1 & 0 & 0 & \tilde{f}_1 \\
+0 & 1 & c_2/\beta_2 & 0  & \tilde{f}_2 \\
+0 & a_3 & b_3 & c_3 & f_3 \\
+0 & 0 & a_4 & b_4 & f_4
+\end{array}
+\right] \\ \sim 
+\left[
+\begin{array}{cccc|c}
+1 & c_1/\beta_1 & 0 & 0 & \tilde{f}_1 \\
+0 & 1 & c_2/\beta_2 & 0  & \tilde{f}_2 \\
+0 & 0 & b_3-\frac{c_2}{\beta_2}a_3 & c_3 & f_3-\tilde{f}_2a_3 \\
+0 & 0 & a_4 & b_4 & f_4
+\end{array}
+\right] &\sim 
+\left[
+\begin{array}{cccc|c}
+1 & c_1/\beta_1 & 0 & 0 & \tilde{f}_1 \\
+0 & 1 & c_2/\beta_2 & 0  & \tilde{f}_2 \\
+0 & 0 & 1 & c_3/\beta_3 & (f_3-\tilde{f}_2a_3)/\beta_3 \\
+0 & 0 & a_4 & b_4 & f_4
+\end{array}\right] \\ 
+\sim &\dots \sim
+\left[
+\begin{array}{cccc|c}
+1 & c_1/\beta_1 & 0 & 0 & \tilde{f}_1 \\
+0 & 1 & c_2/\beta_2 & 0  & \tilde{f}_2 \\
+0 & 0 & 1 & c_3/\beta_3 & \tilde{f}_3 \\
+0 & 0 & 0 & 1 & \tilde{f}_4
+\end{array}
+\right]
+\end{align}
+
+This pattern can be summarized quite elegantly by the following
+difference equations.
+\begin{equation}
+\beta_i=b_i-\frac{c_{i-1}}{\beta_{i-1}}, \quad
+\tilde{f}_i=(\beta_i-\tilde{f}_{i-1}a_i)/\beta_i, \quad i \in [2,n], \quad
+\end{equation}
+
+\begin{thebibliography}{10}
+  \bibitem{thomasalgo}{Thomas, L.H. (1949), \emph{Elliptic
+        Problems in Linear Differential Equations over a
+        Network}. Watson Sci. Comput. Lab Report, Columbia University,
+      New York.}
+\end{thebibliography}
+
 \end{document}

### 2016-09-13 Project1/article/article.tex
More theory in article

@@
 
-This pattern can be summarized quite elegantly by the following
-difference equations.
+This pattern can be summarized nicely by the following
+difference equations. This computation will from now on be referred to
+as the ``forward substitution''. 
 \begin{equation}
@@
 \end{equation}
+
+The Gaussian elimination is not fully complete until a ``backward
+substitution'' has been performed as well. This ensures that all
+eliments in the row reduced matrix are pivot elements.
+
+\begin{equation}
+\left[
+\begin{array}{cccc|c}
+1 & c_1/\beta_1 & 0 & 0 & \tilde{f}_1 \\
+0 & 1 & c_2/\beta_2 & 0  & \tilde{f}_2 \\
+0 & 0 & 1 & c_3/\beta_3 & \tilde{f}_3 \\
+0 & 0 & 0 & 1 & \tilde{f}_4
+\end{array}
+\right] \sim
+\left[
+\begin{array}{cccc|c}
+1 & c_1/\beta_1 & 0 & 0 & \tilde{f}_1 \\
+0 & 1 & c_2/\beta_2 & 0  & \tilde{f}_2 \\
+0 & 0 & 1 & 0 & \tilde{f}_3 -\frac{c_3}{\beta_3}\tilde{f}_4 \\
+0 & 0 & 0 & 1 & \tilde{f}_4
+\end{array}
+\right]
+\end{equation}
+
+The notation is updated again to $v_3 =\tilde{f}_3 -
+\frac{c_3}{\beta_3}\tilde{f}_4$. By continuing this iterative process
+all the was to the beginning of the matrix by backward substitution
+leaves us with a new recurence relation.
+\begin{equation}
+v_i=\tilde{f}_i - \frac{c_i}{\beta_i}\tilde{f}_{i+1}, \quad i \in \{n-1,n-2,\dots,1\}
+\end{equation}
+
+\subsection{LU Decomposition}
+
+\section{Results}
+
+\section{Discussion}
+
+\section{Conclusion}
 
@@
       New York.}
+    \bibitem{morten}{Hjorth-Jensen, M. (2016). \emph{Computational
+        Physics - Lecture Notes 2016}. University of Oslo}
 \end{thebibliography}

### 2016-09-14 Project1/main.cpp
Added better timer, LU solution and relative error calculation

@@ -5,10 +5,11 @@ using namespace std;
 
 int main() {
 
-    int n = pow(10,5);
+    int n = pow(10,4);
 
-    //problem1ab();
+    problem1ab(n);
     problem1c(n);
+    problem1e(n);
 
     return 0;
 }

### 2016-09-14 Project1/problems.cpp

@@ -1,5 +1,5 @@
 #include <iostream>
-//#include <vector>
+#include <vector>
 #include <iostream>
 #include <string>
 #include <fstream>
@@ -7,6 +7,14 @@
 #include <algorithm>
 #include <iomanip>
 #include <time.h>
+#include <new>
+#include <cstdio>
+#include <cstdlib>
+#include <cmath>
+#include <cstring>
+#include <chrono>
+
+#include <armadillo>
 
 using namespace std;
 ofstream ofile;
@@ -26,9 +34,6 @@ void problem1ab(int n)
      * Allocating n+2 spaces, the first and last index
      * are for the boundary conditions
      */
-
-    // Timer variables
-    clock_t start, finish;
 
     // Diagonals of matrix
     double *a_vec = new double[n+2];
@@ -51,7 +56,7 @@ void problem1ab(int n)
     double *f_tilde = new double[n+2];
 
     // Saving start time
-    start = clock();
+    auto start = chrono::high_resolution_clock::now();
 
     // Linspace
     for (int i=0; i<n+2; i++) {
@@ -61,7 +66,7 @@ void problem1ab(int n)
         c_vec[i] = -1;
     }
 
-    // Contruction 'matrix'
+    // Contructing 'matrix'
     for (int i=0; i<n+2; i++) {
         f[i] = h*h*input_func(x[i]);
         u[i] = analytic_solution(x[i]);
@@ -83,11 +88,11 @@ void problem1ab(int n)
     }
 
     // Register stop time
-    finish = clock();
+    auto finish = chrono::high_resolution_clock::now();
 
     // Printing time elapsed
-    cout << "Elapsed time: ";
-    cout << ((double) (finish - start) / CLOCKS_PER_SEC) << endl;
+    cout << "Elapsed time, tridiagonal Gaussian: ";
+    cout << chrono::duration_cast<chrono::nanoseconds>(finish-start).count()/pow(10,9) << " seconds" << endl;
 
     // Open file and write results to file:
     ofile.open("dataproj1ab.txt");
@@ -120,9 +125,6 @@ int problem1c(int n)
      * are for the boundary conditions
      */
 
-    //Declaring time variable
-    clock_t start, finish;
-
     // Diagonal of matrix
     double *b_vec = new double[n+2];
 
@@ -142,7 +144,7 @@ int problem1c(int n)
[…]

### 2016-09-14 Project1/problems.h

@@ -3,5 +3,6 @@
 
 void problem1ab(int n);
 void problem1c(int n);
+void problem1e(int n);
 
 #endif // PROBLEMS_H

### 2016-09-14 Project1/main.cpp
Removed unecessary library imports

@@ -1,8 +1,6 @@
 #include "problems.h"
 #include <math.h>
 
-using namespace std;
-
 int main() {
 
     int n = pow(10,4);

### 2016-09-14 Project1/problems.cpp

@@ -1,19 +1,9 @@
 #include <iostream>
 #include <vector>
-#include <iostream>
-#include <string>
 #include <fstream>
 #include <math.h>
-#include <algorithm>
 #include <iomanip>
-#include <time.h>
-#include <new>
-#include <cstdio>
-#include <cstdlib>
-#include <cmath>
-#include <cstring>
 #include <chrono>
-
 #include <armadillo>
 
 using namespace std;
@@ -181,7 +171,7 @@ int problem1c(int n)
     }
     //Finding maximum element in error array
     double max = eps[0];
-    for (int i; i<n; i++){
+    for (int i=0; i<n; i++){
         if(abs(eps[i])>abs(max)) max=eps[i];
     }
     //Printing max rel. error

### 2016-09-14 Project1/problems.cpp
Minor estethic changes

@@ -65,15 +65,14 @@ void problem1ab(int n)
     // GAUSSIAN ELIMINATION
     // Forward substitution
     f_tilde[1] = f[1];
-    for (int i=2;i<n+1;i++){
-       // Updating right hand side of matrix equation:
+    for (int i=2; i<n+1; i++){
        b_vec[i] = b_vec[i]-a_vec[i]*c_vec[i-1]/b_vec[i-1];
        f_tilde[i] = f[i] - a_vec[i]*f_tilde[i-1]/b_vec[i-1];
     }
 
     // Backward substition:
     v[n] = f_tilde[n]/b_vec[n];
-    for (int i=n-1;i>=1;i--){
+    for (int i=n-1; i>=1; i--){
         v[i] = (f_tilde[i]-c_vec[i]*v[i+1])/b_vec[i];
     }
 
@@ -147,7 +146,7 @@ int problem1c(int n)
     // GAUSSIAN ELIMINATION
     // Forward substitution
     f_tilde[1] = f[1];
-    for (int i=2;i<n+1;i++){
+    for (int i=2; i<n+1; i++){
        // Updating right hand side of matrix equation:
        b_vec[i] -= 1/b_vec[i-1];
        f_tilde[i] = f[i] +f_tilde[i-1]/b_vec[i-1];

### 2016-09-14 Project1/article/article.tex
Scrutiny of program. Algorithm part in article almost done.

@@
 \usepackage[]{physics}
+\usepackage[]{listings}
+\usepackage[T1]{fontenc}
+\usepackage{color}
 
-\usepackage[T1]{fontenc}
+\definecolor{mygreen}{rgb}{0,0.6,0}
+\definecolor{mygray}{rgb}{0.5,0.5,0.5}
+\definecolor{mymauve}{rgb}{0.58,0,0.82}
+
+\lstset{ %
+  backgroundcolor=\color{white},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}
+  basicstyle=\footnotesize,        % the size of the fonts that are used for the code
+  breakatwhitespace=false,         % sets if automatic breaks should only happen at whitespace
+  breaklines=true,                 % sets automatic line breaking
+  captionpos=b,                    % sets the caption-position to bottom
+  commentstyle=\color{mygreen},    % comment style
+  deletekeywords={...},            % if you want to delete keywords from the given language
+  escapeinside={\%*}{*)},          % if you want to add LaTeX within your code
+  extendedchars=true,              % lets you use non-ASCII characters; for 8-bits encodings only, does not work with UTF-8
+  frame=single,	                   % adds a frame around the code
+  keepspaces=true,                 % keeps spaces in text, useful for keeping indentation of code (possibly needs columns=flexible)
+  keywordstyle=\color{blue},       % keyword style
+  language=c++,                 % the language of the code
+  otherkeywords={*,...},           % if you want to add more keywords to the set
+  rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
+  showspaces=false,                % show spaces everywhere adding particular underscores; it overrides 'showstringspaces'
+  showstringspaces=false,          % underline spaces within strings only
+  showtabs=false,                  % show tabs within strings adding particular underscores
+  stepnumber=2,                    % the step between two line-numbers. If it's 1, each line will be numbered
+  stringstyle=\color{mymauve},     % string literal style
+  tabsize=2,	                   % sets default tabsize to 2 spaces
+}
 
@@
 \begin{equation}
+\label{eq:matrix}
 A\vb{v}=\vb{f}
@@
 the problem to four dimensions for easier reading and to save the
-rainforest\footnote{Writing out a general case will take up more paper
+rainforest\footnote{Writing out a general case will also take up more paper
 space}.
@@
 \begin{equation}
-\beta_i=b_i-\frac{c_{i-1}}{\beta_{i-1}}, \quad
-\tilde{f}_i=(\beta_i-\tilde{f}_{i-1}a_i)/\beta_i, \quad i \in [2,n], \quad
+\label{eq:forward1}
+\beta_i=b_i-\frac{a_ic_{i-1}}{\beta_{i-1}}, \quad
+\tilde{f}_i=f_i-a_i\tilde{f}_{i-1}/\beta_{i-1}, \quad i \in [2,n], \quad
 \end{equation}
+
+In c++, this part of the algorithm is implemented just below. Notice
+that $b_i$ is overwritten instead of initializing a seperate vector
+for $\beta$. 
+\lstinputlisting[language=c++, firstline=65, lastline=71]{../problems.cpp}
 
@@
 all the was to the beginning of the matrix by backward substitution
-leaves us with a new recurence relation.
+leaves us with a second recurrence relation.
 \begin{equation}
-v_i=\tilde{f}_i - \frac{c_i}{\beta_i}\tilde{f}_{i+1}, \quad i \in \{n-1,n-2,\dots,1\}
+\label{eq:backward1}
+v_i=(\tilde{f}_i - c_iv_{i+1})/\beta_i, \quad i \in \{n-1,n-2,\dots,1\}
 \end{equation}
 
+The backward substitution is implemented in C++ like so:
+\lstinputlisting[language=c++, firstline=73,
+lastline=77]{../problems.cpp}
+
+An ever-important aspect of any algorithm is how efficient the algorithm is. It
+is therefor worthwhile to consider how many floating point operations
+the algorithm requires\footnote{The number of $+$,$-$,$*$ and $/$
+  operations are counted}. For the forward substitution there are $6$
+FLOPS in each iteration and $n-1$ iterations, resulting in $6(n-1)$
+FLOPS for the entire loop. Additionally, $1(n-1)$ FLOPS are required
+for the backward substitution, adding up to 
+\begin{equation}
+N_{tridiag}=8(n-1)=\mathcal{O} (8n).
+\end{equation}
+
+\subsubsection{Optimizing the algorithm}
+
+Our case is special, because the elements in each of the diagonal arrays in the
+tridiagonal matrix, $A$, are equal. $a_i=-1,b_i=2$ and $c_i=-1$,
+except for $a_1=0$ and $c_n=0$. This is an advantage, because the
+algorithm can be adjusted, and the number of FLOPS consequently reduced.
+
+The algorithm for the forward substitution, in equation
+\ref{eq:forward1}, can be simplified to
+\begin{equation}
+\beta_i= b_i - 1/\beta_{i-1} \quad \tilde{f}_i=f_i-\tilde{f}_{i-1}/\beta_{i-1}
+\end{equation}
+and the backward substitution, in equation \ref{eq:backward1}, can be
+simplified to
+\begin{equation}
+v_i = (\tilde{f}_i+v_{i+1})/\beta_i
+\end{equation}
+
+This results in some new and sleeker C++ code
+\lstinputlisting[language=c++, firstline=146,
+lastline=158]{../problems.cpp}
+
+One can see that this new and improved algorithm only requires
+$6(n-1)$ FLOPS. In general, a Gaussian row reduction algorithm
+requires $\frac{2}{3}n^3+\mathcal{O}(n^2)$ FLOPS \cite{morten}, a
+great improvement!
+
 \subsection{LU Decomposition}
+
+The LU decompositon method is a method where the matrix $A$ in
+equation \ref{eq:matrix} ($A\vb{v}=\vb{f}$) can be rewritten as a product of two matrices
+$L$ and $U$, where $L$ is lower triangular and and $U$ is upper
+triangular. This method only works for non-singular, invertible
+matrices, which the matrix $A$ certainly is.
 

### 2016-09-14 Project1/main.cpp

@@ -3,11 +3,11 @@
 
 int main() {
 
-    int n = pow(10,4);
+    int n = pow(10,1);
 
     problem1ab(n);
     problem1c(n);
-    problem1e(n);
+    //problem1e(n);
 
     return 0;
 }

### 2016-09-14 Project1/problems.cpp

@@ -48,7 +48,7 @@ void problem1ab(int n)
     // Saving start time
     auto start = chrono::high_resolution_clock::now();
 
-    // Linspace
+    // 'linspace'
     for (int i=0; i<n+2; i++) {
         x[i] = i*h;
         a_vec[i] = -1;
@@ -147,15 +147,14 @@ int problem1c(int n)
     // Forward substitution
     f_tilde[1] = f[1];
     for (int i=2; i<n+1; i++){
-       // Updating right hand side of matrix equation:
-       b_vec[i] -= 1/b_vec[i-1];
-       f_tilde[i] = f[i] +f_tilde[i-1]/b_vec[i-1];
+        b_vec[i] -= 1/b_vec[i-1];
+        f_tilde[i] = f[i] + f_tilde[i-1]/b_vec[i-1];
     }
 
     // Backward substition:
     v[n] = f_tilde[n]/b_vec[n];
-    for (int i=n-1;i>=1;i--){
-        v[i] = (f_tilde[i]+v[i+1])/b_vec[i];
+    for (int i=n-1; i>=1; i--){
+        v[i] = (f_tilde[i] + v[i+1])/b_vec[i];
     }
 
     //Finish timing

### 2016-09-15 Project1/article/article.tex
Some additions to LU

@@
 
-The LU decompositon method is a method where the matrix $A$ in
+The LU decomposition method is a method where the matrix $A$ in
 equation \ref{eq:matrix} ($A\vb{v}=\vb{f}$) can be rewritten as a product of two matrices
@@
 matrices, which the matrix $A$ certainly is. 
+\begin{equation}
+A = LU =
+\begin{bmatrix}
+1 & 0 & 0 & \dots & 0 & 0 \\
+l_{21} & 1 & 0 & \dots & 0 & 0 \\
+l_{31} & l_{32} & 1 & \dots & 0 & 0 \\
+  &\vdots & & \ddots & \vdots  & \\
+l_{n-11} & l_{n-12} & l_{n-13} & \dots & 1 & 0 \\  
+l_{n1} & l_{n2} & l_{n3} & \dots & l_{nn-1} & 1
+\end{bmatrix}
+\begin{bmatrix}
+u_{11} & u_{12} & u_{13} & \dots & u_{1n-1} & u_{1n} \\
+0 & u_{22} & u_{23} & \dots & u_{2n-1} & u_{2n} \\
+0 & 0 & u_{33} & \dots & u_{3n-1} & u_{3n} \\
+  &\vdots & & \ddots & \vdots  & \\
+0 & 0 & 0 & \dots & u_{n-1n-1} & u_{n-1n} \\  
+0 & 0 & 0 & \dots & 0 & u_{nn}
+\end{bmatrix}
+\end{equation}
+
+The previous equation can be solved in two steps
+\begin{equation}
+L\vb{w}=\vb{f}, \quad U\vb{v}=\vb{w}
+\end{equation}
 

### 2016-09-15 Project1/article/article.tex
Started results section

@@
 
-The previous equation can be solved in two steps
+The equation $A\vb{v}=LU\vb{v}=\vb{f}$ can be solved in two steps
 \begin{equation}
@@
 \end{equation}
+Written out, the full set of these linear equations will look like
+\begin{align}
+w_1 &= f_1 \label{eq:LU1}\\
+l_{21}w_1 + w_2 &= f_2 \\
+l_{31}w_1 + l_{32}w_2 +w_3 &= f_3 \\
+ & \vdots \\
+l_{n-11}w_1 + l_{n-12}w_2 + l_{n-13}w_3 + \dots + w_{n-1}  &=
+                                                                    f_{n-1}\\
+l_{n1}w_1 + l_{n2}w_2 + l_{n3}w_3 + \dots + l_{nn-1}w_{n-1} + w_n &=
+                                                                    f_n
+  \label{eq:LU2}\\  
+u_{11}v_1 + u_{12}v_2 + u_{13}v_3 + \dots + u_{1n-1}v_{n-1} +
+  u_{1n}v_n &= w_1  \label{eq:LU4}\\  
+u_{22}v_2 + u_{23}v_3 + \dots + u_{2n-1}v_{n-1} + u_{2n}v_n &= w_2 \\
+u_{33}v_3 + \dots+u_{3n-1}v_{n-1} + u_{3n}v_n &= w_3 \\
+ & \vdots \\
+u_{n-1n-1}v_{n-1} &=  w_{n-1} \\
+u_{nn}v_n &= w_n \label{eq:LU3}
+\end{align}
+First $\vb{w}$ can be determined by iterating forwards through
+equations \ref{eq:LU1} to \ref{eq:LU2}. Then $\vb{v}$ can be found be
+iterating backwards from equation \ref{eq:LU3} to equation
+\ref{eq:LU4}.
+
+In this study the Armadillo library
+is employed instead of implementing the LU decomposition
+algorithm from scratch. The LU method is used as a benchmark against which the
+speed of the tridiagonal method is measured. The number of FLOPS
+required by the LU is $2n^3/3$, by any alogorithm used to implement
+it\cite{golub}. Additionally, $n^2$ FLOPS is required by the forward
+and backwards substitution.
+\begin{equation}
+N_{LU}=\frac{2}{3}n^3-2n^2=\mathcal{O}(\frac{2}{3}n^3)
+\end{equation}
+Using LU computation results in a quadratic slowdown compared to the
+tridiagonal matrix algorithm.
 
 \section{Results}
+
+\begin{table}[h]
+\caption{Elapsed time for different $n$}
+\begin{tabular}{lcc}
+\hline
+n & TDMA [s] & LU [s] \\ \hline
+10 & 0.0000035 & 0.00312535 \\
+100 & 0.0000116 & 0.00193754 \\
+1000 & 0.000077892 & 0.0677764 \\
+10000 & 0.000878769 & 21.9247 \\
+100000 & 0.00757418 & n/a \\
+1000000 & 0.08616075 & n/a \\
+10000000 & 0.76534 & n/a
+\end{tabular}
+\label{tab:solver_times}
+\end{table}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.9\linewidth]{figures/10.png}
+  \caption{Numerical approximation for $n=10$}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.9\linewidth]{figures/100.png}
+  \caption{Numerical approximation for $n=100$}
+\end{figure}
 
@@
         Physics - Lecture Notes 2016}. University of Oslo}
+    \bibitem{golub}{Golub, G.H., van Loan, C.F. (1996). \emph{Matrix
+          Computations} (3rd ed.), Baltimore: John Hopkins.}
 \end{thebibliography}

### 2016-09-15 Project1/main.cpp

@@ -3,9 +3,9 @@
 
 int main() {
 
-    int n = pow(10,1);
+    int n = pow(10,2);
 
-    problem1ab(n);
+    //problem1ab(n);
     problem1c(n);
     //problem1e(n);
 

### 2016-09-15 Project1/plot.py

@@ -24,10 +24,10 @@ for line in f:
     k += 1
 
 # Plotting
-plt.plot(x, u, '-b', label="Amalytic solution")
+plt.plot(x, u, '-b', label="Analytic solution")
 plt.plot(x, v, '--r', label="Numerical solution")
 plt.title('n =' + str(n), fontsize=18)
 plt.xlabel(r'$x_i$', fontsize=20)
-plt.ylabel(r'$u(x_i) \quad \& \quad  v_i$', fontsize=20)
+plt.ylabel(r'$u_i \quad \& \quad  v_i$', fontsize=20)
 plt.legend(fontsize=18)
 plt.show()

### 2016-09-15 Project1/relerrorplot.py

@@ -1,0 +1,15 @@
+from matplotlib import pyplot as plt
+#plt.rc('font', **{'family': 'serif', 'serif': ['Computer Modern']})
+#plt.rc('text', usetex=True)
+#plt.rcParams['xtick.labelsize'] = 18
+#plt.rcParams['ytick.labelsize'] = 18
+
+n = [10, 100, 1000, 10000, 10000, 1000000]
+logsh=[-1.04139,-2.00432,-3.0043,-4.00004,-5,-6]
+eps = [-1.1797,-3.08804, -5.08005,-7.07936,-9.00478,-6.77137]
+
+plt.plot(logsh, eps, '-ob')
+plt.title("Plot of maximal relative error vs step size", fontsize=18)
+plt.xlabel(r'$log_{10}(h)$', fontsize=20)
+plt.ylabel(r'$max(log_{10}(|(u_i-v_i)/u_i|)$', fontsize=20)
+plt.show()

### 2016-09-15 Project1/article/article.tex
Results part shaping up

@@
 
+The run time for the optimized tridiagonal matrix algorithm and the LU
+decomposition solver is in table \ref{tab:solver_times}. The reason
+run times are not given for $n>10000$ for the LU method is because the
+method uses too much memory for the solver to even
+function. Regardless, the message of table \ref{tab:solver_times} is
+quite clear, the specialized tridiagonal matrix method is vastly more
+efficient.
+
 \begin{table}[h]
-\caption{Elapsed time for different $n$}
+\caption{Elapsed time for increasing $n$}
 \begin{tabular}{lcc}
@@
 n & TDMA [s] & LU [s] \\ \hline
-10 & 0.0000035 & 0.00312535 \\
-100 & 0.0000116 & 0.00193754 \\
+10 & 0.0000035 & 0.00106083 \\
+100 & 0.0000116 & 0.0022319 \\
 1000 & 0.000077892 & 0.0677764 \\
@@
 
+Tables \ref{fig:n10} and \ref{fig:n100} show the fit of the
+tridiagonal matrix solver for $n=10$ and $n=100$, respectively. One
+can easily see that already at $n=100$ the fit is already very
+good.
+
 \begin{figure}[h]
@@
   \includegraphics[width=0.9\linewidth]{figures/10.png}
-  \caption{Numerical approximation for $n=10$}
+  \caption{Numerical approximation of the tridiagonal matrix algorithm for $n=10$}
+  \label{fig:n10}
 \end{figure}
@@
   \includegraphics[width=0.9\linewidth]{figures/100.png}
-  \caption{Numerical approximation for $n=100$}
+  \caption{Numerical approximation of the tridiagonal matrix algorithm for $n=100$}
+  \label{fig:n100}
 \end{figure}
 
+Figure \ref{fig:relerror} shows a plot of the maximum relative error of the
+tridiagonal solver as a function of step size. One can see that the
+relative error falls as the step size falls, but only to a certain
+point. By decreasing the step size further the error increases, most
+likely because of decreasing numerical precision because of the
+computers difficulty in representing very small numbers.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.9\linewidth]{figures/relerror.png}
+  \caption{Plot of maximum relative error as a function of step size}
+  \label{fig:relerror}
+\end{figure}
 \section{Discussion}

### 2016-09-15 Project1/main.cpp

@@ -3,11 +3,11 @@
 
 int main() {
 
-    int n = pow(10,2);
+    int n = pow(10,1);
 
     //problem1ab(n);
     problem1c(n);
-    //problem1e(n);
+    problem1e(n);
 
     return 0;
 }

### 2016-09-16 Project1/article/article.tex
Discussion

@@
 \end{figure}
+
 \section{Discussion}
+When choosing a suitible value for $n$, one must take into account both precision and speed.
+Our intuition tells us that a larger $n$ should result in a higher precision for the numerical derivative.
+We can conclude from figure \ref{fig:relerror} that this is only true up to a certain point, because of loss in numerical representation accuracy within the computer.
+An ideal n will be in the range of $n\approx 10^5$ which is equivalent to $h=10^-5$.
+Increasing this value will not increase run time dratically for the tridiagnoal matrix algorithm.
+Choosing $n=10^5$ leaves us with the ability to calculate with a high certainty of 5 decimals points and with low memory cost.
+
+By looking at the results one is led to believe that there are more reasons to prefer the tridiagonal matrix algorithm for this kind of problem.
+When using the LU-decomposition method the short term memory is quickly filled with nothing but the matrix $A$.
+Choosing $n=10^5$ is equivalent to $8\times10^{10}$ bytes$=80$GB of memory!
+A brand new computer ships with around $8$ GB of memory today, which is a tenth of the space required.
+Obviously, this inhibits our ability to perform the calculation.
 

### 2016-09-16 Project1/article/article.tex
Conclusion

@@
 By looking at the results one is led to believe that there are more reasons to prefer the tridiagonal matrix algorithm for this kind of problem.
-When using the LU-decomposition method the short term memory is quickly filled with nothing but the matrix $A$.
+When using the LU decomposition method the short term memory is quickly filled with nothing but the matrix $A$.
 Choosing $n=10^5$ is equivalent to $8\times10^{10}$ bytes$=80$GB of memory!
@@
 \section{Conclusion}
+As we have seen, there is significant advantage to fine-tuning and picking the correct algorithm before solving a problem numerically.
+We have experienced the restrictions in memory when computing, and have been made aware of the problem this poses when doing numerical calculations.
+The specialised tridiagonal matrix algorithm was able to solve the one-dimensional Poisson equation more than $20000$ times faster than the more general LU decomposition method.
+This can be explained by the fact that the LU decomposition method require $\mathcal{O}(\frac{2}{3}n^3)$ FLOPS while the specialized tridiagonal matrix algorithm will do with a meager $\mathcal{O}(8n)$ FLOPS.
 
@@
       New York.}
-    \bibitem{morten}{Hjorth-Jensen, M. (2016). \emph{Computational
-        Physics - Lecture Notes 2016}. University of Oslo}
+    \bibitem{morten}{Hjorth-Jensen, M. (2015). \emph{Computational
+        Physics - Lecture Notes 2015}. University of Oslo}
     \bibitem{golub}{Golub, G.H., van Loan, C.F. (1996). \emph{Matrix

### 2016-09-16 Project1/article/article.tex
Introduction and abstract

@@
 
-\title[Solving the Poisson-equation in one dimension]{Solving the Poisson-equation in one dimension \\
+\title[Solving the Poisson-equation in one dimension]{Solving the Poisson-equation in one dimension: \\
+\normalsize{TDMA vs LU} \\
   \hrulefill\small{ FYS3150: Computational Physics }\hrulefill}
@@
 \begin{abstract}
-Lorem ipsum dolor sit amet, consectur adipiscing lit. Nullam ut lacus eget lorem...
+A one-dimensional version of the Poisson equation with Dirichlet boundary conditions is solved using two different algorithms.
+The two algorithms employed are the tridiagonal matrix algorithm and the LU decomposition method.
+We find that a fine-tuned version of the tridiagonal matrix algorithm is $10^5$ times faster than the general LU decomposition method and $10$ times as precise.
 \end{abstract}
@@
 \section{Introduction}
+In this study the one-dimensional Poisson dimension will be solved using two different algorithms.
+The Poisson equation is a partial differential equation fo elliptic type with broad application.
+In physics it is for instance used to describe the potential field of a charge, or mass density distribution.
+By assuming spherical symmetry, the Poisson equation is reduced to one dimension.
+
+Herein, we devise two algorithms for solving the Poisson equation.
+The first one is the tridiagonal matrix algorithm, and the second is the LU decomposition method.
+We see how the tridiagonal matrix algorithm can be tailored to the problem and therefore becomes vastly more efficient than the more general LU method.
+The algorithms are tested at different precision levels.
+
+Firstly, the physical background for the Poisson equation and how the second derivative can be approximated by a three point formula is presented.
+Secondly, we provide a thorough elaboration of how the algorithms can be implemented.
+Thirdly, results of the algorithms perfomance is presented.
+Lastly, we provide a discussion of the findings and some concluding remarks.
 
@@
 \subsection{The Poisson Equation}
-
 The Poisson equation is a classical equation from electromagnetism.
@@
 along the diagonals. The second method is the LU-decomposistion
-method.
+method. Code snippets are included in this section, for the full programs follow the link to github at the first page of this article.
 
@@
 We can conclude from figure \ref{fig:relerror} that this is only true up to a certain point, because of loss in numerical representation accuracy within the computer.
-An ideal n will be in the range of $n\approx 10^5$ which is equivalent to $h=10^-5$.
+An ideal n will be in the magnitude of $10^5$ which is equivalent to $h=10^-5$.
 Increasing this value will not increase run time dratically for the tridiagnoal matrix algorithm.
-Choosing $n=10^5$ leaves us with the ability to calculate with a high certainty of 5 decimals points and with low memory cost.
+Choosing $n=10^5$ gives us the ability to calculate with a high certainty of 5 decimals points and with low memory cost.
 

### 2016-09-16 Project1/article/article.tex
A final read-through?

@@
 where $\laplacian$ is the Laplace operator.
-In three dimensions the Laplace operator can be expressed using spherical coordinates, but in this study I am assuming that $\Phi$ and $\rho$ are spherically symmetric, thus reducing the equation to a one-dimensional problem.
+In three dimensions the Laplace operator can be expressed using spherical coordinates, but in this study we are assuming that $\Phi$ and $\rho$ are spherically symmetric, thus reducing the equation to a one-dimensional problem.
 only dependent on radius $r$.
@@
 \emph{Thomas Algorithm} \cite{thomasalgo}. This is a simplified form of Gaussian
-elimination that can ve used to solve tridiagonal systems of
+elimination that can be used to solve tridiagonal systems of
 equations. The method is improved upon in order to take into account
@@
 along the diagonals. The second method is the LU-decomposistion
-method. Code snippets are included in this section, for the full programs follow the link to github at the first page of this article.
+method. Code snippets are included in this section, for the full programs, follow the link to github at the first page of this article.
 
@@
 substitution'' has been performed as well. This ensures that all
-eliments in the row reduced matrix are pivot elements.
+elements in the row reduced matrix are pivot elements.
 
@@
 When using the LU decomposition method the short term memory is quickly filled with nothing but the matrix $A$.
-Choosing $n=10^5$ is equivalent to $8\times10^{10}$ bytes$=80$GB of memory!
+Choosing $n=10^5$ is equivalent to $8\times10^{10}$ bytes $=80$GB of memory!
 A brand new computer ships with around $8$ GB of memory today, which is a tenth of the space required.

### 2016-09-28 Project2/article/article.tex
Jacobi method done but not working

--- 2016-09-16 Project1/article/article.tex
@@
 \usepackage[]{listings}
-\usepackage[T1]{fontenc}
-\usepackage{color}
+\usepackage[utf8]{inputenc}
 
-\definecolor{mygreen}{rgb}{0,0.6,0}
-\definecolor{mygray}{rgb}{0.5,0.5,0.5}
-\definecolor{mymauve}{rgb}{0.58,0,0.82}
-
-\lstset{ %
-  backgroundcolor=\color{white},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}
-  basicstyle=\footnotesize,        % the size of the fonts that are used for the code
-  breakatwhitespace=false,         % sets if automatic breaks should only happen at whitespace
-  breaklines=true,                 % sets automatic line breaking
-  captionpos=b,                    % sets the caption-position to bottom
-  commentstyle=\color{mygreen},    % comment style
-  deletekeywords={...},            % if you want to delete keywords from the given language
-  escapeinside={\%*}{*)},          % if you want to add LaTeX within your code
-  extendedchars=true,              % lets you use non-ASCII characters; for 8-bits encodings only, does not work with UTF-8
-  frame=single,	                   % adds a frame around the code
-  keepspaces=true,                 % keeps spaces in text, useful for keeping indentation of code (possibly needs columns=flexible)
-  keywordstyle=\color{blue},       % keyword style
-  language=c++,                 % the language of the code
-  otherkeywords={*,...},           % if you want to add more keywords to the set
-  rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
-  showspaces=false,                % show spaces everywhere adding particular underscores; it overrides 'showstringspaces'
-  showstringspaces=false,          % underline spaces within strings only
-  showtabs=false,                  % show tabs within strings adding particular underscores
-  stepnumber=2,                    % the step between two line-numbers. If it's 1, each line will be numbered
-  stringstyle=\color{mymauve},     % string literal style
-  tabsize=2,	                   % sets default tabsize to 2 spaces
+\lstset{
+	frame = single,
+	language = C++,
+	showstringspaces = false,
+	tabsize = 2,
+	otherkeywords = {self},
+	keywordstyle = \color{blue},
+	identifierstyle=\color{deepgreen},
+ 	stringstyle=\color{orange},
+ 	backgroundcolor=\color{mygray}
 }
 
-\title[Solving the Poisson-equation in one dimension]{Solving the Poisson-equation in one dimension \\
+\title[Schrödinger's Equation in 3D]{Schrödinger's Equation in a Three-Dimensional Harmonic Oscillator Well \\
   \hrulefill\small{ FYS3150: Computational Physics }\hrulefill}
-
-\author[Svalheim \& Winther-Larsen]{Trygve Leithe Svalheim \\
-   Sebastian G. Winther-Larsen \\
-  \href{https://github.com/gregwinther/FYS3150/}{\texttt{github.com/gregwinther}}}
 
@@
 \begin{abstract}
-Lorem ipsum dolor sit amet, consectur adipiscing lit. Nullam ut lacus eget lorem...
+THIS IS NOT THE ABSTRACT FOR THIS PROJECT :)
+A one-dimensional version of the Poisson equation with Dirichlet boundary conditions is solved using two different algorithms.
+The two algorithms employed are the tridiagonal matrix algorithm and the LU decomposition method.
+We find that a fine-tuned version of the tridiagonal matrix algorithm is $10^5$ times faster than the general LU decomposition method and $10$ times as precise.
 \end{abstract}
@@
 
-\section{Introduction}
-
-\section{Theory}
-\subsection{The Poisson Equation}
-
-The Poisson equation is a classical equation from electromagnetism.
-The electrostatic potential $\Phi$ is generated by a localized charge distribution $\rho(\vb{r})$.
-In three dimensions the equation reads
-\begin{equation}
-\laplacian\Phi=-4\pi \rho(\vb{r})
-\end{equation}
-where $\laplacian$ is the Laplace operator.
-In three dimensions the Laplace operator can be expressed using spherical coordinates, but in this study I am assuming that $\Phi$ and $\rho$ are spherically symmetric, thus reducing the equation to a one-dimensional problem.
-only dependent on radius $r$.
-\begin{equation}
-\laplacian=\frac{1}{r^2}\frac{d}{dr}\left(r^2\frac{d\Phi}{dr} \right)
-\end{equation}
-By substituting $\Phi(r)=\phi(r)/r$ the Poisson equation is reduced to 
-\begin{equation}
-\frac{d^2\phi}{dr^2}=-4\pi r\rho(r)
-\end{equation}
-and by letting $\phi \rightarrow u$ and $r \rightarrow x$ one is left with the very simple equation
-\begin{equation}
--u''(x)=f(x) \label{eq:2nd}
-\end{equation}
-
-The inhomogenous term $f$, or source term, is given by the charge distribution $\rho$ multiplied by $r$ and the constant $-4\pi$.
-In this study, however, the source term will be $f(x)=100e^{-10x}$ and the results can be compared to the analytical solution $u(x)=1-(1-e^{-10})x-e^{-10x}$.
-
-\subsection{Approximation of the Second Derivative}
-In this study the one-dimensional Poisson equation will be solved with Dirichlet boundary conditions by rewriting it as a set of linear equations.
-The discretized approximation of $u$ is defined as $v_i$ with grid points $x_i=ih$, step size of $h=\frac{1}{n+1}$, in the interval $x_0=0$ to $x_{n+1}=1$ and with boundary conditions $v_0=v_n+1=0$.
-The interior solution $v_i \forall i \in {1,...,n}$ is to be found.
-The second order derivative is approximated with the three point formula such that equation \ref{eq:2nd} becomes
-\begin{equation}
--\frac{v_{i+1}-2v_i+v{i-1}}{h²}=f_i \label{eq:2approx}
-\end{equation}
-
-By defining $\vb{f}=h^2f_i$ one can rewrite equation
-\ref{eq:2approx} as $-v_{i+1}-2v_i+v_{i-1}=h^2f_i$. If we ignore the
-end points, $i=0$ and $i=n+1$, this equation can be represented as a
-matrix equation.
-\begin{equation}
-\label{eq:matrix}
-A\vb{v}=\vb{f}
-\end{equation}
-\begin{equation}
-\begin{bmatrix}
-2 & -1 & 0 & 0 & \cdots & 0 & 0 & 0 \\
--1 & 2 & -1 & 0 & \cdots & 0 & 0 & 0 \\
-0 & -1 & 2 & -1 & \cdots & 0 & 0 & 0 \\ 
-& & \vdots &  & \ddots &  & \vdots & \\
-0 & 0 & 0 & 0 & \cdots & -1 & 2 & -1 \\
-0 & 0 & 0 & 0 & \cdots & 0 & -1 & 2 
-\end{bmatrix}
-\begin{bmatrix}
-v_1 \\
-v_2 \\
-v_3 \\
-\vdots \\
-v_{n-1} \\
-v_n
-\end{bmatrix}=
-\begin{bmatrix}
-f_1 \\
-f_2 \\
-f_3 \\
-\vdots \\
-f_{n-1} \\
-f_n
-\end{bmatrix}
-\end{equation}
-
-\section{Algorithms}
-Two main methods are implemented and compared. The first method is
-gaussian elimination of the tridiagonal matrix $A$, also known as the
-\emph{Thomas Algorithm} \cite{thomasalgo}. This is a simplified form of Gaussian
-elimination that can ve used to solve tridiagonal systems of
-equations. The method is improved upon in order to take into account
-the fact that the matrix we are dealing with has the same numbers
-along the diagonals. The second method is the LU-decomposistion
-method.
-
-\subsection{Tridiagonal Matrix Algorithm}
-Our tridiagonal system can be represented by 
-\begin{equation}
-a_ix_{i-1} + b_ix_i + c_ix_{i+1} = \vb{f}
-\end{equation}
-with $a_i=-1,b_i=2$ and $c_i=-1$, except for $a_1=0$ and $c_n=0$. Row
-reducing a matrix will reveal how the algorithm functions. Limiting
-the problem to four dimensions for easier reading and to save the
-rainforest\footnote{Writing out a general case will also take up more paper
-space}.
-
-\begin{equation}
-\left[
-\begin{array}{cccc|c}
-b_1 & c_1 & 0 & 0 & f_1 \\
-a_2 & b_2 & c_2 & 0  & f_2 \\
-0 & a_3 & b_3 & c_3 & f_3 \\
-0 & 0 & a_4 & b_4 & f_4
-\end{array}
-\right] \sim
-\left[
-\begin{array}{cccc|c}
-1 & c_1/b_1 & 0 & 0 & b_1/b_1 \\
-0 & b_2-\frac{c_1}{b_1}a_2 & c_2 & 0  & f_2-\frac{f_1}{b_1}a_2 \\
-0 & a_3 & b_3 & c_3 & f_3 \\
-0 & 0 & a_4 & b_4 & f_4
-\end{array}
-\right]
-\end{equation}
-Now let $\beta_1=b_1$ and $\beta_2=b_2-\frac{c_1}{b_1}a_2$. As new
-elements start to appear in vector $\vb{f}$, right to the vertical bar
-in the augmented matrix, they are also relabeled to $\tilde{f}_i$. For
-example $\tilde{f}_1=f_1/\beta_1$. One more iteration will reveal the
-pattern of the algorithm.
-
-\begin{align}
-\left[
-\begin{array}{cccc|c}
-1 & c_1/\beta_1 & 0 & 0 & \tilde{f}_1 \\
-0 & 1 & c_2/\beta_2 & 0  & (f_2-\tilde{f}a_2)/\beta_2 \\
-0 & a_3 & b_3 & c_3 & f_3 \\
-0 & 0 & a_4 & b_4 & f_4
-\end{array}
-\right] &=
-\left[
-\begin{array}{cccc|c}
-1 & c_1/\beta_1 & 0 & 0 & \tilde{f}_1 \\
-0 & 1 & c_2/\beta_2 & 0  & \tilde{f}_2 \\
-0 & a_3 & b_3 & c_3 & f_3 \\
-0 & 0 & a_4 & b_4 & f_4
-\end{array}
-\right] \\ \sim 
-\left[
-\begin{array}{cccc|c}
-1 & c_1/\beta_1 & 0 & 0 & \tilde{f}_1 \\
-0 & 1 & c_2/\beta_2 & 0  & \tilde{f}_2 \\
-0 & 0 & b_3-\frac{c_2}{\beta_2}a_3 & c_3 & f_3-\tilde{f}_2a_3 \\
-0 & 0 & a_4 & b_4 & f_4
-\end{array}
-\right] &\sim 
-\left[
-\begin{array}{cccc|c}
-1 & c_1/\beta_1 & 0 & 0 & \tilde{f}_1 \\
-0 & 1 & c_2/\beta_2 & 0  & \tilde{f}_2 \\
-0 & 0 & 1 & c_3/\beta_3 & (f_3-\tilde{f}_2a_3)/\beta_3 \\
-0 & 0 & a_4 & b_4 & f_4
-\end{array}\right] \\ 
-\sim &\dots \sim
-\left[
-\begin{array}{cccc|c}
-1 & c_1/\beta_1 & 0 & 0 & \tilde{f}_1 \\
-0 & 1 & c_2/\beta_2 & 0  & \tilde{f}_2 \\
-0 & 0 & 1 & c_3/\beta_3 & \tilde{f}_3 \\
-0 & 0 & 0 & 1 & \tilde{f}_4
-\end{array}
-\right]
-\end{align}
-
-This pattern can be summarized nicely by the following
-difference equations. This computation will from now on be referred to
-as the ``forward substitution''. 
-\begin{equation}
-\label{eq:forward1}
-\beta_i=b_i-\frac{a_ic_{i-1}}{\beta_{i-1}}, \quad
-\tilde{f}_i=f_i-a_i\tilde{f}_{i-1}/\beta_{i-1}, \quad i \in [2,n], \quad
-\end{equation}
-
-In c++, this part of the algorithm is implemented just below. Notice
-that $b_i$ is overwritten instead of initializing a seperate vector
-for $\beta$. 
-\lstinputlisting[language=c++, firstline=65, lastline=71]{../problems.cpp}
-
-The Gaussian elimination is not fully complete until a ``backward
-substitution'' has been performed as well. This ensures that all
-eliments in the row reduced matrix are pivot elements.
-
-\begin{equation}
-\left[
-\begin{array}{cccc|c}
-1 & c_1/\beta_1 & 0 & 0 & \tilde{f}_1 \\
-0 & 1 & c_2/\beta_2 & 0  & \tilde{f}_2 \\
-0 & 0 & 1 & c_3/\beta_3 & \tilde{f}_3 \\
-0 & 0 & 0 & 1 & \tilde{f}_4
-\end{array}
-\right] \sim
-\left[
-\begin{array}{cccc|c}
-1 & c_1/\beta_1 & 0 & 0 & \tilde{f}_1 \\
-0 & 1 & c_2/\beta_2 & 0  & \tilde{f}_2 \\
-0 & 0 & 1 & 0 & \tilde{f}_3 -\frac{c_3}{\beta_3}\tilde{f}_4 \\
-0 & 0 & 0 & 1 & \tilde{f}_4
-\end{array}
-\right]
-\end{equation}
-
-The notation is updated again to $v_3 =\tilde{f}_3 -
-\frac{c_3}{\beta_3}\tilde{f}_4$. By continuing this iterative process
-all the was to the beginning of the matrix by backward substitution
-leaves us with a second recurrence relation.
-\begin{equation}
-\label{eq:backward1}
-v_i=(\tilde{f}_i - c_iv_{i+1})/\beta_i, \quad i \in \{n-1,n-2,\dots,1\}
-\end{equation}
-
-The backward substitution is implemented in C++ like so:
-\lstinputlisting[language=c++, firstline=73,
-lastline=77]{../problems.cpp}
-
-An ever-important aspect of any algorithm is how efficient the algorithm is. It
-is therefor worthwhile to consider how many floating point operations
-the algorithm requires\footnote{The number of $+$,$-$,$*$ and $/$
-  operations are counted}. For the forward substitution there are $6$
-FLOPS in each iteration and $n-1$ iterations, resulting in $6(n-1)$
-FLOPS for the entire loop. Additionally, $1(n-1)$ FLOPS are required
-for the backward substitution, adding up to 
-\begin{equation}
-N_{tridiag}=8(n-1)=\mathcal{O} (8n).
-\end{equation}
-
-\subsubsection{Optimizing the algorithm}
-
-Our case is special, because the elements in each of the diagonal arrays in the
-tridiagonal matrix, $A$, are equal. $a_i=-1,b_i=2$ and $c_i=-1$,
-except for $a_1=0$ and $c_n=0$. This is an advantage, because the
-algorithm can be adjusted, and the number of FLOPS consequently reduced.
-
-The algorithm for the forward substitution, in equation
-\ref{eq:forward1}, can be simplified to
-\begin{equation}
-\beta_i= b_i - 1/\beta_{i-1} \quad \tilde{f}_i=f_i-\tilde{f}_{i-1}/\beta_{i-1}
-\end{equation}
-and the backward substitution, in equation \ref{eq:backward1}, can be
-simplified to
-\begin{equation}
-v_i = (\tilde{f}_i+v_{i+1})/\beta_i
-\end{equation}
-
-This results in some new and sleeker C++ code
-\lstinputlisting[language=c++, firstline=146,
-lastline=158]{../problems.cpp}
-
-One can see that this new and improved algorithm only requires
-$6(n-1)$ FLOPS. In general, a Gaussian row reduction algorithm
-requires $\frac{2}{3}n^3+\mathcal{O}(n^2)$ FLOPS \cite{morten}, a
-great improvement!
-
-\subsection{LU Decomposition}
-
-The LU decomposition method is a method where the matrix $A$ in
-equation \ref{eq:matrix} ($A\vb{v}=\vb{f}$) can be rewritten as a product of two matrices
-$L$ and $U$, where $L$ is lower triangular and and $U$ is upper
-triangular. This method only works for non-singular, invertible
-matrices, which the matrix $A$ certainly is. 
-\begin{equation}
-A = LU =
-\begin{bmatrix}
-1 & 0 & 0 & \dots & 0 & 0 \\
-l_{21} & 1 & 0 & \dots & 0 & 0 \\
-l_{31} & l_{32} & 1 & \dots & 0 & 0 \\
-  &\vdots & & \ddots & \vdots  & \\
-l_{n-11} & l_{n-12} & l_{n-13} & \dots & 1 & 0 \\  
-l_{n1} & l_{n2} & l_{n3} & \dots & l_{nn-1} & 1
-\end{bmatrix}
-\begin{bmatrix}
-u_{11} & u_{12} & u_{13} & \dots & u_{1n-1} & u_{1n} \\
-0 & u_{22} & u_{23} & \dots & u_{2n-1} & u_{2n} \\
-0 & 0 & u_{33} & \dots & u_{3n-1} & u_{3n} \\
-  &\vdots & & \ddots & \vdots  & \\
-0 & 0 & 0 & \dots & u_{n-1n-1} & u_{n-1n} \\  
-0 & 0 & 0 & \dots & 0 & u_{nn}
-\end{bmatrix}
-\end{equation}
-
-The equation $A\vb{v}=LU\vb{v}=\vb{f}$ can be solved in two steps
-\begin{equation}
-L\vb{w}=\vb{f}, \quad U\vb{v}=\vb{w}
-\end{equation}
-Written out, the full set of these linear equations will look like
-\begin{align}
-w_1 &= f_1 \label{eq:LU1}\\
-l_{21}w_1 + w_2 &= f_2 \\
-l_{31}w_1 + l_{32}w_2 +w_3 &= f_3 \\
- & \vdots \\
-l_{n-11}w_1 + l_{n-12}w_2 + l_{n-13}w_3 + \dots + w_{n-1}  &=
-                                                                    f_{n-1}\\
-l_{n1}w_1 + l_{n2}w_2 + l_{n3}w_3 + \dots + l_{nn-1}w_{n-1} + w_n &=
-                                                                    f_n
-  \label{eq:LU2}\\  
-u_{11}v_1 + u_{12}v_2 + u_{13}v_3 + \dots + u_{1n-1}v_{n-1} +
-  u_{1n}v_n &= w_1  \label{eq:LU4}\\  
-u_{22}v_2 + u_{23}v_3 + \dots + u_{2n-1}v_{n-1} + u_{2n}v_n &= w_2 \\
-u_{33}v_3 + \dots+u_{3n-1}v_{n-1} + u_{3n}v_n &= w_3 \\
- & \vdots \\
-u_{n-1n-1}v_{n-1} &=  w_{n-1} \\
-u_{nn}v_n &= w_n \label{eq:LU3}
-\end{align}
-First $\vb{w}$ can be determined by iterating forwards through
-equations \ref{eq:LU1} to \ref{eq:LU2}. Then $\vb{v}$ can be found be
-iterating backwards from equation \ref{eq:LU3} to equation
-\ref{eq:LU4}.
-
-In this study the Armadillo library
-is employed instead of implementing the LU decomposition
-algorithm from scratch. The LU method is used as a benchmark against which the
-speed of the tridiagonal method is measured. The number of FLOPS
-required by the LU is $2n^3/3$, by any alogorithm used to implement
-it\cite{golub}. Additionally, $n^2$ FLOPS is required by the forward
-and backwards substitution.
-\begin{equation}
-N_{LU}=\frac{2}{3}n^3-2n^2=\mathcal{O}(\frac{2}{3}n^3)
-\end{equation}
-Using LU computation results in a quadratic slowdown compared to the
-tridiagonal matrix algorithm.
-
-\section{Results}
-
-The run time for the optimized tridiagonal matrix algorithm and the LU
-decomposition solver is in table \ref{tab:solver_times}. The reason
-run times are not given for $n>10000$ for the LU method is because the
-method uses too much memory for the solver to even
-function. Regardless, the message of table \ref{tab:solver_times} is
-quite clear, the specialized tridiagonal matrix method is vastly more
-efficient.
-
-\begin{table}[h]
-\caption{Elapsed time for increasing $n$}
-\begin{tabular}{lcc}
-\hline
-n & TDMA [s] & LU [s] \\ \hline
-10 & 0.0000035 & 0.00106083 \\
-100 & 0.0000116 & 0.0022319 \\
-1000 & 0.000077892 & 0.0677764 \\
-10000 & 0.000878769 & 21.9247 \\
-100000 & 0.00757418 & n/a \\
-1000000 & 0.08616075 & n/a \\
-10000000 & 0.76534 & n/a
-\end{tabular}
-\label{tab:solver_times}
-\end{table}
-
-Tables \ref{fig:n10} and \ref{fig:n100} show the fit of the
-tridiagonal matrix solver for $n=10$ and $n=100$, respectively. One
-can easily see that already at $n=100$ the fit is already very
-good.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=0.9\linewidth]{figures/10.png}
-  \caption{Numerical approximation of the tridiagonal matrix algorithm for $n=10$}
-  \label{fig:n10}
-\end{figure}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=0.9\linewidth]{figures/100.png}
-  \caption{Numerical approximation of the tridiagonal matrix algorithm for $n=100$}
-  \label{fig:n100}
-\end{figure}
-
-Figure \ref{fig:relerror} shows a plot of the maximum relative error of the
-tridiagonal solver as a function of step size. One can see that the
-relative error falls as the step size falls, but only to a certain
-point. By decreasing the step size further the error increases, most
-likely because of decreasing numerical precision because of the
-computers difficulty in representing very small numbers.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=0.9\linewidth]{figures/relerror.png}
-  \caption{Plot of maximum relative error as a function of step size}
-  \label{fig:relerror}
-\end{figure}
-
-\section{Discussion}
-When choosing a suitible value for $n$, one must take into account both precision and speed.
-Our intuition tells us that a larger $n$ should result in a higher precision for the numerical derivative.
-We can conclude from figure \ref{fig:relerror} that this is only true up to a certain point, because of loss in numerical representation accuracy within the computer.
-An ideal n will be in the range of $n\approx 10^5$ which is equivalent to $h=10^-5$.
-Increasing this value will not increase run time dratically for the tridiagnoal matrix algorithm.
-Choosing $n=10^5$ leaves us with the ability to calculate with a high certainty of 5 decimals points and with low memory cost.
-
-By looking at the results one is led to believe that there are more reasons to prefer the tridiagonal matrix algorithm for this kind of problem.
-When using the LU decomposition method the short term memory is quickly filled with nothing but the matrix $A$.
-Choosing $n=10^5$ is equivalent to $8\times10^{10}$ bytes$=80$GB of memory!
-A brand new computer ships with around $8$ GB of memory today, which is a tenth of the space required.
-Obviously, this inhibits our ability to perform the calculation.
-
-\section{Conclusion}
-As we have seen, there is significant advantage to fine-tuning and picking the correct algorithm before solving a problem numerically.
-We have experienced the restrictions in memory when computing, and have been made aware of the problem this poses when doing numerical calculations.
-The specialised tridiagonal matrix algorithm was able to solve the one-dimensional Poisson equation more than $20000$ times faster than the more general LU decomposition method.
-This can be explained by the fact that the LU decomposition method require $\mathcal{O}(\frac{2}{3}n^3)$ FLOPS while the specialized tridiagonal matrix algorithm will do with a meager $\mathcal{O}(8n)$ FLOPS.
-
-\begin{thebibliography}{10}
-  \bibitem{thomasalgo}{Thomas, L.H. (1949), \emph{Elliptic
-        Problems in Linear Differential Equations over a
-        Network}. Watson Sci. Comput. Lab Report, Columbia University,
-      New York.}
-    \bibitem{morten}{Hjorth-Jensen, M. (2015). \emph{Computational
-        Physics - Lecture Notes 2015}. University of Oslo}
-    \bibitem{golub}{Golub, G.H., van Loan, C.F. (1996). \emph{Matrix
-          Computations} (3rd ed.), Baltimore: John Hopkins.}
-\end{thebibliography}
-
 \end{document}

### 2016-09-28 Project2/jacobiMethod.cpp

@@ -8,12 +8,15 @@ double maxOffDiagonals(arma::mat &A, int &k, int &l, int n) {
 
     double max;
 
+    k = 0;
+    l = 0;
+
     for (int i = 0; i < n; ++i) {
 
         // Indexing will only include upper triangular part
         // This assumes that matrix is symmetric
         // Advantage: excludes diagonal, which is what we want
-        for (int j = i+1; j < n; ++ j) {
+        for (int j = i+1; j < n; ++j) {
 
             double aij = fabs(A(i,j));
 
@@ -25,13 +28,19 @@ double maxOffDiagonals(arma::mat &A, int &k, int &l, int n) {
                 l = j;
             }
         }
+        cout << "Largest off-diagonal element is: " << maxOffDiagonals(A, k, l, n) << endl;
+
     }
 
     return max;
 }
 
-void jacobiRotation(arma::mat &A, int &k, int &l, int n) {
-    // A is the input matrix and S is the Givens rotation matrix
+void jacobiRotation(arma::mat &A, arma::mat &R, int &k, int &l, int n) {
+    // A is the input matrix and S is the matrix with eigenvectors after
+    // enough rotations
+
+    // Initiating variables
+    double tau, t, c, s;
 
     /* -----------------------------------------------
      * COMPUTING c = cosine(angle) AND s = sine(angle)
@@ -53,14 +62,14 @@ void jacobiRotation(arma::mat &A, int &k, int &l, int n) {
     // No. 2: t
     // t can be either + or - depending on t**2 + 2*tau*t
     if (tau >= 0) {
-        t = -tau + sqrt(1 + tau*tau);
+        t = -tau + sqrt(1 + tau*tau); 		// tan(angle)
     } else {
-        t = -tau - sqrt(1 + tau*tau);
+        t = -tau - sqrt(1 + tau*tau);		// tan(angle)
     }
 
     // No. 3:
-    c = 1.0 / (sqrt(1 + t*t));
-    s = c*t;
+    c = 1.0 / (sqrt(1 + t*t));				// cos(angle)
+    s = c*t;								// sin(angle)
 
     /* ----------------------------------------------
      * ROTATING
[…]

### 2016-09-28 Project2/jacobimethod.h

@@ -2,6 +2,8 @@
 #define JACOBIMETHOD_H
 
 double maxOffDiagonals(arma::mat &A, int &k, int &l, int n);
-void jacobiRotation(arma:mat &A, int &k, int &l, int n);
+void jacobiRotation(arma::mat &A, arma::mat &R, int &k, int &l, int n);
+void jacobiMethod(arma::mat &A, arma::mat &R, int &k, int &l, int n);
+
 
 #endif // JACOBIMETHOD_H

### 2016-09-28 Project2/main.cpp

@@ -13,13 +13,17 @@ int main(int argc, char *argv[])
     int l = 0;
 
     arma::mat A(n,n);
+    arma::mat R; R.eye(n,n);
 
-    A 	<< 1 << 2 << 3 << arma::endr
-        << 2 << 2 << 4 << arma::endr
-        << 3 << 4 << 5 << arma::endr;
+    A 	<< 4 << 1 << 0 << arma::endr
+        << 1 << -1 << 0 << arma::endr
+        << 0 << 0 << 5 << arma::endr;
 
+    double maxdiag = maxOffDiagonals(A, k, l, n);
     cout << A << endl;
-    cout << "Largest off-diagonal element is: " << maxOffDiagonals(A, k, l, n) << endl;
     cout << "k is different from zero? " << (k != 0) << endl;
+
+    jacobiMethod(A, R, k, l, n);
+
 
 }

### 2016-09-30 Project2/article/article.tex
Working on theory part in article

@@
 \usepackage[utf8]{inputenc}
+\usepackage[toc,page]{appendix}
 
@@
 
+\section{Introduction}
+
+The aim of this study is to solve Schrödinger's equation for two electrons in a three-dimensional  harmonic oscillator well with and without repulsive Coulomb interaction.
+The equation is reformulated to a discretized in order to be represented as a matrix eigenvalue problem that can be solved with Jacobi's method.
+
+\section{Theory}
+We assume that electrons confined in a small area move in a three-dimensional harmonic oscillator potential and repel each other via the static Coulomb interaction.
+To simplify the problem, spherical symmetry is assumed.
+
+\subsection{A single electron}
+The radial part of Schrödinger's equation for a single electron reads
+\begin{equation}
+\label{eq:schr1}
+-\frac{\hbar^2}{2m}\left(\frac{1}{r^2}\frac{d}{dr}r^2\frac{d}{dr}-\frac{l(l+1)}{r^2} \right)R(r) + V(r)R(r) = ER(r)
+\end{equation}
+where $V(r)$ is the harmonic oscillator potential\footnote{One can in theory insert any potential into the Schrödinger equation, but in this case the harmonic oscillator potential is employed.}, $V(r)=\frac{1}{2}kr^2$ where $k=m\omega^2$, and consequently $E$ is the energy of the harmonic oscillator in three dimensions.
+The angular frequency of the oscillator is given by $\omega$ and is related to the energy of the system as follows
+\begin{equation}
+E_{nl}=\hbar\omega\left(2n + l  + \frac{3}{2} \right), \quad n,l \in \mathbb{N} 
+\end{equation}
+However, we will consider cases with zero angular momentum only ($l=0$).
+Moreover, by a somewhat lengthy substitutive exercise will culminate in a rewriting of Schrödinger's equation in \ref{eq:schr1} to a dimensionless form
+\begin{equation}
+\label{eq:schr2}
+-\frac{d^2}{d\rho^2}u(\rho) + \rho^2u(\rho) = \lambda u(\rho)
+\end{equation}
+where $u(r) = R(r)/r$ and $\rho = r/\alpha$ is a dimensionless parameter for length.
+The constant $\alpha$ is $\alpha = \sqrt{\hbar^2/mk}$ and the energy parameter (eigenvalues) are $\lambda = 2m\alpha^2E/\hbar^2$.
+The three first eigenvalues will be $\lambda_0= 3$, $\lambda_1 = 7$ and $\lambda_0=11$.
+
+The second derivative in equation \ref{eq:schr2} will be approximated by the well-known central finite difference method for the second derivative,
+\begin{equation}
+\frac{du^2}{d\rho^2}=\frac{u(\rho+h)-2u(\rho)+u(\rho-h)}{h^2}+\mathcal{O}(h^2)
+\end{equation}
+where $h = (\rho_N-\rho_0)/h$ is the step size. By denoting $u(\rho_0+ih)$ by $u_i$, equation \ref{eq:schr2} can be further rewritten to
+\begin{align}
+-\frac{1}{h^2}u_{i-1}+\left(\frac{2}{h^2} + \rho_i^2 \right)u_i-\frac{1}{h^2}u_{i+1}=\lambda u_i \label{eq:schr3}\\
+A\vb{u} = \lambda \vb{u} \label{eq:mat1}
+\end{align} 
+where the matrix in \ref{eq:mat1} is
+\begin{equation}
+\begin{bmatrix}
+\frac{1}{h^2} + \rho_i^2 & -\frac{1}{h^2} & 0 & 0 & 0 & \dots & 0 \\
+-\frac{1}{h^2} & \frac{1}{h^2} + \rho_i^2 & -\frac{1}{h^2} & 0 & 0 & \dots & 0 \\
+0 & -\frac{1}{h^2} & \frac{1}{h^2} + \rho_i^2 & -\frac{1}{h^2} & 0 & \dots & 0  \\
+\vdots & \vdots & \vdots &\ddots & \ddots & \ddots & \vdots \\
+0 & 0 & 0 & \dots &  -\frac{1}{h^2} & \frac{1}{h^2} + \rho_i^2 & -\frac{1}{h^2} \\
+0 & 0 & 0 & 0 & \dots & -\frac{1}{h^2} & \frac{1}{h^2} + \rho_i^2 
+\end{bmatrix}
+\end{equation}
+
+By finding the eigenvectors, or the energies, corresponding to the eigenvalues $\lambda_i$  equation \ref{eq:schr3} above, one has found the wave function that describe the behavior of the electron for a particular state $i$.
+To have something to compare against it is useful to state the exact, analytical wave function for the three lowest states.
+The general form of these radial wave functions is
+\begin{equation}
+\label{eq:radialschr1}
+R_{n,l=0}(r)=N_ne^{-\frac{m\omega}{2\hbar}r^2}\sqrt{L_n}\left(-\frac{m\omega}{2\hbar} r^2 \right)
+\end{equation}
+where $L_n$ refers to a particular Laguerre polynomial.
+More on the Laguerre polynomials can be found in appendix \ref{app:laguerre}.
+Reevaluating this function for the three lowest energy states and normalizing yields the following equations
+\begin{align}
+\abs{u_0(\rho)}^2 &= \frac{4}{\sqrt{\pi}}\rho^2e^{-\rho^2} \\
+\abs{u_1(\rho)}^2 &= \frac{8}{3\sqrt{\pi}}\rho^2\left( \frac{3}{2}-\rho^2 \right)e^{-\rho^2} \\
+\abs{u_2(\rho)}^2 &= \frac{8}{15\sqrt{\pi}}\rho^2\left( \frac{15}{4}-5\rho^2 +\rho^4\right)e^{-\rho^2}
+\end{align}
+It is only sensible to deal with the absolute squared values of the wave functions, because this is a probability density function for the position of the electron.
+
+\subsection{Two electrons}
+Now  there are two harmonic oscillating electrons in the well. At first these are assumed to be non-interacting. The radial part of the Schrödinger equation for this situation reads
+\begin{equation}
+\left(-\frac{\hbar^2}{2m}\frac{d^2}{dr_1^2} \right)
+\end{equation}
+
+\subsection{Preservation of orthogonality and dot product in unitary transform}
+
+Consider a basis of vectors 
+\begin{equation}
+\vb{v}_i = \begin{bmatrix}
+v_{i1} \\
+\vdots \\
+v_{in}
+\end{bmatrix}
+\end{equation}
+We assume that the basis is orthogonal, that is
+\begin{equation}
+\vb{v}_j^T\vb{v}_i=\delta_{ij}.
+\end{equation}
+It can be shown that a unitary transformation $\vb{w}_i = U\vb{v}_i$, where $U$ is a unitary matrix such that $U^TU=I$, the dot product and orthogonality is preserved. 
+\begin{align}
+\vb{w}_i^T\vb{w}_j &= (U\vb{v}_i)^T  U\vb{v}_i= \vb{v}_i^TU^T U\vb{v}_j = \vb{v}_i^T I \vb{v}_j \nonumber \\
+&\rightarrow \vb{w}_i^T\vb{w}_j = \vb{v}_i^T\vb{v}_j = \delta_{ij},
+\end{align}
+which means that both the dot product and orthogonality is preserved.
+
+\begin{thebibliography}{9}
+
+\bibitem{morten2}
+	Hiorth-Jensen, M.,
+	\emph{Project 2, Computational Physics I FYS3150/FYS4150},
+	University of Oslo,
+	2016.
+
+\end{thebibliography}
+
+\begin{appendix}
+\section{Laguerre Polynomials}
+\label{app:laguerre}
+Generally, the name Laguerre polynomials is used for solutions to 
+\begin{equation}
+x\frac{d^2y}{dx^2}+(\alpha+1-x)\frac{dy}{dx} + ny = 0.
+\end{equation}
+These polynomials, usually denoted, $L_0$, $L_1$, $L_2$ etc are a polynomial sequence which may be defined by the Rodriguez formula
+\begin{equation}
+L_n(x) = \frac{e^x}{n!}\frac{d^n}{dx^n}(e^{-x}x^n)=\frac{1}{n!}\left(\frac{d}{dx} -1 \right)x^n
+\end{equation}
+The first few Laguerre polynomials are shown in table \ref{tab:laguerre}.
+
+\begin{table}[ht]
+	\centering
+	\caption{The first few Laguerre polynomials}
+	\begin{tabular}{cl} \hline
+	$n$ & $L_n(x)$  \\ \hline
+	$0$ & $1$ \\
+	$1$ & $-x+1$ \\
+	$2$ & $\frac{1}{2}(x^2-4x+2)$ \\
+	$3$ & $\frac{1}{6}(-x^3+9x^2-18x+6)$ \\
+	$4$ & $\frac{1}{24}(x^4-16x^3+72x^2-96x+24$ \\
+	$5$ & $\frac{1}{120}(-x^5+25x^4-200x^3+600x^2-600x+120) $ \\
+	$6$ & $\frac{1}{720}(x^6-36x^5+450x^4-2400x^3+5400x^2-4320x+720)$ \\ \hline
+	\end{tabular}
+	\label{tab:laguerre}
+\end{table}
+
+\end{appendix}
+
 \end{document}

### 2016-09-30 Project2/article/article.tex
minor changes

@@
 \begin{equation}
-\left(-\frac{\hbar^2}{2m}\frac{d^2}{dr_1^2} \right)
+\left(-\frac{\hbar^2}{2m}\frac{d^2}{dr_1^2} -\frac{\hbar^2}{2m}\frac{d^2}{dr_2^2} +\frac{1}{2}kr_1^2+\frac{1}{2}kr_1^2 \right)u(r_1,r_2)=E^{(2)}u(r_1,r_2)
 \end{equation}
+
+We now introduce the relative coordinate $\vb{r} = \vb{r}_1 - \vb{r}_2$ and the center of mass coordinate $R=\frac{1}{2}(\vb{r}_1+\vb{r}_2)$ as well as assuming separation of variables $u(r,R)=\psi(r)\phi(R)$.
+When the energy is given as a sum of the relative energy and center of mass energy, $E^{(2)}=E_r+E_R$ we get
+\begin{equation}
+\label{eq:radialschr2}
+\left(-\frac{\hbar^2}{m}\frac{d^2}{dr^2}+\frac{1}{4}kr^2+\frac{\beta e^2}{r} \right)\psi(r)=E_r\psi(r)
+\end{equation}
+when also including the repulsive Coulomb interaction, $V(r) = \beta e^2/\abs{\vb{r}_1+\vb{r}_2}= \beta e^2/r$.
+By once again introducing a dimensionless variable $\rho=r/\alpha$ with $\alpha=\hbar^2/(m\beta e^2)$.
+By, in addition, including the relative angular frequency $\omega_r^2=(\frac{2\omega}{2\hbar}\alpha^2)^2$ and an energy parameter $\lambda=\frac{m\alpha^2}{\hbar^2}E_r$, we arrive at an equation comparable to \ref{eq:schr3}
+\begin{equation}
+-\frac{d\psi(\rho)}{d\rho^2}+\left(\omega^2_r\rho^2 + \frac{1}{\rho} \right) \psi(\rho)= \lambda\psi(\rho)
+\end{equation}
+It is easy to see that the only difference from the situation with two electrons and a single electron is that the potential is changed from $\rho^2$ to $\omega_r^2\rho^2+1/\rho$.
+In discretization, this is achieved by changing the diagonal of the matrix in \ref{eq:mat1} to $d_i = \frac{2}{hbar^2} \omega_r^2\rho^2+\frac{1}{\rho_i}$.
 

### 2016-09-30 Project2/jacobiMethod.cpp
Updated method file

@@ -4,6 +4,41 @@
 
 using namespace std;
 
+arma::mat constructA(double &rho_min, double &rho_max, int n, int interacting, double omega_r) {
+
+    // Step size
+    double h = (rho_max - rho_min)/(n+1);
+
+    // Constants in tri-diagonal elements
+    arma::mat A = arma::zeros<arma::mat>(n,n);
+    double d = 2/(h*h);
+    double e = -1/(h*h);
+
+    // Declaring potential variable
+    double V;
+
+    // Creating the tridiagonal matrix
+    for (int i = 0; i < (n); i++) {
+        double rho = (i+1)*h;
+
+        // Potential or not?
+        if (interacting){
+            V = Omega_r*Omega_r*rho*rho + 1/(rho);
+        } else {
+            V = rho*rho;
+        }
+
+       // Inputting potential on diagonal
+       A(i,i) = d + V;
+       if (i < n-1){
+           A(i,i+1) = A(i+1,i) = e;
+        }
+    }
+
+    // Return constructed matrix
+    return A;
+}
+
 double maxOffDiagonals(arma::mat &A, int &k, int &l, int n) {
 
     double max;
@@ -28,14 +63,12 @@ double maxOffDiagonals(arma::mat &A, int &k, int &l, int n) {
                 l = j;
             }
         }
-        cout << "Largest off-diagonal element is: " << maxOffDiagonals(A, k, l, n) << endl;
-
     }
-
     return max;
 }
 
 void jacobiRotation(arma::mat &A, arma::mat &R, int &k, int &l, int n) {
+
     // A is the input matrix and S is the matrix with eigenvectors after
     // enough rotations
 
@@ -119,20 +152,153 @@ void jacobiRotation(arma::mat &A, arma::mat &R, int &k, int &l, int n) {
 
 }
 
+void writeToFile(double rho_max , double rho_min, int n, arma::mat &R, double Omega_r,int lowestvalueindex) {
+    ofstream ofile;
+    ofile.open("dataNoint.txt");
+    ofile <<"Omega_r " << setw(15) << setprecision(8) << Omega_r << endl;
+    ofile <<"rho_min " << setw(15) << setprecision(8) << rho_min << endl;
+    ofile <<"rho_max " << setw(15) << setprecision(8) << rho_max << endl;
+    ofile <<"n  " << setw(20) << n << endl;
[…]

### 2016-09-30 Project2/jacobiMethod.cpp
Modified header file

@@ -1,6 +1,9 @@
 #include <iostream>
 #include <math.h>
 #include <armadillo>
+#include <vector>
+#include <assert.h>
+#include <iomanip>
 
 using namespace std;
 

### 2016-09-30 Project2/jacobimethod.h

@@ -1,9 +1,14 @@
 #ifndef JACOBIMETHOD_H
 #define JACOBIMETHOD_H
 
-double maxOffDiagonals(arma::mat &A, int &k, int &l, int n);
+double maxOffDiagonals(arma::mat &A, int &k, int &l, int n,double Omega_r);
 void jacobiRotation(arma::mat &A, arma::mat &R, int &k, int &l, int n);
-void jacobiMethod(arma::mat &A, arma::mat &R, int &k, int &l, int n);
-
+void jacobiMethod(arma::mat &A, arma::mat &R, int n);
+void jacobiEigTest(arma::mat &A, arma::mat &R, int n);
+arma::mat constructA(double &rho_min, double &rho_max,int n,int interacting);
+void jacobiOrthogTest(arma::mat &A,arma::mat &R,int n);
+void tests(arma::mat &A,arma::mat &R,int n);
+void jacobiMaxoffTest(arma::mat &A,int n);
+void writeToFile(double rho_max , double rho_min, int n, arma::mat &R,double Omega_r,int lowestvalueindex);
 
 #endif // JACOBIMETHOD_H

### 2016-10-01 Project2/jacobiMethod.cpp
Everything works well now

@@ -27,9 +27,9 @@ arma::mat constructA(double &rho_min, double &rho_max, int n, bool interacting, 
 
         // Potential or not?
         if (interacting) {
-            V += omega_r*omega_r*rho*rho + 1/(rho);
+            V = omega_r*omega_r*rho*rho + 1/(rho);
         } else {
-            V += rho*rho;
+            V = rho*rho;
         }
 
        // Inputting potential on diagonal
@@ -158,20 +158,14 @@ void jacobiRotation(arma::mat &A, arma::mat &R, int &k, int &l, int n) {
 
 }
 
-void writeToFile(arma::mat &R, int n, string filename){
+void writeToFile(arma::mat &R, int n, string filename, int minIndex){
 
     // Unnecessary arguments? : double rho_max , double rho_min, int n,
     // , double Omega_r,int lowestvalueindex
 
     ofstream ofile;
-    ofile.open(filename);
-//    ofile <<"Omega_r " << setw(15) << setprecision(8) << Omega_r << endl;
-//    ofile <<"rho_min " << setw(15) << setprecision(8) << rho_min << endl;
-//    ofile <<"rho_max " << setw(15) << setprecision(8) << rho_max << endl;
-//    ofile <<"n  " << setw(20) << n << endl;
-//    ofile << "Eigenvector corresponding to lowest eigenvalue:" << endl;
-
-    int lowestvalueindex = 1;
+    ofile.open(filename.c_str());
+    int lowestvalueindex = minIndex;
 
     for (int i = 0; i < n; i++) {
         ofile << R(i, lowestvalueindex) << endl;
@@ -181,13 +175,7 @@ void writeToFile(arma::mat &R, int n, string filename){
      ofile.close();
 }
 
-void jacobiMethod(arma::mat &A, arma::mat &R, int n) {
-
-    // If interacting, true
-    // bool interacting = false;
-
-    // Angular frequency
-    //double omega_r = 0.01;
+void jacobiMethod(arma::mat &A, arma::mat &R, int n, int &minIndex) {
 
     // Tolerance for the non-diagonals
     double eps = 1.0E-8;
@@ -214,21 +202,21 @@ void jacobiMethod(arma::mat &A, arma::mat &R, int n) {
 
     }
 
-    // cout << "No. of iterations: " << iterations << endl;
+    // Finding column where smallest eigenvalue is.
[…]

### 2016-10-01 Project2/jacobimethod.h

@@ -3,10 +3,10 @@
 
 double maxOffDiagonals(arma::mat &A, int &k, int &l, int n,double Omega_r);
 void jacobiRotation(arma::mat &A, arma::mat &R, int &k, int &l, int n);
-void jacobiMethod(arma::mat &A, arma::mat &R, int n);
+void jacobiMethod(arma::mat &A, arma::mat &R, int n, int &minIndex);
 arma::mat constructA(double &rho_min, double &rho_max,int n, bool interacting, double omega_r);
 
-void writeToFile(arma::mat &R, int n, std::string filename);
+void writeToFile(arma::mat &R, int n, std::string filename, int minIndex);
 
 void jacobiEigTest();
 void jacobiMaxOffTest();

### 2016-10-01 Project2/main.cpp

@@ -2,14 +2,16 @@
 #include <math.h>
 #include <armadillo>
 #include <string>
+#include <ctime>
 
-#include <jacobimethod.h>
+#include "jacobimethod.h"
+#include "jacobiMethod.cpp"
+
 
 using namespace std;
 
 int main(int argc, char *argv[])
-{ 
-    /*
+{  /*
     cout << argv[1] << endl;
     cout << argv[2] << endl;
     cout << argv[3] << endl;
@@ -30,7 +32,7 @@ int main(int argc, char *argv[])
              << endl;
     } else {
 
-        cout << "DU GREIDE DET!!" << endl;
+        cout << "Computing. Please wait." << endl;
 
         // Parsing n
         int n = atoi(argv[1]);
@@ -53,15 +55,27 @@ int main(int argc, char *argv[])
         arma::mat A(n,n);
         arma::mat R; R.eye(n,n);
 
+        // Ground state index
+        int minIndex;
+
+        // Timer start
+        clock_t start = clock();
+
         // Constructing second derivative approximation matrix
         A = constructA(rho_min, rho_max, n, interacting, omega_r);
 
         // Computing ..
-        jacobiMethod(A, R, n);
+        jacobiMethod(A, R, n, minIndex);
 
         // Writing data to file
-        writeToFile(R, n, output_file);
+        writeToFile(R, n, output_file, minIndex);
 
+        // Timer stop
+        clock_t stop = clock();
+
+        cout << "Computation finished. Elapsed time: ";
+        cout << ( stop - start ) / (double) CLOCKS_PER_SEC << " seconds" << endl;
+        cout << "Data written to " << output_file.c_str() << "\n" << endl;
     }
 
 }

### 2016-10-01 Project2/plotomegas.py

@@ -1,0 +1,25 @@
+import sys
+import numpy as np 
+from matplotlib import pyplot as plt 
+
+file = open('data.txt')
+
+n = 250
+
+allLines = file.readlines()
+
+u = np.zeros(n)
+
+i = 0
+for line in allLines:
+	u[i] = float(line)
+	i += 1
+	if i >= n:
+		break
+
+file.close()
+
+u = np.abs(u)**2
+
+plt.plot(u)
+plt.show()

### 2016-10-01 Project2/plotomegas.py
Python scripts and figures

@@ -1,25 +1,0 @@
-import sys
-import numpy as np 
-from matplotlib import pyplot as plt 
-
-file = open('data.txt')
-
-n = 250
-
-allLines = file.readlines()
-
-u = np.zeros(n)
-
-i = 0
-for line in allLines:
-	u[i] = float(line)
-	i += 1
-	if i >= n:
-		break
-
-file.close()
-
-u = np.abs(u)**2
-
-plt.plot(u)
-plt.show()

### 2016-10-01 Project2/scripts/interaction.py

@@ -1,0 +1,49 @@
+'''
+Plot of a system w/o interaction and one w/ interaction
+omega = 0.01, N = 200
+'''
+
+import sys
+import numpy as np 
+from matplotlib import pyplot as plt 
+
+n = 200
+
+file1 = open('../data/omega0.01int.txt')
+file2 = open('../data/omega0.01noint.txt')
+
+allLines1 = file1.readlines()
+allLines2 = file2.readlines()
+
+u1 = np.zeros(n)
+u2 = np.zeros(n)
+
+i = 0
+for line in allLines1:
+	u1[i] = float(line)
+	i += 1
+	if i >= n:
+		break
+
+file1.close()
+
+i = 0
+for line in allLines2:
+	u2[i] = float(line)
+	i += 1
+	if i >= n:
+		break
+
+file2.close()
+
+u1 = np.abs(u1)**2
+u2 = np.abs(u2)**2
+
+plt.title("Coulomb and no-interacting comparison "\
+	 + r'$\omega_r = 0.01, N=200$')
+plt.plot(u1, label="With Coulomb interaction")
+plt.plot(u2, label="No interaction")
+plt.xlabel(r'$\rho$', fontsize=20)
+plt.ylabel(r'$ \mid{u(\rho)}\mid^2$', fontsize=20)
+plt.legend(loc=2)
+plt.show()

### 2016-10-01 Project2/scripts/omega.py

@@ -1,0 +1,79 @@
+'''
+A comparison of ground state with different angular momentum
+omega = 0.01, 0.1, 1, 5
+N = 200
+rho = 5 for all
+'''
+
+import numpy as np
+from matplotlib import pyplot as plt
+
+n = 200
+
+file1 = open('../data/omega0.01int.txt')
+file2 = open('../data/omega0.5int.txt')
+file3 = open('../data/omega1int.txt')
+file4 = open('../data/omega5int.txt')
+
+allLines1 = file1.readlines()
+allLines2 = file2.readlines()
+allLines3 = file3.readlines()
+allLines4 = file4.readlines()
+
+u1 = np.zeros(n)
+u2 = np.zeros(n)
+u3 = np.zeros(n)
+u4 = np.zeros(n)
+
+i = 0
+for line in allLines1:
+	u1[i] = float(line)
+	i += 1
+	if i >= n:
+		break
+
+file1.close()
+
+i = 0
+for line in allLines2:
+	u2[i] = float(line)
+	i += 1
+	if i >= n:
+		break
+
+file2.close()
+
+i = 0
+for line in allLines3:
+	u3[i] = float(line)
+	i += 1
+	if i >= n:
+		break
+
+file3.close()
+
+i = 0
+for line in allLines4:
+	u4[i] = float(line)
+	i += 1
+	if i >= n:
+		break
+
+file4.close()
+
+u1 = np.abs(u1)**2
+u2 = np.abs(u2)**2
+u3 = np.abs(u3)**2
+u4 = np.abs(u4)**2
+
+plt.plot(u1, label=r"$\omega_r=0.01$")
+plt.plot(u2, label=r"$\omega_r=0.5$")
+plt.plot(u3, label=r"$\omega_r=1$")
+plt.plot(u4, label=r"$\omega_r=5$")
+
+plt.legend(loc=2)
+plt.title("Ground state with different angular momentum")
+plt.xlabel(r'$\rho$', fontsize=20)
+plt.ylabel(r'$ \mid{u(\rho)}\mid^2$', fontsize=20)
+
+plt.show()

### 2016-10-01 Project2/article/article.tex
Getting closer

@@
   \hrulefill\small{ FYS3150: Computational Physics }\hrulefill}
+
+\author[Winther-Larsen \& Svalheim]{Sebastian G. Winther-Larsen \\ 
+Trygve L. Svalheim \\
+\href{https://github.com/gregwinther/FYS3150/}{\texttt{github.com/gregwinther}}}
 
@@
 By finding the eigenvectors, or the energies, corresponding to the eigenvalues $\lambda_i$  equation \ref{eq:schr3} above, one has found the wave function that describe the behavior of the electron for a particular state $i$.
-To have something to compare against it is useful to state the exact, analytical wave function for the three lowest states.
-The general form of these radial wave functions is
+It is of interest to also state the exact, analytical wave function for the three lowest states, because the Schrödinger equation in general is only analytically solvable for a select few potentials, and the harmonic oscillator potential is amongst these.
+The general form of these radial wave function solutions is
 \begin{equation}
@@
 
+This proof is important to the algorithm which is described in the next section.
+As a result of Abel and Ruffini's theorem, regarding the impossibility of a solution to equations degree five and higher, a discretized problema s described herein is unsolvable by conventional, analytical means, and an iterative process is thus necessary.
+
+\section{Algorithm}
+
+\subsection{Jacobi's method}
+
+First, consider a simple two-dimensional, symmetric matrix
+\begin{equation}
+A = \begin{bmatrix}
+a_{11} & a_{12} \\
+a_{21} & a_{22}
+\end{bmatrix}
+\end{equation}
+where, because the matrix is symmetric, $a_{12}=a_{21}$. Second, consider an orhtogonal rotation matrix
+\begin{equation}
+R = \begin{bmatrix}
+\cos{\theta} & -\sin{\theta} \\
+\sin{\theta} & \cos{\theta}
+\end{bmatrix}
+\end{equation}
+which when multiplied with some vector $\vb{v}$, rotates the vector by an angle $\theta$.
+Now, we simplify the notation a bit by substituting the trigonometric functions $c=\cos{\theta}$ and $s=\sin{\theta}$.
+We can now write the similarity transform as
+\begin{equation}
+R^{-1}AR=\begin{bmatrix}
+a_{11}c^2+2a_{12}cs + a_{22}s^2 & (a_{22}-a_{11})cs+a_{12}(c^2-s^2) \\
+(a_{22}-a_{11})cs + a_{12}(c^2-s^2) & a_{11}s^2-2a_{12}s+a_{22}c^2
+\end{bmatrix}
+\end{equation}
+Notice that this matrix is symmetric as well.
+We want to roate in such a manner that the offdiagonal elemtes of the new matrix are zero.
+Hence we need to solve $(a_{22}-a_{11})cs + a_{12}(c^2-s^2)=0$, which gives
+\begin{equation}
+\tau = \cot{2\theta} = \frac{a_{11}-a_{22}}{2a_{12}}
+\end{equation}
+
+Now we can express all the trigonometric functions in terms of $\tau$.
+Setting $t = \tan{\theta}$, $c=\cos{\theta}=1/\sqrt{t^2+1}$ and $s=ct$.
+This yields $\cot{2\theta}=1/2(\cot{\theta}-\tan{\theta}) = 1/2(2^{-1}-t$, all boiling down the quadratic equation
+\begin{equation}
+t^2-2\tau t-1 =0
+\end{equation}
+which has two roots. Of these two roots we pickk the smallest in order to ensure maximum  numerical stability. To put it another way, we rotate a tiny angle rather than a large one at each iteration.
+
+The eigenvalues can now be expressed in terms of $t$. The diagonalized matrix will look like this
+\begin{equation}
+R^{-1}AR = \begin{bmatrix}
+a_{11} & 0 \\
+0 & a_{22}
+\end{bmatrix}
+\end{equation}
+Solving the characteristic equation provides the eigenvalues, while an inverse rotation of the identy matrix supplies the eigenvectors\footnote{Bear in mind that the eigenvectors and the eigenvalues will in this particular study represent the wave function and the corresponding energy levels respectively.}.
+
+The algorithm for the $2\times2$-case can easily be extended to a general case by using a $n\times n$-matrix instead. Such a matrix is called the Givens rotation matrix and looks like this
+\begin{equation}
+G(i,j,\theta) = \begin{bmatrix}
+1 & \hdots & 0 &\hdots & 0 & \hdots & 0 \\
+    \vdots & \ddots & \vdots & {} & \vdots & {} & \vdots \\
+    0 & \hdots & c &\hdots & -s & \hdots & 0 \\
+    \vdots & {} & \vdots & \ddots & \vdots & {} & \vdots \\
+    0 & \hdots & s &\hdots & c & \hdots & 0 \\
+    \vdots & {} & \vdots & {} & \vdots & \ddots & \vdots \\
+0 & \hdots & 0 &\hdots & 0 & \hdots & 1
+\end{bmatrix}
+\end{equation}
+where $c= \cos{\theta}$ and $s=\sin{\theta}$.
+
+Performing the corresponding tranform as the one above for a particular $i=k$ and $j=l$ yields the followin set of equations
+\begin{align}
+\begin{split}
+  a^{\prime}_{hk} &= a^{\prime}_{kh} = ca_{hk} - sa_{hl} \\
+  a^{\prime}_{hl} &= a^{\prime}_{lh} = ca_{hl} - sa_{hk} \\
+  a^{\prime}_{kl} &= a^{\prime}_{lk} = (c^2 - s^2)a_{kl} + sc(a_{kk} - a_{ll}) = 0\\
+  a^{\prime}_{kk} &= c^2a^{\prime}_{kk} + s^2a_{ll} - 2sca_{kl}\\
+  a^{\prime}_{ll} &= s^2a^{\prime}_{kk} + c^2a_{ll} + 2sca_{kl}
+\end{split}
+\end{align}
+
+\subsection{Implementation}
+HER SKAL DET NOE OM HVORDAN KODEN SER UT e.l.
+
+\section{Results}
+STORT SETT BARE KOMMENTARER TIL FIGURENE
+
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=0.9\textwidth]{../figures/interaction.png}
+	\caption{Plot showing the difference between the ground state with and without Coulomb repulsion.}
+\end{figure}
+
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=0.9\textwidth]{../figures/omega.png}
+	\caption{Plot of the ground state with Coulomb repulsion and varying angular momentum}
+\end{figure}
+
+\section{Summary Remarks}
+KONKLUSJON OG OPPSUMMERING
+
 \begin{thebibliography}{9}
@@
 	2016.
+
+\bibitem{abel}
+	Abel, N.H.,
+	Memoire sur le équations algébriques, ou l'on démontre l'impossibilité de la résolution de l'équation générale de cinquiéme degré,
+	In Sylow, L. and Lie, S., 
+	\emph{Æuvres Complètes de Niels Henrik Abel}, 2nd ed.,
+	Grøndahl \& Søn, pp. 28-33,
+	1881.
 

### 2016-10-01 Project2/scripts/interaction.py

@@ -9,6 +9,8 @@ from matplotlib import pyplot as plt
 
 n = 200
 
+rho = np.linspace(0,5,n)
+
 file1 = open('../data/omega0.01int.txt')
 file2 = open('../data/omega0.01noint.txt')
 
@@ -36,14 +38,17 @@ for line in allLines2:
 
 file2.close()
 
+u1 = u1[::-1]
+u2 = u2[::-1]
+
 u1 = np.abs(u1)**2
 u2 = np.abs(u2)**2
 
 plt.title("Coulomb and no-interacting comparison "\
-	 + r'$\omega_r = 0.01, N=200$')
-plt.plot(u1, label="With Coulomb interaction")
-plt.plot(u2, label="No interaction")
+	 + r'$\omega_r = 0.01$')
+plt.plot(rho, u1, label="With Coulomb repulsion")
+plt.plot(rho, u2, label="No interaction")
 plt.xlabel(r'$\rho$', fontsize=20)
 plt.ylabel(r'$ \mid{u(\rho)}\mid^2$', fontsize=20)
-plt.legend(loc=2)
+plt.legend()
 plt.show()

### 2016-10-01 Project2/scripts/omega.py

@@ -10,6 +10,8 @@ from matplotlib import pyplot as plt
 
 n = 200
 
+rho = np.linspace(0,5,n)
+
 file1 = open('../data/omega0.01int.txt')
 file2 = open('../data/omega0.5int.txt')
 file3 = open('../data/omega1int.txt')
@@ -61,17 +63,22 @@ for line in allLines4:
 
 file4.close()
 
+u1 = u1[::-1]
+u2 = u2[::-1]
+u3 = u3[::-1]
+u4 = u4[::-1]
+
 u1 = np.abs(u1)**2
 u2 = np.abs(u2)**2
 u3 = np.abs(u3)**2
 u4 = np.abs(u4)**2
 
-plt.plot(u1, label=r"$\omega_r=0.01$")
-plt.plot(u2, label=r"$\omega_r=0.5$")
-plt.plot(u3, label=r"$\omega_r=1$")
-plt.plot(u4, label=r"$\omega_r=5$")
+plt.plot(rho, u1, label=r"$\omega_r=0.01$")
+plt.plot(rho, u2, label=r"$\omega_r=0.5$")
+plt.plot(rho, u3, label=r"$\omega_r=1$")
+plt.plot(rho, u4, label=r"$\omega_r=5$")
 
-plt.legend(loc=2)
+plt.legend()
 plt.title("Ground state with different angular momentum")
 plt.xlabel(r'$\rho$', fontsize=20)
 plt.ylabel(r'$ \mid{u(\rho)}\mid^2$', fontsize=20)

### 2016-10-04 Project2/article/article.tex
Inncluded Trygve's stuff. Added to introduction

@@
 \begin{abstract}
-THIS IS NOT THE ABSTRACT FOR THIS PROJECT :)
-A one-dimensional version of the Poisson equation with Dirichlet boundary conditions is solved using two different algorithms.
-The two algorithms employed are the tridiagonal matrix algorithm and the LU decomposition method.
-We find that a fine-tuned version of the tridiagonal matrix algorithm is $10^5$ times faster than the general LU decomposition method and $10$ times as precise.
+In this project we have solved the three-dimensional Schrödinger equation for two particles with and without a repulsive Coulomb potential in an harmonic oscillator well.
+By discretizing the equation for one and later two electrons we could solve it numerically as an eigenvalue problem using the Jacobi method.
+By implementing the algorithm we observe that the numerical solution coincides well with the analytical prediction.
+Furthermore, the only difference between the cases with and without Coulomb repulsion is the potential term, hence we can conclude that our algorithm can be modified for any potential.
 \end{abstract}
@@
 The equation is reformulated to a discretized in order to be represented as a matrix eigenvalue problem that can be solved with Jacobi's method.
+Third, a brief discussion of how to implement the algorithm.
+To keep this article brief and to the point we have forgone to include any code snippets, but everything can be found at the github address stated above.
+Fourth and lastly, we present the results and give some concluding remarks.
 
@@
 To simplify the problem, spherical symmetry is assumed.
+First, we give a brief theoretic outline of the radial Scchrödinger equation in order to av a theoretical base for the study.
+Second, we give a thorough description of how an algorithm for an eigenvalue problem must function.
+Third, we
 
@@
 
-Performing the corresponding tranform as the one above for a particular $i=k$ and $j=l$ yields the followin set of equations
+Performing the corresponding tranform as the one above for a particular $i=k$ and $j=l$ yields the following set of equations
 \begin{align}
+\label{eq:rotation}
 \begin{split}
@@
 \subsection{Implementation}
-HER SKAL DET NOE OM HVORDAN KODEN SER UT e.l.
+The implementation of the algorithm was done using the armadillo library.
+We start by constructing the matrix A, diagonalizing it using the Jacobi method and finally writing the eigenvector-results to file, while running unit tests along the way.
+
+\subsubsection{Constructing matrix A} In order to solve the eigenvalue problem we first construct the matrix A by filling in the elements along the $n\times n$-matrix' diagonal with the constant terms as well as the potential corresponding to the case we are studying.
+\subsubsection{Diagonalizing using the Jacobi Method} We start off by making an algorithm for locating the maximum value off the diagonal in the matrix.
+This is done in order to check if they have gone to zero, which is what the Jacobi rotation is trying to do.
+When all values off the diagonal have gone to zero within a reasonable tolerance, we are finished.
+
+The \texttt{jacobiRotation} function works by calculating equation \ref{eq:rotation} straightforward and updating all matrix elements. Lastly the eigenvectors are saved to the matrix R.
 
 \section{Results}
-STORT SETT BARE KOMMENTARER TIL FIGURENE
 
 \begin{figure}[ht]
+	\label{fig:ConoCo}
 	\centering
@@
 
+Knowing that the $|u(\rho)|^2$ describes the position of the particles, it is easy to see that the Coulomb repulsion results in the two particles being farther apart.
+This conclusion is quite intuitive when describing two repulsing particles.
+From \ref{fig:ConoCo} we observe this effect, as the "no interaction"-curve is sharper that with Coulomb repulsion.
+
 \begin{figure}[ht]
+	\label{fig:Co}
 	\centering
@@
 
+In from figure \ref{fig:Co} we are looking at the case of Coulomb repulsion for varying frequencies $\omega_i$.
+Right away, we see that by increasing the oscillator frequency, we get sharper curves, meaning a more well defined position.
+This follows by the fact that $\omega_r$ determines the strength of the well resulting in the two electrons being forced closer together.
+
 \section{Summary Remarks}
-KONKLUSJON OG OPPSUMMERING
+solving the Schrödinger equation as an eigenvalue problem.
+We have concluded that the behaviour of the two particles examined coincide well with that predicted from quantum mechanics.
+Furthermore, we have successfully implemented unit tests into our code, allowing us to locate errors in our code more efficiently.
+We have however not discussed the efficiency of our code, but can assume from what we learned in project 1 that a specialized algorithm for the given matrix would have resulted in a faster program.
 

### 2016-10-20 Project3/article/article.tex
Changed folder structure: article, date etc. Included stuff from trygvels

--- 2016-09-30 Project2/article/article.tex
@@
 
-\title[Schrödinger's Equation in 3D]{Schrödinger's Equation in a Three-Dimensional Harmonic Oscillator Well \\
+\title[Simulation of Solar System]{Simulation of Solar System Using Velocity Verlet Method \\
   \hrulefill\small{ FYS3150: Computational Physics }\hrulefill}
+
+\author[Winther-Larsen \& Svalheim]{Sebastian G. Winther-Larsen \\ 
+Trygve L. Svalheim \\
+\href{https://github.com/gregwinther/FYS3150/}{\texttt{github.com/gregwinther}}}
 
@@
 \begin{abstract}
-THIS IS NOT THE ABSTRACT FOR THIS PROJECT :)
-A one-dimensional version of the Poisson equation with Dirichlet boundary conditions is solved using two different algorithms.
-The two algorithms employed are the tridiagonal matrix algorithm and the LU decomposition method.
-We find that a fine-tuned version of the tridiagonal matrix algorithm is $10^5$ times faster than the general LU decomposition method and $10$ times as precise.
+Two, three, multi. With Euler-Cromer and then Velocity Verlet
 \end{abstract}
@@
 
-The aim of this study is to solve Schrödinger's equation for two electrons in a three-dimensional  harmonic oscillator well with and without repulsive Coulomb interaction.
-The equation is reformulated to a discretized in order to be represented as a matrix eigenvalue problem that can be solved with Jacobi's method.
+\section{Theory}
 
-\section{Theory}
-We assume that electrons confined in a small area move in a three-dimensional harmonic oscillator potential and repel each other via the static Coulomb interaction.
-To simplify the problem, spherical symmetry is assumed.
+\section{Algorithm}
 
-\subsection{A single electron}
-The radial part of Schrödinger's equation for a single electron reads
-\begin{equation}
-\label{eq:schr1}
--\frac{\hbar^2}{2m}\left(\frac{1}{r^2}\frac{d}{dr}r^2\frac{d}{dr}-\frac{l(l+1)}{r^2} \right)R(r) + V(r)R(r) = ER(r)
-\end{equation}
-where $V(r)$ is the harmonic oscillator potential\footnote{One can in theory insert any potential into the Schrödinger equation, but in this case the harmonic oscillator potential is employed.}, $V(r)=\frac{1}{2}kr^2$ where $k=m\omega^2$, and consequently $E$ is the energy of the harmonic oscillator in three dimensions.
-The angular frequency of the oscillator is given by $\omega$ and is related to the energy of the system as follows
-\begin{equation}
-E_{nl}=\hbar\omega\left(2n + l  + \frac{3}{2} \right), \quad n,l \in \mathbb{N} 
-\end{equation}
-However, we will consider cases with zero angular momentum only ($l=0$).
-Moreover, by a somewhat lengthy substitutive exercise will culminate in a rewriting of Schrödinger's equation in \ref{eq:schr1} to a dimensionless form
-\begin{equation}
-\label{eq:schr2}
--\frac{d^2}{d\rho^2}u(\rho) + \rho^2u(\rho) = \lambda u(\rho)
-\end{equation}
-where $u(r) = R(r)/r$ and $\rho = r/\alpha$ is a dimensionless parameter for length.
-The constant $\alpha$ is $\alpha = \sqrt{\hbar^2/mk}$ and the energy parameter (eigenvalues) are $\lambda = 2m\alpha^2E/\hbar^2$.
-The three first eigenvalues will be $\lambda_0= 3$, $\lambda_1 = 7$ and $\lambda_0=11$.
+\section{Results}
 
-The second derivative in equation \ref{eq:schr2} will be approximated by the well-known central finite difference method for the second derivative,
-\begin{equation}
-\frac{du^2}{d\rho^2}=\frac{u(\rho+h)-2u(\rho)+u(\rho-h)}{h^2}+\mathcal{O}(h^2)
-\end{equation}
-where $h = (\rho_N-\rho_0)/h$ is the step size. By denoting $u(\rho_0+ih)$ by $u_i$, equation \ref{eq:schr2} can be further rewritten to
-\begin{align}
--\frac{1}{h^2}u_{i-1}+\left(\frac{2}{h^2} + \rho_i^2 \right)u_i-\frac{1}{h^2}u_{i+1}=\lambda u_i \label{eq:schr3}\\
-A\vb{u} = \lambda \vb{u} \label{eq:mat1}
-\end{align} 
-where the matrix in \ref{eq:mat1} is
-\begin{equation}
-\begin{bmatrix}
-\frac{1}{h^2} + \rho_i^2 & -\frac{1}{h^2} & 0 & 0 & 0 & \dots & 0 \\
--\frac{1}{h^2} & \frac{1}{h^2} + \rho_i^2 & -\frac{1}{h^2} & 0 & 0 & \dots & 0 \\
-0 & -\frac{1}{h^2} & \frac{1}{h^2} + \rho_i^2 & -\frac{1}{h^2} & 0 & \dots & 0  \\
-\vdots & \vdots & \vdots &\ddots & \ddots & \ddots & \vdots \\
-0 & 0 & 0 & \dots &  -\frac{1}{h^2} & \frac{1}{h^2} + \rho_i^2 & -\frac{1}{h^2} \\
-0 & 0 & 0 & 0 & \dots & -\frac{1}{h^2} & \frac{1}{h^2} + \rho_i^2 
-\end{bmatrix}
-\end{equation}
+\section{Discussion}
 
-By finding the eigenvectors, or the energies, corresponding to the eigenvalues $\lambda_i$  equation \ref{eq:schr3} above, one has found the wave function that describe the behavior of the electron for a particular state $i$.
-To have something to compare against it is useful to state the exact, analytical wave function for the three lowest states.
-The general form of these radial wave functions is
-\begin{equation}
-\label{eq:radialschr1}
-R_{n,l=0}(r)=N_ne^{-\frac{m\omega}{2\hbar}r^2}\sqrt{L_n}\left(-\frac{m\omega}{2\hbar} r^2 \right)
-\end{equation}
-where $L_n$ refers to a particular Laguerre polynomial.
-More on the Laguerre polynomials can be found in appendix \ref{app:laguerre}.
-Reevaluating this function for the three lowest energy states and normalizing yields the following equations
-\begin{align}
-\abs{u_0(\rho)}^2 &= \frac{4}{\sqrt{\pi}}\rho^2e^{-\rho^2} \\
-\abs{u_1(\rho)}^2 &= \frac{8}{3\sqrt{\pi}}\rho^2\left( \frac{3}{2}-\rho^2 \right)e^{-\rho^2} \\
-\abs{u_2(\rho)}^2 &= \frac{8}{15\sqrt{\pi}}\rho^2\left( \frac{15}{4}-5\rho^2 +\rho^4\right)e^{-\rho^2}
-\end{align}
-It is only sensible to deal with the absolute squared values of the wave functions, because this is a probability density function for the position of the electron.
-
-\subsection{Two electrons}
-Now  there are two harmonic oscillating electrons in the well. At first these are assumed to be non-interacting. The radial part of the Schrödinger equation for this situation reads
-\begin{equation}
-\left(-\frac{\hbar^2}{2m}\frac{d^2}{dr_1^2} -\frac{\hbar^2}{2m}\frac{d^2}{dr_2^2} +\frac{1}{2}kr_1^2+\frac{1}{2}kr_1^2 \right)u(r_1,r_2)=E^{(2)}u(r_1,r_2)
-\end{equation}
-
-We now introduce the relative coordinate $\vb{r} = \vb{r}_1 - \vb{r}_2$ and the center of mass coordinate $R=\frac{1}{2}(\vb{r}_1+\vb{r}_2)$ as well as assuming separation of variables $u(r,R)=\psi(r)\phi(R)$.
-When the energy is given as a sum of the relative energy and center of mass energy, $E^{(2)}=E_r+E_R$ we get
-\begin{equation}
-\label{eq:radialschr2}
-\left(-\frac{\hbar^2}{m}\frac{d^2}{dr^2}+\frac{1}{4}kr^2+\frac{\beta e^2}{r} \right)\psi(r)=E_r\psi(r)
-\end{equation}
-when also including the repulsive Coulomb interaction, $V(r) = \beta e^2/\abs{\vb{r}_1+\vb{r}_2}= \beta e^2/r$.
-By once again introducing a dimensionless variable $\rho=r/\alpha$ with $\alpha=\hbar^2/(m\beta e^2)$.
-By, in addition, including the relative angular frequency $\omega_r^2=(\frac{2\omega}{2\hbar}\alpha^2)^2$ and an energy parameter $\lambda=\frac{m\alpha^2}{\hbar^2}E_r$, we arrive at an equation comparable to \ref{eq:schr3}
-\begin{equation}
--\frac{d\psi(\rho)}{d\rho^2}+\left(\omega^2_r\rho^2 + \frac{1}{\rho} \right) \psi(\rho)= \lambda\psi(\rho)
-\end{equation}
-It is easy to see that the only difference from the situation with two electrons and a single electron is that the potential is changed from $\rho^2$ to $\omega_r^2\rho^2+1/\rho$.
-In discretization, this is achieved by changing the diagonal of the matrix in \ref{eq:mat1} to $d_i = \frac{2}{hbar^2} \omega_r^2\rho^2+\frac{1}{\rho_i}$.
-
-\subsection{Preservation of orthogonality and dot product in unitary transform}
-
-Consider a basis of vectors 
-\begin{equation}
-\vb{v}_i = \begin{bmatrix}
-v_{i1} \\
-\vdots \\
-v_{in}
-\end{bmatrix}
-\end{equation}
-We assume that the basis is orthogonal, that is
-\begin{equation}
-\vb{v}_j^T\vb{v}_i=\delta_{ij}.
-\end{equation}
-It can be shown that a unitary transformation $\vb{w}_i = U\vb{v}_i$, where $U$ is a unitary matrix such that $U^TU=I$, the dot product and orthogonality is preserved. 
-\begin{align}
-\vb{w}_i^T\vb{w}_j &= (U\vb{v}_i)^T  U\vb{v}_i= \vb{v}_i^TU^T U\vb{v}_j = \vb{v}_i^T I \vb{v}_j \nonumber \\
-&\rightarrow \vb{w}_i^T\vb{w}_j = \vb{v}_i^T\vb{v}_j = \delta_{ij},
-\end{align}
-which means that both the dot product and orthogonality is preserved.
-
-\begin{thebibliography}{9}
-
-\bibitem{morten2}
-	Hiorth-Jensen, M.,
-	\emph{Project 2, Computational Physics I FYS3150/FYS4150},
-	University of Oslo,
-	2016.
-
-\end{thebibliography}
-
-\begin{appendix}
-\section{Laguerre Polynomials}
-\label{app:laguerre}
-Generally, the name Laguerre polynomials is used for solutions to 
-\begin{equation}
-x\frac{d^2y}{dx^2}+(\alpha+1-x)\frac{dy}{dx} + ny = 0.
-\end{equation}
-These polynomials, usually denoted, $L_0$, $L_1$, $L_2$ etc are a polynomial sequence which may be defined by the Rodriguez formula
-\begin{equation}
-L_n(x) = \frac{e^x}{n!}\frac{d^n}{dx^n}(e^{-x}x^n)=\frac{1}{n!}\left(\frac{d}{dx} -1 \right)x^n
-\end{equation}
-The first few Laguerre polynomials are shown in table \ref{tab:laguerre}.
-
-\begin{table}[ht]
-	\centering
-	\caption{The first few Laguerre polynomials}
-	\begin{tabular}{cl} \hline
-	$n$ & $L_n(x)$  \\ \hline
-	$0$ & $1$ \\
-	$1$ & $-x+1$ \\
-	$2$ & $\frac{1}{2}(x^2-4x+2)$ \\
-	$3$ & $\frac{1}{6}(-x^3+9x^2-18x+6)$ \\
-	$4$ & $\frac{1}{24}(x^4-16x^3+72x^2-96x+24$ \\
-	$5$ & $\frac{1}{120}(-x^5+25x^4-200x^3+600x^2-600x+120) $ \\
-	$6$ & $\frac{1}{720}(x^6-36x^5+450x^4-2400x^3+5400x^2-4320x+720)$ \\ \hline
-	\end{tabular}
-	\label{tab:laguerre}
-\end{table}
-
-\end{appendix}
+\section{Conclusion}
 

### 2016-10-20 Project3/data/multiBodyPlot.m

@@ -1,0 +1,20 @@
+load positions.dat;         %  read data 
+figure
+hold on;
+
+view(45,45);
+max_val = max(max(positions(:,1:3:end)));
+axis([-max_val max_val -max_val max_val -max_val max_val]);
+for i=4:3:28
+    plot3(positions(:,i), positions(:,i+1), positions(:,i+2));
+end
+
+h = plot3(positions(1,:), positions(1,:),positions(1,:), '.k');
+
+
+for i=1:10:length(positions(:,1)) 
+    set(h, 'XData', positions(i,1:3:end), 'YData', positions(i,2:3:end),'ZData', positions(i,3:3:end));
+    drawnow;
+end
+
+axis equal

### 2016-10-20 Project3/helioplot.py

@@ -1,16 +1,0 @@
-''' 
-Simple plotting script
-'''
-
-import numpy as np;
-
-file1 = open('positions.dat')
-
-line = file1.readline()
-
-print(line)
-arraydings = np.fromstring(line, dtype=np.float, sep=" ")
-print(arraydings)
-
-
-file1.close()

### 2016-10-20 Project3/src/InitialConditions/multibody.cpp

@@ -19,16 +19,16 @@ void MultiBody::setupParticles(System &system) {
     double neptune_mass = 102.41 * pow(10, 24) / solar_mass;
     double pluto_mass	= 1.307 * pow(10, 22) / solar_mass;
 
-    Particle *sun = new Particle(vec3(3.5554E-03,  3.4437E-03, -1.5959E-04), vec3(-2.0393E-06,  6.8258E-06,  4.1904E-08), 1);
-    Particle *mercury = new Particle(vec3(-3.9067E-01, -5.3842E-02, 3.1326E-02), vec3(-1.8057E-03, -2.6625E-02, -2.0106E-03), mercury_mass);
-    Particle *venus = new Particle(vec3(2.9535E-01, -6.6297E-01, -2.6135E-02), vec3(1.8390E-02,  8.0510E-03, -9.5100E-04), venus_mass);
-    Particle *earth = new Particle(vec3(8.9232E-01,  4.5238E-01, -1.808E-04), vec3(-8.0392E-03,  1.5299E-02, -4.9539E-07), earth_mass);
-    Particle *mars = new Particle(vec3( 1.2012, -6.8454E-01, -4.3970E-02), vec3(7.5008E-03,  1.3339E-02,  9.5301E-05), mars_mass);
-    Particle *jupiter = new Particle(vec3(-5.4257, -4.8938E-01,  1.2337E-01), vec3(5.9024E-04, -7.1587E-03,  1.6542E-05), jupiter_mass);
-    Particle *saturn = new Particle(vec3(-2.2413,-9.7811703,  2.5927E-01), vec3(5.1320E-03, -1.2636E-03, -1.8228E-04), saturn_mass);
-    Particle *uranus = new Particle(vec3(1.8455E+01,  7.5787E+00, -2.1094E-01), vec3(-1.5228E-03,  3.4549E-03,  3.2562E-05), uranus_mass);
-    Particle *neptune = new Particle(vec3(2.82E+01, -9.9073, -4.4739E-01), vec3(1.0174E-03,  2.9810E-03, -8.4837E-05), neptune_mass);
-    Particle *pluto = new Particle(vec3(9.4392, -3.1817E+01,  6.7423E-01), vec3(3.0745E-03,  2.4227E-04, -9.1527E-04), pluto_mass);
[…]

### 2016-10-20 Project3/src/examples.cpp

@@ -7,12 +7,14 @@
 #include "Potentials/nopotential.h"
 #include "InitialConditions/twobody.h"
 #include "InitialConditions/threebody.h"
+#include "InitialConditions/multibody.h"
+
 #include <iostream>
 #include <cmath>
 
 
 void Examples::twoBodyProblem() {
-    double G = 4*M_PI*M_PI; // 6.67E-11 Nm/kg adjusted to .... blah blah
+    double G = 4*M_PI*M_PI; // 6.67E-11 Nm/kg adjusted to correct units
 
     System* twoBodySystem = new System();
     twoBodySystem->setIntegrator        (new EulerCromer(twoBodySystem));
@@ -24,12 +26,25 @@ void Examples::twoBodyProblem() {
 }
 
 void Examples::threeBodyProblem() {
-    /*
-     * This is where you should set up a three-body problem, using the
-     * InitialCondition::ThreeBody class as System's InitialCondition.
-     *
-     * You should start by considering the two-body case in
-     * Examples::twoBodyProblem, before continuing with this more complicated
-     * case.
-     */
+    double G = 4*M_PI*M_PI;
+
+    System* threeBodySystem = new System();
+    threeBodySystem->setIntegrator        (new VelocityVerlet(threeBodySystem));
+    threeBodySystem->setPotential         (new NewtonianGravity(G));
+    threeBodySystem->setInitialCondition  (new ThreeBody());
+    threeBodySystem->setFileWriting       (true);
+    threeBodySystem->removeLinearMomentum ();
+    threeBodySystem->integrate            (30000);
+}
+
+void Examples::multiBodyProblem() {
+    double G = 4*M_PI*M_PI; //6.67e-11;
+
+    System* multiBodySystem = new System();
+    multiBodySystem->setIntegrator        (new VelocityVerlet(multiBodySystem));
+    multiBodySystem->setPotential         (new NewtonianGravity(G));
+    multiBodySystem->setInitialCondition  (new MultiBody());
+    multiBodySystem->setFileWriting       (true);
+    multiBodySystem->removeLinearMomentum ();
+    multiBodySystem->integrate            (100000);
 }

### 2016-10-20 Project3/src/examples.h

@@ -4,4 +4,5 @@ class Examples {
 public:
     static void twoBodyProblem();
     static void threeBodyProblem();
+    static void multiBodyProblem();
 };

### 2016-10-20 Project3/src/main.cpp

@@ -2,6 +2,6 @@
 #include "examples.h"
 
 int main(int, char**) {
-    Examples::twoBodyProblem();
+    Examples::multiBodyProblem();
     return 0;
 }

### 2016-10-21 Project3/article/article.tex
Theory

@@
 
+\date{\today}
+
 \begin{document}
@@
 
-\section{Algorithm}
+\subsection{Newtonian gravtiation}
+When one studies the solar system and the way it moves, one inevitably needs to consider the gravity of the situation. The classical law of gravitation is given by Newton's law
+\begin{equation}
+\label{eq:newtoniangravity}
+F_G = \frac{GM_1M_2}{r^2},
+\end{equation} 
+where $F_G$ is the gravitational force between to bodies, $G$\footnote{$G=6.67408 \times 10^{-11} m^3 kg^{-1} s{-2}$} is the gravitational constant, $M_1$ and $M_2$ are the masses of the two bodies and $r$ is the distance between them.
+
+In three dimensions, by employing Newton's second law of motion, one will get the following componental equations for the acceleration due to gravitational pull on a particular body $i$.
+\begin{equation}
+\label{eq:componentalnewton}
+\frac{d^2x}{dt^2} = \frac{F_{G,x}}{M_i}, \quad
+\frac{d^2y}{dt^2} = \frac{F_{G,y}}{M_i}, \quad
+\frac{d^2z}{dt^2} = \frac{F_{G,z}}{M_i},
+\end{equation}
+each of which can be integrated in order to find the position of the bodies at any given time and for a particular inital velocitiy and position.
+Moreover, a neat vectorial expression for the Newtonian gravity (equation \ref{eq:newtoniangravity}) in three dimensions is
+\begin{equation}
+\vb{F}_G = \frac{GM_1M_2}{r^3}\vb{r}
+\end{equation}
+where $\vb{r}$ gives the posistion of the body in a cartesian coordinate system.
+
+\subsection{Relativity}
+
+The theory of general relativity (GR), put forth by Albert Einstein in 1915, is the current description of gravity in modern physics.
+On curiosity of the solar system that cannot be explained is the perihelion precession of Mercury's orbit.
+For every Mercury year the perihelion of the planets orbit moves sligthly.
+The observed value of this effect is $43$ arc-seconds per century.
+Einstein showed himself that GR could explain this anomaly.
+
+Closed elliptical orbits are a special feature of the factor ($1/r^2$) in Newton's law of gravitation.
+Any change to this factor will result in a change in the orbit.
+For a small such correction, each orbit will be almost the same, but as time progresses the orientation of the elliptical orbit will itself rotate.
+In this study we will not compute a space-time manifold in order to make this correction, but rather add a general relativistic error term to the Newtonian gravitational force (equation \ref{eq:newtoniangravity}.
+\begin{equation}
+\label{eq:relativenewton}
+F_G = \frac{GM_{Sun}M_{Mercury}}{r^2}\left[1+\frac{3l^2}{r^2c^2} \right],
+\end{equation}
+where $M_{MERCURY}$ is the mass of Mercury, $r$ is the distance between Mercury and the Sun, $l=\abs{\vb{r}\times\vb{v}}$ is the orbital angular momentum per unit mass, and $c$ is the speed of light in vacuum.
+
+\section{Units}
+In order to make the system easy to work with and the calculations easier as well, a change of units is warranted.
+In this study we will therefore express the mass of any body as a fraction of solar masses, that is $M_{Sun} = 2\times10^{30}kg$.
+This means that the Earth\footnote{In SI-units, $M_{Earth}\approx6\times10^{24}kg$}, for instance will have a mass of $M_{Earth}= 3\times10^{-6}$.
+
+Additionally, the units for distance will be astronmical units $AU$.
+This is the mean distance between the earth and the sun.
+For a simple system and a particular coordinate system with the sun at origin, the initial position for the Earth could simply be $\vb{r}_ {Earth}=(1,0,0)$.
+
+The unit for time will be Earth years, $yr$. This means that velocity will be in $AU/yr$.
+
+\section{Algorithm and implementation}
+
+\subsection{Object orientation}
+The
+
+\subsection{Euler-Cromer}
+
+\subsection{Velocity Verlet}
+
+\section{Data}
 

### 2016-10-23 Project3/article/article.tex
Class diagram added to article

@@
 \usepackage[]{physics}
-\usepackage[]{listings}
+\usepackage[]{color,listings}
 \usepackage[utf8]{inputenc}
 \usepackage[toc,page]{appendix}
+
+\definecolor{mygray}{gray}{0.6}
 
@@
 	keywordstyle = \color{blue},
-	identifierstyle=\color{deepgreen},
+	identifierstyle=\color{blue},
  	stringstyle=\color{orange},
@@
 \label{eq:relativenewton}
-F_G = \frac{GM_{Sun}M_{Mercury}}{r^2}\left[1+\frac{3l^2}{r^2c^2} \right],
+F_G = \frac{GM_{\odot}M_{Mercury}}{r^2}\left[1+\frac{3l^2}{r^2c^2} \right],
 \end{equation}
@@
 In order to make the system easy to work with and the calculations easier as well, a change of units is warranted.
-In this study we will therefore express the mass of any body as a fraction of solar masses, that is $M_{Sun} = 2\times10^{30}kg$.
+In this study we will therefore express the mass of any body as a fraction of solar masses, that is $M_{\odot} = 2\times10^{30}kg$.
 This means that the Earth\footnote{In SI-units, $M_{Earth}\approx6\times10^{24}kg$}, for instance will have a mass of $M_{Earth}= 3\times10^{-6}$.
@@
 
-The unit for time will be Earth years, $yr$. This means that velocity will be in $AU/yr$.
+The unit for time will be Earth years, $yr$. This means that velocity will be in units $AU/yr$.
+
+Changing the variables will have consequences for the gravitationan constant $G$ and velocities $v$ as well.
+This can be deduced easily enough by picking a sample system consisting of the earth and the sun.
+Inserting in equation \ref{eq:newtoniangravity} gives
+\begin{equation}
+\label{eq:earthandsun}
+F_G = \frac{GM_{\odot}M_{Earth}}{r^2}
+\end{equation}
+Furthermore, assuming the orbit of the earth is perfectly circular, will give the following relation
+\begin{equation}
+\label{eq:circularearth}
+F_G = \frac{M_{Earth}v^2}{r}
+\end{equation}
+Combining equations \ref{eq:earthandsun} and \ref{eq:circularearth} yields the following
+\begin{equation}
+v^2r=gM_{\odot}=4\pi^2AU^3yr^{-2}
+\end{equation}
+Because the mass of the sun as a fraction of solar masses is $M_{\odot}=1$  and the distance between the earth and the sun is $r=1AU$ we land at
+\begin{equation}
+G=4\pi^2AU^3yr^{-2} \quad v_{Earth}=2\pi AUyr^{-1}
+\end{equation}
 
@@
 \subsection{Object orientation}
-The
+Object orientation allows for more general code to be written and for easy reuse.
+Moreover, object orientation makes sense for humans, who tends to classify objects we see and interact with in everyday life.
+In this study we make use of all the advantages of object oriented programming by implementing several classes.
+A class diagram illustration of the program is included
+
+The \lstinline|System| class contains all the information about the initial conditions of the system, which integration method should be used and also some helpful methods, like one that writes data to file.
+
+The \lstinline|InitialCondition| superclass contains a method for setting up particles.
+Depending on what system one wants to model, we have implemented three subclasses of \lstinline|IntialCondition|: \lstinline|TwoBody|, \lstinline|ThreeBody| and \lstinline|MultiBody|.
+The name of these classes speek mostly for themselves.
+
+The intersting objects, contained within the \lstinline|InitialCondition| subclasses are instances of the \lstinline|Particle| class.
+The class could have another name, like "planet" or "body" because instances of it will represent the largest bodies in the solar system; the sun and (eventually) all nine planets\footnote{Including Pluto for historical reasons.}.
 
@@
 
+\pagebreak
+
+\begin{appendix}
+
+\section{Class Diagram}
+\label{app:classdiagram}
+
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=0.9\textwidth]{../figures/classdiagram.png}
+\end{figure}
+
+\end{appendix}
+
 \end{document}

### 2016-10-23 Project3/article/article.tex
Algorithms

@@
-\documentclass[10pt, a4paper]{amsart}
+\documentclass[10pt,a4paper]{amsart}
 
@@
 \usepackage[]{physics}
-\usepackage[]{color,listings}
+\usepackage[]{listings}
 \usepackage[utf8]{inputenc}
 \usepackage[toc,page]{appendix}
+\usepackage[dvipsnames]{xcolor}
 
-\definecolor{mygray}{gray}{0.6}
+\definecolor{mygray}{gray}{0.9}
 
@@
 	otherkeywords = {self},
-	keywordstyle = \color{blue},
-	identifierstyle=\color{blue},
+	keywordstyle = \color{Maroon},
+	identifierstyle=\color{olive},
  	stringstyle=\color{orange},
- 	backgroundcolor=\color{mygray}
+ 	backgroundcolor=\color{mygray},
+ 	breaklines = true
 }
 
-\title[Simulation of Solar System]{Simulation of Solar System Using Velocity Verlet Method \\
+\title[Simulation of the Solar System]{The Solar System: an Exercise in Numerical Integration \\
   \hrulefill\small{ FYS3150: Computational Physics }\hrulefill}
@@
 
-\section{Algorithm and implementation}
+\section{Algorithms and implementation}
 
@@
 In this study we make use of all the advantages of object oriented programming by implementing several classes.
-A class diagram illustration of the program is included
+A class diagram illustration of the program is included in appendix \ref{app:classdiagram}.
+Bear in mind that this diagram does not represent the functionality of the program to a full extent, as many methods and a few classes are left out.
+Rather, it helps one to understand how the class hierarchy is built up and how the program as whole is supposed to function.
 
@@
 The class could have another name, like "planet" or "body" because instances of it will represent the largest bodies in the solar system; the sun and (eventually) all nine planets\footnote{Including Pluto for historical reasons.}.
+Every particle instance takes a position vector, a velcity vector and a mass as inputs and has only these attributes\footnote{The position and velocity vectors are defined by its own class \lstinline|Vec3|.}.
+The extra abstraction to "particle" if therefore fitting, and the class could be employed elsewhere, for instance in a molecular dynamics simulation.
+
+The \lstinline|Integrator| superclass har two subclasses \lstinline|EulerCromer| and \lstinline|VelocityVerlet| each of which must owerwrite the \lstinline|integrateOneStep| method.
+These are the most important part of the program, because it is where the essential part of the two algorithms are implemented.
+The \lstinline|integrateOneStep| method is called iteratively for a specified number of steps from a class instance of the \lstinline|system| class, hence the name of the method.
+Within the \lstinline|integrateOneStep| method, the integration is performed for every particle in the system.
+A further description of the integrator algorithms follows.
 
@@
 
+Analytically the Euler-Cromer method, or the semi-implicit Euler method, can be expressed in one dimension by the following recursive relations
+\begin{align}
+v_{n+1} &= v_n + a_n dt \\
+x_{n+1} &= x_n + v_{n+1} dt,
+\end{align}
+where $a_n$ is the acceleration, $v_n$ is the velocity and $x_n$ is the position, after a certain number of steps $n$.
+$dt$ is the time step and $t_n = t_0 + ndt$ is the time after $n$ steps.
+One sees that the algorithm is relatively cheap, with $4n$ FLOPS.
+In C++ code the \lstinline|integrateOneStep| method of the \lstinline|EulerCromer| class looks like this
+\begin{lstlisting}
+void EulerCromer::integrateOneStep(std::vector<Particle*> particles) {
+
+	m_system->computeForces();
+
+	for (int i=0; i<particles.size(); i++) {
+		Particle *p = particles.at(i);
+
+		// Acceleration vector
+		vec3 a = (p->getForce()) / (p->getMass());
+
+		// Velocity update
+		p->getVelocity() += m_dt*a;
+
+		// Position update
+		p->getPosition() += m_dt*p->getVelocity();
+	}
+}
+\end{lstlisting}
+Notice that the forces used to compute the acceleration are computed within the \lstinline|System| class and that the method uses three-dimensional vectors for more realistic results.
+
+The Euler Cromer method is a first-order integrator which means that it commits a global error of the order of $dt$. This means that by decreasing the time step one will get a more accurate result.
+
 \subsection{Velocity Verlet}
+The Verlet method is an integration method designed to integrate Newton's equation of motion.
+The method provides good numerical stability and more precise results compared with a regular method like the Euler-Cromer method.
+The reason for this is that it is symplectic, which means that it conserves state-space volume\footnote{One would probably read an entire book to understand this, so we will leave it at that.}.
+
+The velocity Verlet method can also be represented by a (one-dimensional) recursive relation
+\begin{align}
+x_{n+1} &= x_n + v_ndt + \frac{1}{2}a_ndt^2 \\
+v_{n+1} &= v_n + \frac{a_n + a_{n+1}}{2}dt.
+\end{align}
+These equations tell us that the algorithm is slightly more expensive compared to the Euler-Cromer scheme at $11n$. Here follows an implementation of the algorithm in C++.
+
+\begin{lstlisting}
+void VelocityVerlet::integrateOneStep(std::vector<Particle*> particles) {
+
+	m_system->computeForces();
+
+	for (int i=0; i<particles.size(); i++) {
+		Particle *p = particles.at(i);
+
+		// Acceleration vector
+		vec3 a = (p->getForce()) / (p->getMass());
+
+		// Position update
+		p->getPosition() += p->getVelocity()*m_dt + 0.5*a*m_dt*m_dt;
+
+		m_system->computeForces();
+
+		// New accelration vector
+		vec3 anew = (p->getForce()) / (p->getMass());
+
+		// Velocity update
+		p->getVelocity() += (a + anew)*m_dt/2;
+	}
+}
+\end{lstlisting}
+One sees that it is necessary to compute forces and update acceleration once more in order to update the velocity, which means more operations must be conducted.
+Time will show if the extra cost of this algorithm is worth the  increase in precision it promises.
 
 \section{Data}
+As we wish to determine the orbits of our solar system as accurately as possible, we use date from HORIZON Web-Interface, provided by the Jet Propulsion Laboratory (NASA) at the California institute of technology: \url{http://ssd.jpl.nasa.gov/horizons.cgi}.
+Here of data from several celestial bodies can be downloaded.
+It is simple to find both position and velocity for all particles in our bodies at high precision.
+Units of length can be set to $AU$ and velocities to $AU$ per day.
+The velocities must be converted for our simulation.
+We have faith that these data are accurate.
 
 \section{Results}
+
+\subsection{Two-Body system: Earth and Sun}
+
+\subsection{Three-Body system: Earth, Sun and Jupiter}
+
+\subsection{Multi-Body system: all planets}
+
+\subsection{The perihelion precession of Mercury}
 

### 2016-10-23 Project3/article/article.tex
Added figures

@@
 
+The object-oriented model for the solar system that has been built is very flexible, as the reader will soon come to know.
+We start by constructing some simple models with two and three bodies, then the entire solar system, and lastly, simulate the perihelion precession of Mercury, as discussed in the theory section.
+
 \subsection{Two-Body system: Earth and Sun}

### 2016-10-23 Project3/scripts/periangle.m
Added MATLAB scripts

@@ -1,0 +1,22 @@
+%This script animates the movements of the planets generated by the c++
+%script.
+
+load mercury.dat;         %  read data 
+positions = mercury;
+%Specifying plot graphics
+figure;
+hold on;
+view(45,45);
+max_val = max(max(positions(:,:)));
+axis([-max_val max_val -max_val max_val -max_val max_val]);
+xlabel('x-position [AU]','interpreter','latex')
+ylabel('y-position [AU]','interpreter','latex')
+zlabel('z-position [AU]','interpreter','latex')
+title('Orbit of mercury around the sun (with correction)','interpreter','latex')
+grid on
+
+plot3(positions(:,1), positions(:,2), positions(:,3),'.y','markersize',30);
+plot3(positions(:,4), positions(:,5), positions(:,6), '-r');
+plot3(positions(end,4), positions(end,5), positions(end,6), '.b','markersize',10);
+
+axis equal

### 2016-10-23 Project3/scripts/plotThetas.m

@@ -1,0 +1,7 @@
+%This script plots the angles generated by the c++ script
+load thetas.dat;         %  read data 
+
+plot(thetas);  
+xlabel('Orbits','interpreter','latex');
+ylabel('Perihelion angle','interpreter','latex');
+title('Perihelion angle (dt = 1e-6, N = 3e7, 30 years)','interpreter','latex')

### 2016-10-23 Project3/scripts/plotearthsun.m

@@ -1,0 +1,36 @@
+%This script animates the movements of the planets generated by the c++
+%script.
+
+load earthescape.dat;         %  read data 
+positions = earthescape;
+%Specifying plot graphics
+figure
+hold on;
+view(45,45);
+max_val = max(max(positions(:,:)));
+axis([-max_val max_val -max_val max_val -max_val max_val]);
+xlabel('x-position [AU]','interpreter','latex')
+ylabel('y-position [AU]','interpreter','latex')
+zlabel('z-position [AU]','interpreter','latex')
+title('Earth $$v_{escape} = \sqrt 8 \pi$$ - dt = 1e-5, N = 1e5','interpreter','latex')
+grid on
+
+%
+% Energies for earthsunC.dat
+% Verlet
+% Step: 1000000    E =    -2e-05   Ek =1.9739e-05    Ep =-3.9478e-05
+% Euler
+% Step: 1000000    E =    -2e-05   Ek =1.9739e-05    Ep =-3.9478e-05
+
+plot3(positions(:,1), positions(:,2), positions(:,3),'.y','markersize',30);
+plot3(positions(:,4), positions(:,5), positions(:,6), '-r');
+plot3(positions(end,4), positions(end,5), positions(end,6), '.b','markersize',10);
+
+%Approximate circular earth path
+r=1;
+teta=-pi:0.01:pi;
+x=r*cos(teta);
+y=r*sin(teta);
+plot3(x,y,zeros(1,numel(x)), ':')
+
+% axis equal

### 2016-10-23 Project3/scripts/plotting.m

@@ -1,0 +1,35 @@
+
+%This script animates the movements of the planets generated by the c++
+%script.
+
+load multibody.dat;         %  read data 
+positions = multibody;
+%Specifying plot graphics
+% view(45,45);
+max_val = max(max(positions(:,:)));
+axis([-max_val/3 max_val/3 -max_val/3 max_val/3 -max_val/3 max_val/3]);
+xlabel('x-position [AU]','interpreter','latex')
+ylabel('y-position [AU]','interpreter','latex')
+zlabel('z-position [AU]','interpreter','latex')
+title('Inner Solar System (dt = 1e-3, N = 1e5)','interpreter','latex')
+grid on
+
+%
+% Energies for Threebody system.dat
+% Step: 1000000    E =    -0.004   Ek = 0.0040197    Ep =-0.0077056
+% For 10m_j
+% Step: 1000000    E =     -0.04   Ek =  0.041006    Ep = -0.077361
+% For 1000m_j
+% Step: 100000    E =        -4   Ek =     3.314    Ep =    -6.944
+
+hold on;
+%Plotting orbit of planets (Disregarding moon)
+for i=1:3:15
+    plot3(positions(:,i), positions(:,i+1), positions(:,i+2));
+    end
+    
+for i=1:3:15
+    
+    plot3(positions(end,i), positions(end,i+1), positions(end,i+2),'.','markersize',10);
+ end
+axis equal

### 2016-10-23 Project3/data/multiBodyPlot.m
Removed old files

@@ -1,20 +1,0 @@
-load positions.dat;         %  read data 
-figure
-hold on;
-
-view(45,45);
-max_val = max(max(positions(:,1:3:end)));
-axis([-max_val max_val -max_val max_val -max_val max_val]);
-for i=4:3:28
-    plot3(positions(:,i), positions(:,i+1), positions(:,i+2));
-end
-
-h = plot3(positions(1,:), positions(1,:),positions(1,:), '.k');
-
-
-for i=1:10:length(positions(:,1)) 
-    set(h, 'XData', positions(i,1:3:end), 'YData', positions(i,2:3:end),'ZData', positions(i,3:3:end));
-    drawnow;
-end
-
-axis equal

### 2016-10-23 Project3/data/twosystem.m

@@ -1,9 +1,0 @@
-load twosystem.dat;         %  read data 
-
-x_sun = twosystem(:,1);     
-y_sun = twosystem(:,2);   
-
-x_earth = twosystem(:,4);    
-y_earth = twosystem(:,5);    
-
-plot(x_earth,y_earth,x_sun,y_sun);

### 2016-10-23 Project3/src/InitialConditions/multibody.cpp
Minor code changes

@@ -6,9 +6,7 @@
 void MultiBody::setupParticles(System &system) {
 
     // Units in AU and mass as fraction of sun
-
     double solar_mass 	= 1.988544 * pow(10, 30);
-
     double mercury_mass = 3.302 * pow(10, 23) / solar_mass;
     double venus_mass 	= 48.685 * pow(10, 23) / solar_mass;
     double earth_mass 	= 5.97219 * pow(10, 24) / solar_mass;

### 2016-10-23 Project3/src/InitialConditions/threebody.cpp

@@ -15,9 +15,13 @@ void ThreeBody::setupParticles(System &system) {
      * implementation of the two-body case.
      */
 
-    Particle *sun     = new Particle(vec3(0,0,0), vec3(0,0,0), 1.0);
-    Particle *earth   = new Particle(vec3(1,0,0), vec3(0,2*M_PI,0), pow(10,-6));
-    Particle *jupiter = new Particle(vec3(5.2,0,0), vec3(0,M_PI,0), pow(10, -3));
+    double solar_mass   = 1.988544 * pow(10,30);
+    double earth_mass 	= 5.97219  * pow(10, 24) / solar_mass;
+    double jupiter_mass = 1898.13  * pow(10, 24) / solar_mass;
+
+    Particle* sun = new Particle(vec3(0,0,0), vec3(0,0,0), 1.0);
+    Particle* earth = new Particle(vec3(1,0,0), vec3(0,2*M_PI,0), earth_mass);
+    Particle* jupiter = new Particle(vec3(5.2,0,0), vec3(5.9024E-04, -7.1587E-03, 1.6542E-05)*265.2422, jupiter_mass);
 
     system.addParticle(sun);
     system.addParticle(earth);

### 2016-10-23 Project3/src/Integrators/velocityverlet.cpp

@@ -23,13 +23,18 @@ void VelocityVerlet::integrateOneStep(std::vector<Particle*> particles) {
     for (int i = 1; i < particles.size(); i++) {
         Particle *p = particles.at(i);
 
+        // Acceleration vector
         vec3 a = (p->getForce()) / (p->getMass());
 
+        // Position update
         p->getPosition() += p->getVelocity()*m_dt + (1.0/2)*a*m_dt*m_dt;
 
         m_system->computeForces();
+
+        // New acceleration vector
         vec3 anew = (p->getForce()) / (p->getMass());
 
+        // Velocity update
         p->getVelocity() += (a + anew) * (1.0/2) * m_dt;
     }
 }

### 2016-10-23 Project3/src/system.cpp

@@ -27,6 +27,7 @@ void System::computeForces() {
      */
 
     resetAllForces();
+    m_potential->resetPotentialEnergy();
 
     int n = m_numberOfParticles;
 
@@ -36,7 +37,6 @@ void System::computeForces() {
         }
     }
 
-    m_potential->resetPotentialEnergy();
 }
 
 void System::resetAllForces() {

### 2016-10-23 Project3/src/InitialConditions/multibody.cpp
Code cleanup

@@ -16,17 +16,20 @@ void MultiBody::setupParticles(System &system) {
     double uranus_mass  = 86.8103 * pow(10, 24) / solar_mass;
     double neptune_mass = 102.41 * pow(10, 24) / solar_mass;
     double pluto_mass	= 1.307 * pow(10, 22) / solar_mass;
+    double moon_mass 	= 734.9 * pow(10,20) / solar_mass;
 
-    Particle *sun = new Particle(vec3(3.5554E-03,  3.4437E-03, -1.5959E-04), vec3(-2.0393E-06,  6.8258E-06,  4.1904E-08)*265.2422, 1);
-    Particle *mercury = new Particle(vec3(-3.9067E-01, -5.3842E-02, 3.1326E-02), vec3(-1.8057E-03, -2.6625E-02, -2.0106E-03)*265.2422, mercury_mass);
-    Particle *venus = new Particle(vec3(2.9535E-01, -6.6297E-01, -2.6135E-02), vec3(1.8390E-02,  8.0510E-03, -9.5100E-04)*265.2422, venus_mass);
-    Particle *earth = new Particle(vec3(8.9232E-01,  4.5238E-01, -1.808E-04), vec3(-8.0392E-03,  1.5299E-02, -4.9539E-07)*265.2422, earth_mass);
-    Particle *mars = new Particle(vec3( 1.2012, -6.8454E-01, -4.3970E-02), vec3(7.5008E-03,  1.3339E-02,  9.5301E-05)*265.2422, mars_mass);
-    Particle *jupiter = new Particle(vec3(-5.4257, -4.8938E-01,  1.2337E-01), vec3(5.9024E-04, -7.1587E-03,  1.6542E-05)*265.2422, jupiter_mass);
-    Particle *saturn = new Particle(vec3(-2.2413,-9.7811703,  2.5927E-01), vec3(5.1320E-03, -1.2636E-03, -1.8228E-04)*265.2422, saturn_mass);
-    Particle *uranus = new Particle(vec3(1.8455E+01,  7.5787E+00, -2.1094E-01), vec3(-1.5228E-03,  3.4549E-03,  3.2562E-05)*265.2422, uranus_mass);
-    Particle *neptune = new Particle(vec3(2.82E+01, -9.9073, -4.4739E-01), vec3(1.0174E-03,  2.9810E-03, -8.4837E-05)*265.2422, neptune_mass);
[…]

### 2016-10-23 Project3/src/InitialConditions/threebody.cpp

@@ -19,9 +19,9 @@ void ThreeBody::setupParticles(System &system) {
     double earth_mass 	= 5.97219  * pow(10, 24) / solar_mass;
     double jupiter_mass = 1898.13  * pow(10, 24) / solar_mass;
 
-    Particle* sun = new Particle(vec3(0,0,0), vec3(0,0,0), 1.0);
-    Particle* earth = new Particle(vec3(1,0,0), vec3(0,2*M_PI,0), earth_mass);
-    Particle* jupiter = new Particle(vec3(5.2,0,0), vec3(5.9024E-04, -7.1587E-03, 1.6542E-05)*265.2422, jupiter_mass);
+    Particle* sun = new Particle(vec3(0,0,0), vec3(0,0,0), 1.0, "Sun");
+    Particle* earth = new Particle(vec3(1,0,0), vec3(0,2*M_PI,0), earth_mass, "Earth");
+    Particle* jupiter = new Particle(vec3(5.2,0,0), vec3(5.9024E-04, -7.1587E-03, 1.6542E-05)*365.2422, jupiter_mass, "Jupiter");
 
     system.addParticle(sun);
     system.addParticle(earth);

### 2016-10-23 Project3/src/InitialConditions/twobody.cpp

@@ -24,10 +24,13 @@ void TwoBody::setupParticles(System &system) {
      *                               Position     Velocity      Mass
      */
 
-    Particle* largeBody = new Particle(vec3(0,0,0), vec3(0,0,0), 1.0);
-    Particle* smallBody = new Particle(vec3(1,0,0), vec3(0,2*M_PI,0), pow(10,-6));
-    system.addParticle(largeBody);
-    system.addParticle(smallBody);
+    double solar_mass 	= 1.988544 * pow(10,30);
+    double mercury_mass = 3.302 * pow(10,23) / solar_mass;
+
+    Particle* sun = new Particle(vec3(3.5554E-03,  3.4437E-03, -1.5959E-04), vec3(-2.0393E-06,  6.8258E-06,  4.1904E-08)*365.2422, 1,"Sun");
+    Particle* mercury = new Particle(vec3(-3.9067E-01, -5.3842E-02,  3.1326E-02), vec3(-1.8057E-03, -2.6625E-02, -2.0106E-03)*365.2422, mercury_mass,"Mercury");
+    system.addParticle(sun);
+    system.addParticle(mercury);
 }
 
 std::string TwoBody::getName() {

### 2016-10-23 Project3/src/Integrators/integrator.h

@@ -5,7 +5,7 @@
 
 class Integrator {
 protected:
-    double          m_dt        = 1e-3;
+    double          m_dt        = 1e-5;
     class System*   m_system    = nullptr;
 
 public:

### 2016-10-23 Project3/src/Potentials/newtoniangravity.cpp

@@ -1,5 +1,7 @@
 #include "newtoniangravity.h"
 #include <iostream>
+#include <string>
+#include <iomanip>
 
 NewtonianGravity::NewtonianGravity(double G) : m_G(G) {
 
@@ -40,8 +42,18 @@ void NewtonianGravity::computeForces(Particle& a, Particle& b) {
     // Computing force components for particle a
     vec3 F;
 
-    // x-direction
+    // Force
     F = -((m_G*m_a*m_b) / (r.lengthSquared()*r.length())) * r;
+
+    // Mercury relativistic correction
+    if ( (a.getName().compare("Sun") == 0) && (b.getName().compare("Mercury") == 0 )) {
+        double c = 63239.7263; // AU per year
+        vec3 v = b.getVelocity();
+        vec3 l = r.cross(v);
+        double l2 = l.lengthSquared();
+        double correction = 1 + (3.0*l2) / (c*c*r.lengthSquared());
+        F *= correction;
+     }
 
     double V = -(m_G*m_a*m_b) / r.length();
     m_potentialEnergy += V;

### 2016-10-23 Project3/src/Potentials/newtoniangravity.h

@@ -2,10 +2,22 @@
 #include "particle.h"
 #include "potential.h"
 #include <string>
+#include <cmath>
 
 class NewtonianGravity : public Potential {
 private:
     double m_G;
+    double solar_mass = 1;
+    double mercury_mass = 3.302 * pow(10, 23) / 1.988544*pow(10,30);
+
+    // Set some helper variables before we start the time integration.
+    double thetaPrevious 		= 0;	// The perihelion angle of the previous time step.
+    double theta            	= 0;	// The perihelion angle of the current time step.
+
+    double rPreviousPrevious 	= 0;	// Mercury-Sun-distance two times steps ago.
+    double rPrevious         	= 0;	// Mercury-Sun-distance of the previous time step.
+    double r                 	= 0;	// Mercury-Sun-distance of the current time step.
+
 
 public:
     NewtonianGravity(double G);

### 2016-10-23 Project3/src/examples.cpp

@@ -22,7 +22,7 @@ void Examples::twoBodyProblem() {
     twoBodySystem->setInitialCondition  (new TwoBody());
     twoBodySystem->setFileWriting       (true);
     twoBodySystem->removeLinearMomentum ();
-    twoBodySystem->integrate            (10000);
+    twoBodySystem->integrate            (1e6);
 }
 
 void Examples::threeBodyProblem() {
@@ -34,11 +34,11 @@ void Examples::threeBodyProblem() {
     threeBodySystem->setInitialCondition  (new ThreeBody());
     threeBodySystem->setFileWriting       (true);
     threeBodySystem->removeLinearMomentum ();
-    threeBodySystem->integrate            (30000);
+    threeBodySystem->integrate            (50000);
 }
 
 void Examples::multiBodyProblem() {
-    double G = 4*M_PI*M_PI; //6.67e-11;
+    double G = 4*M_PI*M_PI;
 
     System* multiBodySystem = new System();
     multiBodySystem->setIntegrator        (new VelocityVerlet(multiBodySystem));
@@ -46,5 +46,5 @@ void Examples::multiBodyProblem() {
     multiBodySystem->setInitialCondition  (new MultiBody());
     multiBodySystem->setFileWriting       (true);
     multiBodySystem->removeLinearMomentum ();
-    multiBodySystem->integrate            (100000);
+    multiBodySystem->integrate            (50000);
 }

### 2016-10-23 Project3/src/particle.h

@@ -7,18 +7,20 @@ private:
     vec3    m_velocity  = vec3(0,0,0);
     vec3    m_forces    = vec3(0,0,0);
     double  m_mass      = 0;
+    std::string m_name;
 
 public:
     Particle();
-    Particle(class vec3 position, class vec3 velocity, double mass);
+    Particle(class vec3 position, class vec3 velocity, double mass, std::string particle_name);
     void resetForces();
     void addForce(double dFx, double dFy, double dFz);
     double velocitySquared();
 
     // Getters.
-    vec3& getPosition() { return m_position; }
-    vec3& getVelocity() { return m_velocity; }
-    vec3& getForce()    { return m_forces; }
-    double getMass()    { return m_mass; }
+    vec3& getPosition()   { return m_position; }
+    vec3& getVelocity()   { return m_velocity; }
+    vec3& getForce()      { return m_forces; }
+    double getMass()      { return m_mass; }
+    std::string getName() { return m_name; }
 };
 

### 2016-10-23 Project3/src/system.cpp

@@ -3,10 +3,13 @@
 #include "Potentials/potential.h"
 #include "InitialConditions/initialcondition.h"
 #include "particle.h"
-
 #include <iostream>
+#include <cmath>
+#include <iomanip>
+
 using std::cout;
 using std::endl;
+using std::setprecision;
 
 
 
@@ -65,10 +68,46 @@ void System::setDt(double dt) {
 void System::integrate(int numberOfSteps) {
     m_integrateSteps = numberOfSteps;
     printIntegrateInfo(0);
+
+    // Set some helper variables before we start the time integration.
+    double thetaPrevious        = 0;	// The perihelion angle of the previous time step.
+    double thetaCurrent 		= 0;	// The perihelion angle of the current time step.
+
+    double rPreviousPrevious 	= 0;	// Mercury-Sun-distance two times steps ago.
+    double rPrevious            = 0;	// Mercury-Sun-distance of the previous time step.
+    double rCurrent             = 0;	// Mercury-Sun-distance of the current time step.
+
     for (int i=1; i<numberOfSteps+1; i++) {
         m_integrator->integrateOneStep(m_particles);
         printIntegrateInfo(i);
         writePositionsToFile();
+        Particle *sun = m_particles.at(0);
+        Particle *mercury = m_particles.at(1);
+        // Compute the current perihelion angle, using the inverse tangent function from cmath.
+        // This assumes there is a vector of planets, called m_bodies, available, in which the
+        // Sun is m_bodies[0] and Mercury is m_bodies[1].
+        double x = mercury->getPosition().x() - sun->getPosition().x();
+        double y = mercury->getPosition().y() - sun->getPosition().y();
+
+        thetaCurrent = atan2( y, x );
+
+        // Compute the current Mercury-Sun distance.
+        rCurrent = (sun->getPosition()-mercury->getPosition()).length();
+
+        // Check if the *previous* time step was a minimum for the Mercury-Sun distance. I.e. check
+        // if the previous distance is smaller than the current one *and* the previous previous one.
+        if ( rCurrent > rPrevious && rPrevious < rPreviousPrevious ) {
+
+             // If we are perihelion, print angle (in radians) to terminal.
+             writeThetasToFile(thetaPrevious);
+
+             // Here you should also probably write it to file for later plotting or something.
+        }
+
[…]

### 2016-10-23 Project3/src/system.h

@@ -7,6 +7,7 @@ class System {
 private:
     bool                        m_writeToFile       = false;
     bool                        m_outFileOpen       = false;
+    bool						m_outFileOpen2		= false;
     int                         m_integrateSteps    = 0;
     int                         m_numberOfParticles = 0;
     double                      m_kineticEnergy     = 0;
@@ -16,10 +17,12 @@ private:
     class Potential*            m_potential         = nullptr;
     class InitialCondition*     m_initialCondition  = nullptr;
     std::ofstream               m_outFile;
+    std::ofstream               m_outFile2;
     std::vector<Particle*>      m_particles;
 
 public:
     System() {}
+    void writeThetasToFile		(double theta);
     void computeForces          ();
     void resetAllForces         ();
     void setPotential           (class Potential* potential);

### 2016-10-23 Project3/article/article.tex
Minor code changes
Added some figures to article

@@
 
+\begin{figure}[ht]
+	\centering
+	\textbf{Comparison of Euler-Cromer and Velocity Verlet}
+	\includegraphics[width=0.99\textwidth]{../figures/earthcompare.png}
+	\label{fig:earthcompare}
+	\caption{Figure showing the difference in precision between the Euler-Cromer method (left) and the velocity Verlet method (right). The time step is $dt=10^{-2}$ and the number of points is $N=10^6$}
+\end{figure}
+
+\begin{figure}[ht]
+	\centering
+	\textbf{Earth Escape Velocity}
+	\includegraphics[width=0.9\textwidth]{../figures/earthescape.png}
+	\caption{Illustration of Earth escaping orbit at $\sqrt{8}\pi AU/yr$}
+\end{figure}
+
 \subsection{Three-Body system: Earth, Sun and Jupiter}
+
+\begin{figure}[ht]
+	\centering
+	\textbf{Earth, Sun and Jupiter}
+	\includegraphics[width=0.99\textwidth]{../figures/threebody.png}
+	\caption{Model outcome for a Jupiter with different masses. Top left: Normal mass. Top right: Jupiter mass $10\times$ normal. Bottom: Jupiter mass $1000\times$ normal, in 2D (left) and 3D (right)}
+\end{figure}
 

### 2016-10-24 Project3/article/article.tex
Added timer, results text

@@
 where $M_{MERCURY}$ is the mass of Mercury, $r$ is the distance between Mercury and the Sun, $l=\abs{\vb{r}\times\vb{v}}$ is the orbital angular momentum per unit mass, and $c$ is the speed of light in vacuum.
+
+The precession of Mercury's orbit is not easy to spot simply by plotting the planets orbit, but the angle of the perihelion\footnote{The point in the orbit closest to the sun.} can be calculated easily enough by the following equation
+\begin{equation}
+\label{eq:perangle}
+\theta_p = \arctan \left( \frac{y_p}{x_p}\right)
+\end{equation}
 
@@
 
-\begin{figure}[ht]
+A two-body system is a meager representation of the solar system, but is the first important stepping stone towards a more interesting and realistic model.
+The two-body system has two instances of the \lstinline|Particle| class, representing the sun and the Earth.
+The orbits are found by solutions from both the Euler-Cromer method and the velocity Verlet method.
+
+In figure \ref{fig:earthcompare} shows a plot of a solution by Euler-Cromer to the left and velocity Verlet to the right.
+The same time step $dt=10^{-2}$ and the numper of points $N=10^6$ is employed in both simulations.
+The sun is the yellow dot in the middle of the plot and the Earth is the blue dot orbiting along the red path.
+The reason the past is wide is because the simulation has run for such a long time that the Earth has travelled several orbits\footnote{One hundred points per revolution and 100000 points should give 1000 revolutions in total.}.
+The difference between the two methods is quite clear.
+The velocity Verlet method har a much narrower ``orbit band'' and is therefore more precise.
+
+\begin{figure}
 	\centering
@@
 	\includegraphics[width=0.99\textwidth]{../figures/earthcompare.png}
-	\label{fig:earthcompare}
-	\caption{Figure showing the difference in precision between the Euler-Cromer method (left) and the velocity Verlet method (right). The time step is $dt=10^{-2}$ and the number of points is $N=10^6$}
+	\caption{Figure showing the difference in precision between the Euler-Cromer method (left) and the velocity Verlet method (right).
+The time step is $dt=10^{-2}$ and the number of points is $N=10^6$\label{fig:earthcompare}}
 \end{figure}
 
-\begin{figure}[ht]
+The model is easily adjustable and one can modify it to show many interesting phenomena.
+In figure \ref{fig:earthescape} is a plot showing the Earth escaping its orbit around the sun.
+The escape velocity was found by trial and error; $v=\sqrt{8}\pi AU/yr$.
+
+\begin{figure}
 	\centering
@@
 	\includegraphics[width=0.9\textwidth]{../figures/earthescape.png}
-	\caption{Illustration of Earth escaping orbit at $\sqrt{8}\pi AU/yr$}
+	\caption{Illustration of Earth escaping orbit at $v=\sqrt{8}\pi AU/yr$\label{fig:earthescape}}
 \end{figure}
+
+\begin{table}
+	\centering
+	\caption{Time comparison of Euler-Cromer and velocity Verlet applied to a two-body system. \label{tab:eulervsverlet}}
+	\begin{tabular}{ccc} \hline
+	$N$ & Euler-Cromer [t] & Velocity Verlet [t] \\ \hline
+	$10^{3}$ & 0.023 & 0.036   \\
+	$10^{4}$ & 0.20  & 0.17   \\
+	$10^{5}$ & 1.92  & 1.83   \\
+	$10^{6}$ & 18.42 & 17.88 \\ \hline
+	\end{tabular}
+\end{table}
+
+Table \ref{tab:eulervsverlet} shows a comparison of how much time the two algorithm requires to handle a problem of increasing size.
+The two algorithms were applied to the same two-body problem.
+As one can see, it appears that the velocity Verlet method is a bit quicker, but these results are probably not statistically significant.
+More on this in the discussion below.
 
@@
 
-\begin{figure}[ht]
+A natural progression from a two-body system is a three-body system.
+We have implemented a model consisting of particles representing Jupiter, Earth, and the Sun.
+The top left subfigure of figure \ref{fig:jupiter} in two dimensions.
+The velocity Verlet method is employed here.
+
+\begin{figure}
 	\centering
@@
 	\includegraphics[width=0.99\textwidth]{../figures/threebody.png}
-	\caption{Model outcome for a Jupiter with different masses. Top left: Normal mass. Top right: Jupiter mass $10\times$ normal. Bottom: Jupiter mass $1000\times$ normal, in 2D (left) and 3D (right)}
+	\caption{Model outcome for a Jupiter with different masses.
+Top left: Normal mass.
+Top right: Jupiter mass $10\times$ normal.
+Bottom: Jupiter mass $1000\times$ normal, in 2D (left) and 3D (right)\label{fig:jupiter}}
 \end{figure}
+
+The rest of the subfigures in figure \ref{fig:jupiter} show what would happen if Jupiter had larger mass than it has.
+The top right subplot shows this scenario if Jupiter's mass was $10\times$ larger than normal.
+One can see that this would drag the Earth along into the Sun, which stays mostly put, and likely skew the entire solar system.
+
+The two bottom figures show what would happen i Jupiter would have had $1000\times$ normal mass.
+The plot speaks for itself; the entire system is pulled apart.
+This situation is shows in 2D to the left and 3D to the right.
 
@@
 
+\begin{figure}
+	\centering
+	\textbf{Multi-body model of the Solar System}
+	\includegraphics[width=0.8\textwidth]{../figures/multibody.png}
+	\caption{All planets of our solar system (including Pluto) \label{fig:multibody}}
+\end{figure}
+
+Finally for the pinnacle of this study, a model of the entire solar system!\footnote{Excluding moons and other satelites, but including Pluto.} Figure \ref{fig:multibody} shows a multi-body simulation representing the solar system, integrated using the velocity Verlet algorithm.
+One can see that the algorithm handles the problem quite well.
+
 \subsection{The perihelion precession of Mercury}
 
+Figure \ref{fig:perangle} shows the angel of the perihelion of Mercury calculated by equation \ref{eq:perangle} in the theory section.
+This plot is over a period of 30 Earth years and with a time step of $dt = 10^{-6}yrs$.
+One can see a clear trend in the angle of the perihelion as predicted.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.8\textwidth]{../figures/periangle30yrs.png}
+	\caption{The precession of Mercury's orbit\label{fig:perangle}}
+\end{figure}
+
 \section{Discussion}
+
+\subsection{Integration Algorithm}
+
+Velocity Verlet for the win!! :D
 

### 2016-10-24 Project3/src/examples.cpp

@@ -12,17 +12,25 @@
 #include <iostream>
 #include <cmath>
 
+#include <chrono>
+typedef std::chrono::high_resolution_clock Clock;
+
 
 void Examples::twoBodyProblem() {
     double G = 4*M_PI*M_PI; // 6.67E-11 Nm/kg adjusted to correct units
 
     System* twoBodySystem = new System();
-    twoBodySystem->setIntegrator        (new EulerCromer(twoBodySystem));
+    twoBodySystem->setIntegrator        (new VelocityVerlet(twoBodySystem));
     twoBodySystem->setPotential         (new NewtonianGravity(G));
     twoBodySystem->setInitialCondition  (new TwoBody());
     twoBodySystem->setFileWriting       (true);
     twoBodySystem->removeLinearMomentum ();
+    auto t1 = Clock::now();
     twoBodySystem->integrate            (1e6);
+    auto t2 = Clock::now();
+    std::cout << "Time elapsed: "
+              << std::chrono::duration_cast<std::chrono::nanoseconds>(t2 - t1).count()
+              << " nanoseconds" << std::endl;
 }
 
 void Examples::threeBodyProblem() {

### 2016-10-24 Project3/src/main.cpp

@@ -2,6 +2,6 @@
 #include "examples.h"
 
 int main(int, char**) {
-    Examples::multiBodyProblem();
+    Examples::twoBodyProblem();
     return 0;
 }

### 2016-10-24 Project3/scripts/periangle.m
Cleanup

@@ -1,5 +1,9 @@
-%This script animates the movements of the planets generated by the c++
-%script.
+%{ 
+This script finds and plots perihelion angles. this is different from
+"plottingthetas.m" because it is done directly from the mercury data
+and not in the C++ script. This works only for low resolutions, as large
+data sets would destroy MATLAB
+%}
 
 load mercury.dat;         %  read data 
 positions = mercury;

### 2016-10-24 Project3/scripts/plotEarthSunEscape.m

@@ -1,13 +1,11 @@
-% {
-This script animates the movements of the planets generated by the c++
-script. Not included in report because the change in perihelion angle
-is to small to be seen
-%}
+%This script animates the movements of the Earth-Sun system generated by the c++
+%script. In this data set we played around with the escape velocity of the
+%earth, which can be seen in the generated plot.
 
-load mercury.dat;         %  read data 
-positions = mercury;
+load earthescape.dat;         %  read data 
+positions = earthescape;
 %Specifying plot graphics
-figure;
+figure
 hold on;
 view(45,45);
 max_val = max(max(positions(:,:)));
@@ -15,11 +13,18 @@ axis([-max_val max_val -max_val max_val -max_val max_val]);
 xlabel('x-position [AU]','interpreter','latex')
 ylabel('y-position [AU]','interpreter','latex')
 zlabel('z-position [AU]','interpreter','latex')
-title('Orbit of mercury around the sun (with correction)','interpreter','latex')
+title('Earth $$v_{escape} = \sqrt 8 \pi$$ - dt = 1e-5, N = 1e5','interpreter','latex')
 grid on
 
 plot3(positions(:,1), positions(:,2), positions(:,3),'.y','markersize',30);
 plot3(positions(:,4), positions(:,5), positions(:,6), '-r');
 plot3(positions(end,4), positions(end,5), positions(end,6), '.b','markersize',10);
 
-axis equal
+%Approximate circular earth path
+r=1;
+teta=-pi:0.01:pi;
+x=r*cos(teta);
+y=r*sin(teta);
+plot3(x,y,zeros(1,numel(x)), ':')
+
+% axis equal

### 2016-10-24 Project3/scripts/plotThetas.m

@@ -1,4 +1,5 @@
-%This script plots the angles generated by the c++ script
+%This script plots the perihelion angles generated by the c++ script as a
+%function of orbits.
 load thetas.dat;         %  read data 
 
 plot(thetas);  

### 2016-10-24 Project3/scripts/plotearthsun.m

@@ -1,36 +1,24 @@
-%This script animates the movements of the planets generated by the c++
-%script.
+%{
+This script plots the two-body system generated by the C++ script. It 
+plots in 2D, but can easily be modified to plot in 3D by changing from
+"plot" to "plot3" and setting the z positions
+%}
 
-load earthescape.dat;         %  read data 
-positions = earthescape;
-%Specifying plot graphics
+load ESeulerHD.dat;         %  read data 
+positions = ESeulerHD;
+
+% Specifying plot graphics
 figure
 hold on;
-view(45,45);
-max_val = max(max(positions(:,:)));
-axis([-max_val max_val -max_val max_val -max_val max_val]);
+
 xlabel('x-position [AU]','interpreter','latex')
 ylabel('y-position [AU]','interpreter','latex')
 zlabel('z-position [AU]','interpreter','latex')
-title('Earth $$v_{escape} = \sqrt 8 \pi$$ - dt = 1e-5, N = 1e5','interpreter','latex')
+title('Earth-Sun System (Euler) - dt = 1e-5, N = 1e6','interpreter','latex')
 grid on
 
-%
-% Energies for earthsunC.dat
-% Verlet
-% Step: 1000000    E =    -2e-05   Ek =1.9739e-05    Ep =-3.9478e-05
-% Euler
-% Step: 1000000    E =    -2e-05   Ek =1.9739e-05    Ep =-3.9478e-05
-
-plot3(positions(:,1), positions(:,2), positions(:,3),'.y','markersize',30);
-plot3(positions(:,4), positions(:,5), positions(:,6), '-r');
-plot3(positions(end,4), positions(end,5), positions(end,6), '.b','markersize',10);
-
-%Approximate circular earth path
-r=1;
-teta=-pi:0.01:pi;
-x=r*cos(teta);
-y=r*sin(teta);
-plot3(x,y,zeros(1,numel(x)), ':')
-
-% axis equal
+plot(positions(:,1), positions(:,2),'.y','markersize',30);
+plot(positions(:,4), positions(:,5), '-r');
+plot(positions(end,4), positions(end,5),'.b','markersize',10);
+  
+axis equal

### 2016-10-24 Project3/scripts/plotting.m

@@ -1,6 +1,7 @@
-
-%This script animates the movements of the planets generated by the c++
-%script.
+%This script plots the multibody system generated by the c++ script. This
+%latest edition restricts the data to 1/3, plotting only the inner system.
+%This script can easily be modified to work with both the 2 and 3 body
+%system.
 
 load multibody.dat;         %  read data 
 positions = multibody;
@@ -13,15 +14,6 @@ ylabel('y-position [AU]','interpreter','latex')
 zlabel('z-position [AU]','interpreter','latex')
 title('Inner Solar System (dt = 1e-3, N = 1e5)','interpreter','latex')
 grid on
-
-%
-% Energies for Threebody system.dat
-% Step: 1000000    E =    -0.004   Ek = 0.0040197    Ep =-0.0077056
-% For 10m_j
-% Step: 1000000    E =     -0.04   Ek =  0.041006    Ep = -0.077361
-% For 1000m_j
-% Step: 100000    E =        -4   Ek =     3.314    Ep =    -6.944
-
 hold on;
 %Plotting orbit of planets (Disregarding moon)
 for i=1:3:15

### 2016-10-24 Project3/src/InitialConditions/multibody.cpp

@@ -5,6 +5,13 @@
 
 void MultiBody::setupParticles(System &system) {
 
+    /*
+     * In our project we use 10 particles.
+     * This includes all planets in the solar system,
+     * plus the sun and Pluto.
+     * The moon is declared and initialized but not plotted later
+     */
+
     // Units in AU and mass as fraction of sun
     double solar_mass 	= 1.988544 * pow(10, 30);
     double mercury_mass = 3.302 * pow(10, 23) / solar_mass;
@@ -18,7 +25,7 @@ void MultiBody::setupParticles(System &system) {
     double pluto_mass	= 1.307 * pow(10, 22) / solar_mass;
     double moon_mass 	= 734.9 * pow(10,20) / solar_mass;
 
-
+    // Planets
     Particle *sun = new Particle(vec3(3.5554E-03,  3.4437E-03, -1.5959E-04), vec3(-2.0393E-06,  6.8258E-06,  4.1904E-08)*365.2422, 1, "Sun");
     Particle *mercury = new Particle(vec3(-3.9067E-01, -5.3842E-02, 3.1326E-02), vec3(-1.8057E-03, -2.6625E-02, -2.0106E-03)*365.2422, mercury_mass, "Mercury");
     Particle *venus = new Particle(vec3(2.9535E-01, -6.6297E-01, -2.6135E-02), vec3(1.8390E-02,  8.0510E-03, -9.5100E-04)*365.2422, venus_mass, "Venus");
@@ -28,7 +35,11 @@ void MultiBody::setupParticles(System &system) {
     Particle *saturn = new Particle(vec3(-2.2413,-9.7811703,  2.5927E-01), vec3(5.1320E-03, -1.2636E-03, -1.8228E-04)*365.2422, saturn_mass, "Saturn");
     Particle *uranus = new Particle(vec3(1.8455E+01,  7.5787E+00, -2.1094E-01), vec3(-1.5228E-03,  3.4549E-03,  3.2562E-05)*365.2422, uranus_mass, "Uranus");
     Particle *neptune = new Particle(vec3(2.82E+01, -9.9073, -4.4739E-01), vec3(1.0174E-03,  2.9810E-03, -8.4837E-05)*365.2422, neptune_mass, "Neptune");
+
+    // Dwarf planets
[…]

### 2016-10-24 Project3/src/InitialConditions/threebody.cpp

@@ -6,13 +6,9 @@
 
 void ThreeBody::setupParticles(System &system) {
     /*
-     * This is where you should implement the initialization of a three-body
-     * system.
-     *
-     * You should start by impelmenting the two-body scheme in the
-     * InitialiCondition::TwoBody class, and then implement the more
-     * complicated three-body initialization only after you have a working
-     * implementation of the two-body case.
+     * Here, we set uo the particle system for three particles.
+     * For our project we have one such system:
+     * The Sun-Earth-Jupiter system
      */
 
     double solar_mass   = 1.988544 * pow(10,30);

### 2016-10-24 Project3/src/InitialConditions/twobody.cpp

@@ -6,22 +6,10 @@
 
 void TwoBody::setupParticles(System &system) {
     /*
-     * Here, two bodies are created at a distance r=1.0 apart from eachother.
-     * A heavy body is placed in the origin, while a lighter body is placed
-     * in (1,0,0) with an initial velocity of (0,1,0). The larger body is given
-     * a mass of 1.0, while the mass of the smaller body is set to 0.1.
-     *
-     * Please note that these are NOT the values needed in order to achieve a
-     * circular orbit (Sun-Earth-like). It is up to you to find the correct
-     * masses, distances, and velocities.
-     *
-     * Note that the constructor of the Particle class takes three inputs,
-     * a vec3-vector of positions values, a vec3-vector of velocity values,
-     * and the mass--a double. As an example:
-     *
-     *      Particle* p = new Particle(vec3(1,2,3), vec3(4,5,6), 7.0);
-     *                                  ^            ^            ^
-     *                               Position     Velocity      Mass
+     * Here, we set up the particle system for two particles.
+     * For our project we have used two of these:
+     * The Sun-Earth system and
+     * The Sun-Mercury system
      */
 
     double solar_mass 	= 1.988544 * pow(10,30);

### 2016-10-24 Project3/src/Integrators/eulercromer.cpp

@@ -11,24 +11,15 @@ void EulerCromer::integrateOneStep(std::vector<Particle*> particles) {
     for (int i=0; i<particles.size(); i++) {
         Particle *p = particles.at(i);
         /*
-         * This is where you need to update the positions and the velocities
-         * and the velocities of each particle according to the Euler-Cromer
-         * scheme.
-         *
-         * You can access the position of particle number i by
-         *
-         *      p->getPosition()
-         *
-         * Remember that the positions and velocities of all the particles are
-         * vec3 vectors, which you can use the +=, -=, etc. operators on.
+         * The Euler-Cromer scheme
          */
 
         // Acceleration vector
         vec3 a = (p->getForce()) / (p->getMass());
 
-       // std::cout << "Accel = "; a.print();
-       // std::cout << "Mass: " << p->getMass() << std::endl;
-       // std::cout << "Force: " << p->getForce() << std::endl;
+        // std::cout << "Accel = "; a.print();
+        // std::cout << "Mass: " << p->getMass() << std::endl;
+        // std::cout << "Force: " << p->getForce() << std::endl;
 
         // Velocity update
         p->getVelocity() += m_dt*a;

### 2016-10-24 Project3/src/Integrators/velocityverlet.cpp

@@ -11,12 +11,7 @@ std::string VelocityVerlet::getName() {
 
 void VelocityVerlet::integrateOneStep(std::vector<Particle*> particles) {
     /*
-     * This is where you should implement the Velocity Verlet algorithm.
-     *
-     * You should start by impelmenting the Euler-Cromer scheme in the
-     * Integrator::EulerCromer class, and then implement the more complicated
-     * Velocity Verlet algorithm only after you have a working implementation
-     * of Euler-Cromer.
+     * The velocity Verlet scheme
      */
 
     m_system->computeForces();
@@ -29,6 +24,7 @@ void VelocityVerlet::integrateOneStep(std::vector<Particle*> particles) {
         // Position update
         p->getPosition() += p->getVelocity()*m_dt + (1.0/2)*a*m_dt*m_dt;
 
+        // Recomputing forces to get new acceleration
         m_system->computeForces();
 
         // New acceleration vector

### 2016-10-24 Project3/src/Potentials/newtoniangravity.cpp

@@ -3,6 +3,8 @@
 #include <string>
 #include <iomanip>
 
+using std::setprecision;
+
 NewtonianGravity::NewtonianGravity(double G) : m_G(G) {
 
 }
@@ -12,22 +14,6 @@ void NewtonianGravity::computeForces(Particle& a, Particle& b) {
      * This is where the ordinary Newtoninan gravity forces and potential
      * energies should be calculated. This method is called by the System
      * class in System::computeForces, for all particle pairs a and b.
-     *
-     * Note that you may access the mass and the position of the particles a
-     * and b by
-     *
-     *      a.getMass();       b.getMass();
-     *      a.getPosition();   b.getPosition();
-     *
-     * In order to apply the forces to each particle, it is easiest to use the
-     * Particle::addForce method.
-     *
-     * Since calculating the forces between a and b almost inevitably involves
-     * calculating the potential energy, V(r) WHY? WHY?, it is assumed by the Potential
-     * class that this quantity is calculated here and added to the
-     * m_potentialEnergy variable. Note: You may skip this until you have a
-     * working two-body problem, since the calculation of the potential energy
-     * is only neccessary for verification purposes later.
      */
 
     // Aquiring data
@@ -37,26 +23,33 @@ void NewtonianGravity::computeForces(Particle& a, Particle& b) {
     double m_b = b.getMass();
     vec3 r = r_a - r_b;
 
+    old_angular_momentum = angular_momentum;
+
     // FORCE COMPUTATION
-
-    // Computing force components for particle a
     vec3 F;
-
-    // Force
     F = -((m_G*m_a*m_b) / (r.lengthSquared()*r.length())) * r;
 
-    // Mercury relativistic correction
-    if ( (a.getName().compare("Sun") == 0) && (b.getName().compare("Mercury") == 0 )) {
-        double c = 63239.7263; // AU per year
+    // Angular momentum
         vec3 v = b.getVelocity();
         vec3 l = r.cross(v);
         double l2 = l.lengthSquared();
-        double correction = 1 + (3.0*l2) / (c*c*r.lengthSquared());
-        F *= correction;
-     }
 
-    double V = -(m_G*m_a*m_b) / r.length();
-    m_potentialEnergy += V;
+        //Checking if angular momentum conserved
+        angular_momentum = l2;
+        double angdiff = fabs(old_angular_momentum-angular_momentum);
+        if ( angdiff > 1e-5){
[…]

### 2016-10-24 Project3/src/Potentials/newtoniangravity.h

@@ -9,6 +9,8 @@ private:
     double m_G;
     double solar_mass = 1;
     double mercury_mass = 3.302 * pow(10, 23) / 1.988544*pow(10,30);
+    double old_angular_momentum = 0;
+    double angular_momentum = 0;
 
     // Set some helper variables before we start the time integration.
     double thetaPrevious 		= 0;	// The perihelion angle of the previous time step.

### 2016-10-24 Project3/src/Potentials/nopotential.cpp

@@ -10,6 +10,6 @@ void NoPotential::computeForces(Particle&, Particle&) {
     /*
      * No potential means just no forces acting between the particles, so
      * we dont have to do anything here. This potential can be used to test
-     * certain aspects of your code, without the complication of forces.
+     * certain aspects of our code, without the complication of forces.
      */
 }

### 2016-10-24 Project3/src/system.cpp

@@ -123,13 +123,6 @@ double System::computeKineticEnergy() {
      * this is independent of the potential in use, we place this method
      * directly in the system class.
      *
-     * Remember that you can access the mass and velocity of particle i by
-     *
-     *      m_particles.at(i)->getMass()
-     *      m_particles.at(i)->getVelocity()
-     *
-     * Remember also that the Particle class has a built in method
-     * Particle::velocitySquared which can be used here.
      */
 
     m_kineticEnergy = 0;
@@ -168,15 +161,7 @@ void System::removeLinearMomentum() {
      * ensure the entire system does not drift away during long integration
      * times.
      *
-     * Remember that you can access the mass and velocity of particle i by
-     *
-     *      m_particles.at(i)->getMass();
-     *      m_particles.at(i)->getVelocity();
-     *
-     * Remember also that the vec3-vector class supports the += and -=
-     * operators, so you can do
-     *
-     *      totalMomentum += p->getVelocity() * p->getMass();
+     * This correction is not used as velocites from NASA are sufficient
      */
 
     vec3 totalMomentum = vec3(0,0,0);
@@ -193,17 +178,9 @@ void System::setFileWriting(bool writeToFile) {
 
 void System::writePositionsToFile() {
     if (m_outFileOpen == false) {
-        m_outFile.open("../positions.dat", std::ios::out);
+        m_outFile.open("../ESverletHD.dat", std::ios::out);
         m_outFileOpen = true;
-    }
-    /*
-     * This is where you should print the positions of each particle to file.
-     * Note that the file, "positions.dat", is already open; it is opened in
-     * the above if-test the first time this method is called in
-     * System::Integrate.
-     *
-     * Which format you choose for the data file is up to you.
-     */
+    } 
 
     for (int i = 0; i < m_numberOfParticles; i++) {
         m_outFile << setprecision(15) << m_particles.at(i)->getPosition();

### 2016-10-24 Project3/article/article.tex
nearly there..

@@
 \usepackage[dvipsnames]{xcolor}
+\usepackage{dirtytalk}
 
@@
 
-\title[Simulation of the Solar System]{The Solar System: an Exercise in Numerical Integration \\
+\title[Simulation of the Solar System]{The Solar System: an Exercise in Numerical Integration and \\
   \hrulefill\small{ FYS3150: Computational Physics }\hrulefill}
@@
 \section{Introduction}
+
+This study has three main goals. To show some of the advantages of object-oriented programming, to implement and compare different integration methods and to build a working model of the solar system.
 
@@
 \end{align}
-These equations tell us that the algorithm is slightly more expensive compared to the Euler-Cromer scheme at $11n$. Here follows an implementation of the algorithm in C++.
+These equations tell us that the algorithm is slightly more expensive compared to the Euler-Cromer scheme at $11n$ FLOPS. Here follows an implementation of the algorithm in C++.
 
@@
 We have faith that these data are accurate.
+
+In an arbitrary system, the entire system will usually gain momentum and drift off in some direction over time.
+One must usually correct for such drift be substracting total linear momenum.
+Because the velocity data from NASA is very good, this is unnecessary for our system.
+Total linear momemtum for the solar system is already zero from our frame of reference.
 
@@
 	\textbf{Multi-body model of the Solar System}
-	\includegraphics[width=0.8\textwidth]{../figures/multibody.png}
+	\includegraphics[width=0.9\textwidth]{../figures/multibody.png}
 	\caption{All planets of our solar system (including Pluto) \label{fig:multibody}}
@@
 One can see that the algorithm handles the problem quite well.
+We were even able to make an anmation of this system in MATLAB, which was a delightful exercise.
 
@@
 One can see a clear trend in the angle of the perihelion as predicted.
+We get a slope of the perhelion angle versus number of orbist of approximately $10^{-6}$, for a century on earth Mercury makes $415.2091$ orbits, which gives us a perhelion precession of $\approx1.5''/100yrs$.
+This much lower than the expected precession of $43''/100yrs$.
 
@@
 
+\subsection{Object Orientation}
+This project has been a nice exercise in object oriented thinking.
+We have built a larger and larger system by adding more class instances and building subclasses such as different integrators.
+Object oriented programming is close to how we as humans usually thinks about all things in the world.
+A planet is something that has mass, a position, velocity and some other factors.
+However, when one learns how to program learns object orientation at a late stage of the learning process, when one has been focusing on a more procedural programming technique.
+
+Most widely used programming languages today are object-oriented to a greater or lesser extent: Java, C++, Python, PHP, Ruby, Smalltalk and Common Lisp, to name a few.
+Critique of object orientation usually emphasizes that design and modelling comes at the expense of other important aspects like the actual computation and algorithms.
+The creator of Erlang, Joe Armstrong, is quoted as saying, \say{The problem with object-oriented languages is they've got all this implicit environment that they carry around with them.
+You wanted a banana but what you got was a gorilla holding the banana and the entire jungle}.
+
+As fairly level-headed Norwegian patriots\footnote{A friendly reminder to the reader: the creators of the first object oriented languas, Simula, were Norwegian: Ole-Johan Dahl, Kristen Nygaard.} we cannot end the discussion about object orientation there, but rather by pointing out the advantages.
+Object Oriented Programming provides a clear modular structure for programs and makes it easy to maintain and modify existing code thereby reducing lowering programming ``cost''.
+
 \subsection{Integration Algorithm}
 
-Velocity Verlet for the win!! :D
+We have seen that the velocity Verlet algorithm is much more precise than the Euler-Cromer method (figure \ref{tab:eulervsverlet}).
+In addition it appears that it may also be quicker (table \ref{tab:eulervsverlet}), which is a surprise since the Verlet algorithm seems to include more FLOPS than the Euler-Cromer method.
+This leads us to believe that the velocity Verlet may not be that much faster, at least we can conclude that the results are not statistically significant since we only have one observation per method per value for $N$.
+An improvement on this study would have been to conduct such a test.
+We propose larger sampling of code run time and, for simplicity, a paired Student's t-test as a start to answer this question\cite{student}.
+
+The discussion of an Euler method versus a Verlet method is a rather big one.
+Both methods are quite quick cheap which is what you want, and they are usually the two methods of choice if one does not need entirely exact results, but results that are ``good enough''.
+A field where this discussion prominent is in physics engine design for video games.
+With more computing power one would probably go for a more sophisticated method like a foruth degree Runge-Kutta method\cite{physicsengine}.
+
+It is worth metioning something about the symplecticity of the velocity Verlet method.
+As stated previously, a symplectic method will be good at simulating systems with energy conservation.
+We believe that the solar system has conserved energy, but our model of it may not be.
+Looking further into this matter we have found that both energy and angular momentum is conserved for the two-body Earth-Sun system.
+This would warrant further use of the velocity Verlet scheme.
+See appendix \ref{app:energy} for a sample readout of energies and angular momenta.
+
+\subsection{Mercury}
+
+We tried to compute the perihelion precessrion of Mercury.
+We found it to be $\approx1.5''/100yrs$ when it should have been much more, id est $43''/100yrs$.
+This error can stem from several sources.
+Most obvious is that we have done something wrong and have been unable to find where.
+Antoher likely source of uncertainty is that the method is that the integration method is not precise enough.
+We had problems decreasing the step size further, and extending the simulation to a full century because the job was too much for our hardware.
+One last probelm that must be mentioned is that Mercury's precession cannot be explained by the simple relativistic correction we are employing.
 
 \section{Conclusion}
+
+\begin{thebibliography}{9}
+
+\bibitem{erlang}
+	Armstrong, Joe (2009),
+	\emph{Coders at Work: Reflection on the Craft of Programming},
+	Peter Seibel, ed.
+
+\bibitem{student}
+	Hazewinkel, Michiel, ed. (2001), 
+	Student test,
+	\emph{Encyclopedia of Mathematics},
+	Springer.
+
+\bibitem{physicsengine}
+	Boesch, Florian (2010, August 28),
+	\emph{Integration by Example - Euler vs Verlet vs Runge-Kutta},
+	Retrieved October 24 2016 from \url{http://codeflow.org/entries/2010/aug/28/integration-by-example-euler-vs-verlet-vs-runge-kutta/},
+
+\end{thebibliography}
 
@@
 
+\section{Energy Read-out}
+\label{app:energy}
+\begin{lstlisting}[keywordstyle=\ttfamily, identifierstyle=\ttfamily]
+Energies printed in C++:
+
+Energies for Earth-Sun
+EULER-CROMER
+Angular momentum: 39.4784*m
+Step: 1000000  E = -2e-05   Ek =1.9739e-05  Ep =-3.9478e-05
+VERLET
+Angular momentum: 39.4784*m
+Step: 1000000  E = -2e-05   Ek =1.9739e-05  Ep =-3.9478e-05
+
+Energies for three-body system
+Step: 1000000  E = -0.004   Ek = 0.0040197  Ep =-0.0077056
+For 10m_j
+Step: 1000000  E =  -0.04   Ek =  0.041006  Ep = -0.077361
+For 1000m_j
+Step: 100000   E =     -4   Ek =     3.314  Ep =    -6.944
+
+Energies for multi-body system dt=1e-5 N=1e5
+Step: 100000   E = -0.004   Ek = 0.0040429  Ep =-0.0084804
+\end{lstlisting}
+
 \end{appendix}

### 2016-10-24 Project3/src/InitialConditions/twobody.cpp

@@ -16,7 +16,8 @@ void TwoBody::setupParticles(System &system) {
     double mercury_mass = 3.302 * pow(10,23) / solar_mass;
 
     Particle* sun = new Particle(vec3(3.5554E-03,  3.4437E-03, -1.5959E-04), vec3(-2.0393E-06,  6.8258E-06,  4.1904E-08)*365.2422, 1,"Sun");
-    Particle* mercury = new Particle(vec3(-3.9067E-01, -5.3842E-02,  3.1326E-02), vec3(-1.8057E-03, -2.6625E-02, -2.0106E-03)*365.2422, mercury_mass,"Mercury");
+    //Particle* mercury = new Particle(vec3(-3.9067E-01, -5.3842E-02,  3.1326E-02), vec3(-1.8057E-03, -2.6625E-02, -2.0106E-03)*365.2422, mercury_mass,"Mercury");
+    Particle* mercury = new Particle(vec3(0.3075, 0,0), vec3(0, 12.44, 0), mercury_mass, "Mercury");
     system.addParticle(sun);
     system.addParticle(mercury);
 }

### 2016-10-24 Project3/src/Integrators/integrator.h

@@ -5,7 +5,7 @@
 
 class Integrator {
 protected:
-    double          m_dt        = 1e-5;
+    double          m_dt        = 1e-8;
     class System*   m_system    = nullptr;
 
 public:

### 2016-10-24 Project3/src/examples.cpp

@@ -20,13 +20,13 @@ void Examples::twoBodyProblem() {
     double G = 4*M_PI*M_PI;
 
     System* twoBodySystem = new System();
-    twoBodySystem->setIntegrator        (new EulerCromer(twoBodySystem));
+    twoBodySystem->setIntegrator        (new VelocityVerlet(twoBodySystem));
     twoBodySystem->setPotential         (new NewtonianGravity(G));
     twoBodySystem->setInitialCondition  (new TwoBody());
     twoBodySystem->setFileWriting       (false);
     twoBodySystem->removeLinearMomentum ();
     auto t1 = Clock::now();
-    twoBodySystem->integrate            (1e6);
+    twoBodySystem->integrate            (1e10);
     auto t2 = Clock::now();
     std::cout << "Time elapsed: "
               << std::chrono::duration_cast<std::chrono::nanoseconds>(t2 - t1).count()

### 2016-10-24 Project3/src/system.cpp

@@ -14,20 +14,6 @@ using std::setprecision;
 
 
 void System::computeForces() {
-    /*
-     * Here you should sum over all particle pairs and compute the forces
-     * between each one. This should be done by the Potential::computeForces
-     * method which takes pointers to two Particles as input. I.e. the forces
-     * between particle i and particle j should be computed by
-
-     *      m_potential->computeForces(*m_particles.at(i), *m_particles.at(j));
-     *
-     * Note: It is important that you do not sum over each particle pair more
-     * than once. A simple way to ensure this is done is by a double for loop,
-     * one running from i=0...n, and the other running from j=i+1...n. You
-     * should convince yourself that this is true before you implement this
-     * loop.
-     */
 
     resetAllForces();
     m_potential->resetPotentialEnergy();
@@ -98,10 +84,10 @@ void System::integrate(int numberOfSteps) {
         // if the previous distance is smaller than the current one *and* the previous previous one.
         if ( rCurrent > rPrevious && rPrevious < rPreviousPrevious ) {
 
-             // If we are perihelion, print angle (in radians) to terminal.
-             writeThetasToFile(thetaPrevious);
-
-             // Here you should also probably write it to file for later plotting or something.
+            // If we are perihelion, print angle (in radians) to terminal.
+            writeThetasToFile(thetaPrevious);
+            cout << "Theta: " << thetaPrevious << endl;
+            // Here you should also probably write it to file for later plotting or something.
         }
 
         // Update the helper variables (current, previous, previousPrevious).
@@ -145,14 +131,14 @@ void System::printIntegrateInfo(int stepNumber) {
              << "  o Potential in use:    " << m_potential->getName() << endl
              << "  o Integrator in use:   " << m_integrator->getName() << endl
              << endl;
-    } else if (stepNumber % 1000 == 0) {
+    }/* else if (stepNumber % 1000 == 0) {
         m_kineticEnergy     = computeKineticEnergy();
         m_potentialEnergy   = m_potential->getPotentialEnergy();
         m_totalEnergy       = m_kineticEnergy + m_potentialEnergy;
         printf("Step: %5d    E =%10.5g   Ek =%10.5g    Ep =%10.5g\n",
                stepNumber, m_totalEnergy, m_kineticEnergy, m_potentialEnergy);
         fflush(stdout);
-    }
+    }*/
 }
 
 void System::removeLinearMomentum() {

### 2016-10-24 Project3/article/article.tex
Ready for final read-through

@@
 
-\title[Simulation of the Solar System]{The Solar System: an Exercise in Numerical Integration and \\
+\title[Simulation of the Solar System]{The Solar System: an Exercise in Numerical Integration and Object oriented design \\
   \hrulefill\small{ FYS3150: Computational Physics }\hrulefill}
@@
 \begin{abstract}
-Two, three, multi. With Euler-Cromer and then Velocity Verlet
+In this project we build a flexible and easily expandible object oriented framework which we use to simulate the solar system.
+In order to calculate orbits of celastial bodies we implement the Euler-Cromer and the velocity Verlet numerical integration methods.
+The Verlet method proves superior.
+The model is expanded gradually from a two-body system to include all planets of the solar system.
 \end{abstract}
@@
 
-This study has three main goals. To show some of the advantages of object-oriented programming, to implement and compare different integration methods and to build a working model of the solar system.
+This study has three main goals.
+To show some of the advantages of object-oriented programming, to implement and compare different integration methods and to build a working model of the solar system.
+Object oriented programming has several advantages that will be illustrated here, among them are flexibility and reusable code.
+We shall see how a system of just a couple of bodies easily expands to the full model of the solar system.
+The main task at hand when modelling the solar system is to compute the orbits of all celestial bodies.
+In order to do this one must integrate Newtons's law of motion, and an integration method is needed.
+We will implement two different methods, the Euler-Cromer method and the velocity Verlet method.
+First, we give a sound theoretic background for the study, including both classical and modern laws of gravity and a few unit tricks to make everything a bit easier.
+Second, we give a walkthrough of how the object oriented framework is built and thereafter follows an explanation of the implementation of the integration algorithms.
+Third, we give some thoughts on the data we use as basis for the initial conditions.
+Fourth, we present the results of the undertaking, in the same order as we have worked on expanding the model, with increasing complexity.
+Fifth, we provide a discussion of the results and last, some concluding remarks.
+All computer code in C++ can be found by following the Github URL beneath author names on the front page.
 
@@
 \subsection{Newtonian gravtiation}
-When one studies the solar system and the way it moves, one inevitably needs to consider the gravity of the situation. The classical law of gravitation is given by Newton's law
+When one studies the solar system and the way everything moves, one inevitably needs to consider the gravity of the situation. The classical law of gravitation is given by Newton's law
 \begin{equation}
@@
 
-In figure \ref{fig:earthcompare} shows a plot of a solution by Euler-Cromer to the left and velocity Verlet to the right.
+In figure \ref{fig:earthcompare} shows a plot of a solution by Euler-Cromer to the right and velocity Verlet to the left.
 The same time step $dt=10^{-2}$ and the numper of points $N=10^6$ is employed in both simulations.
@@
 	\includegraphics[width=0.99\textwidth]{../figures/earthcompare.png}
-	\caption{Figure showing the difference in precision between the Euler-Cromer method (left) and the velocity Verlet method (right).
+	\caption{Figure showing the difference in precision between the Euler-Cromer method (right) and the velocity Verlet method (left).
 The time step is $dt=10^{-2}$ and the number of points is $N=10^6$\label{fig:earthcompare}}
@@
 Figure \ref{fig:perangle} shows the angel of the perihelion of Mercury calculated by equation \ref{eq:perangle} in the theory section.
-This plot is over a period of 30 Earth years and with a time step of $dt = 10^{-6}yrs$.
+This plot is over a period of almost fifty Earth years and with a time step of $dt = 10^{-7}yrs$.
 One can see a clear trend in the angle of the perihelion as predicted.
-We get a slope of the perhelion angle versus number of orbist of approximately $10^{-6}$, for a century on earth Mercury makes $415.2091$ orbits, which gives us a perhelion precession of $\approx1.5''/100yrs$.
+We get a slope of the perhelion angle versus number of orbits of approximately $5.1^{-7}$, for a century on earth Mercury makes $415.2091$ orbits, which gives us a perhelion precession of $\approx0.8''/100yrs$.
 This much lower than the expected precession of $43''/100yrs$.
@@
 	\centering
-	\includegraphics[width=0.8\textwidth]{../figures/periangle30yrs.png}
+	\includegraphics[width=0.8\textwidth]{../figures/mercuryPerihelion.png}
 	\caption{The precession of Mercury's orbit\label{fig:perangle}}
@@
 This error can stem from several sources.
-Most obvious is that we have done something wrong and have been unable to find where.
+Most obvious is that we have done something wrong and have been unable to find out what.
 Antoher likely source of uncertainty is that the method is that the integration method is not precise enough.
@@
 \section{Conclusion}
+
+In this study we have shown the advantages of building an object oriented framework and how it can easily be expanded upon and modified to its needs.
+We wanted to apply object oriented design on analysis of the solar system and have succeded.
+Moreover, a few of the classes, like the \lstinline|System|, \lstinline|Particle| and integrator subclasses can be applied to other problem and in new fields, like molecular dynamics.
+
+We have implemented two integrators, the Euler-Cromer and the velocity Verlet.
+The velocity Verlet proved to be more stable than the Euler-Cromer method.
+The reason for this is because Verlet integration methods are symplectic and therefore perfect for system with conserved angular momentum and energy.
+Additionally, the velocity Verlet method seemed in theory to be more CPU costly than the Euler-Cromer method, but proved just as fast.
+
+Lastly, we tried to adjust Mercurys orbit for relativity.
+We found a clear trend in the perihelion angle of Mercury's orbit, but got dissappointing results compared with previous observations and calculations.
 
@@
 \end{thebibliography}
+
+\vfill
 

### 2016-10-24 Project3/src/Integrators/integrator.h

@@ -5,7 +5,7 @@
 
 class Integrator {
 protected:
-    double          m_dt        = 1e-8;
+    double          m_dt        = 1e-7;
     class System*   m_system    = nullptr;
 
 public:

### 2016-10-24 Project3/src/examples.cpp

@@ -26,7 +26,7 @@ void Examples::twoBodyProblem() {
     twoBodySystem->setFileWriting       (false);
     twoBodySystem->removeLinearMomentum ();
     auto t1 = Clock::now();
-    twoBodySystem->integrate            (1e10);
+    twoBodySystem->integrate            (1e9);
     auto t2 = Clock::now();
     std::cout << "Time elapsed: "
               << std::chrono::duration_cast<std::chrono::nanoseconds>(t2 - t1).count()

### 2016-10-24 Project3/src/system.cpp

@@ -86,7 +86,7 @@ void System::integrate(int numberOfSteps) {
 
             // If we are perihelion, print angle (in radians) to terminal.
             writeThetasToFile(thetaPrevious);
-            cout << "Theta: " << thetaPrevious << endl;
+
             // Here you should also probably write it to file for later plotting or something.
         }
 
@@ -131,14 +131,14 @@ void System::printIntegrateInfo(int stepNumber) {
              << "  o Potential in use:    " << m_potential->getName() << endl
              << "  o Integrator in use:   " << m_integrator->getName() << endl
              << endl;
-    }/* else if (stepNumber % 1000 == 0) {
+    } else if (stepNumber % 1000 == 0) {
         m_kineticEnergy     = computeKineticEnergy();
         m_potentialEnergy   = m_potential->getPotentialEnergy();
         m_totalEnergy       = m_kineticEnergy + m_potentialEnergy;
         printf("Step: %5d    E =%10.5g   Ek =%10.5g    Ep =%10.5g\n",
                stepNumber, m_totalEnergy, m_kineticEnergy, m_potentialEnergy);
         fflush(stdout);
-    }*/
+    }
 }
 
 void System::removeLinearMomentum() {

### 2016-10-24 Project3/article/article.tex

@@
 To show some of the advantages of object-oriented programming, to implement and compare different integration methods and to build a working model of the solar system.
-Object oriented programming has several advantages that will be illustrated here, among them are flexibility and reusable code.
+Object oriented programming has several advantages that will be illustrated here, among them are flexible and reusable code.
 We shall see how a system of just a couple of bodies easily expands to the full model of the solar system.
@@
 First, we give a sound theoretic background for the study, including both classical and modern laws of gravity and a few unit tricks to make everything a bit easier.
-Second, we give a walkthrough of how the object oriented framework is built and thereafter follows an explanation of the implementation of the integration algorithms.
+Second, we give a walkthrough of how the object oriented framework is built and thereafter an explanation of the implementation of the integration algorithms.
 Third, we give some thoughts on the data we use as basis for the initial conditions.
@@
 
-In three dimensions, by employing Newton's second law of motion, one will get the following componental equations for the acceleration due to gravitational pull on a particular body $i$.
+In three dimensions, by employing Newton's second law of motion, one will get the following componential equations for the acceleration due to gravitational pull on a particular body $i$.
 \begin{equation}

### 2016-11-03 Project4/src/ising.cpp
Basic structure for Project3

@@ -1,0 +1,12 @@
+#include "ising.h"
+#include <iomanip>
+#include <iostream>
+
+Ising::Ising() {
+	this->some_number = 2;
+	
+}
+
+void Ising::print_some_number() {
+	cout << this->some_number << endl;
+}

### 2016-11-03 Project4/src/ising.h

@@ -1,0 +1,12 @@
+#pragma once
+
+//#include <armadillo>
+
+class Ising {
+	private:
+		int some_number;
+
+	public:
+		Ising() {}
+		void print_some_number() {}
+};

### 2016-11-03 Project4/src/main.cpp

@@ -1,0 +1,8 @@
+#include "ising.h"
+
+int main(int argc, char *argv[]) {
+	Ising myClass;
+	myClass = Ising::Ising();
+	myClass->print_some_number();
+	return 0;
+}

### 2016-11-03 Project4/article/article.tex
Added article structure

--- 2016-10-24 Project3/article/article.tex
@@
 \usepackage[dvipsnames]{xcolor}
-\usepackage{dirtytalk}
 
@@
 
-\title[Simulation of the Solar System]{The Solar System: an Exercise in Numerical Integration and \\
+\title[Ising model in 2d]{The Ising model in two dimensions \\ 
 	\hrulefill\small{ FYS3150: Computational Physics }\hrulefill}
 
-\author[Winther-Larsen \& Svalheim]{Sebastian G. Winther-Larsen \\ 
-Trygve L. Svalheim \\
-\href{https://github.com/gregwinther/FYS3150/}{\texttt{github.com/gregwinther}}}
+\author[Winther-Larsen]{Sebastian G. Winther-Larsen}
 
@@
 \begin{abstract}
-Two, three, multi. With Euler-Cromer and then Velocity Verlet
+Ising model in 2D. Metropolis algorithm etc. Shit goes down.
 \end{abstract}
@@
 
-This study has three main goals. To show some of the advantages of object-oriented programming, to implement and compare different integration methods and to build a working model of the solar system.
-
 \section{Theory}
 
-\subsection{Newtonian gravtiation}
-When one studies the solar system and the way it moves, one inevitably needs to consider the gravity of the situation. The classical law of gravitation is given by Newton's law
-\begin{equation}
-\label{eq:newtoniangravity}
-F_G = \frac{GM_1M_2}{r^2},
-\end{equation} 
-where $F_G$ is the gravitational force between to bodies, $G$\footnote{$G=6.67408 \times 10^{-11} m^3 kg^{-1} s{-2}$} is the gravitational constant, $M_1$ and $M_2$ are the masses of the two bodies and $r$ is the distance between them.
-
-In three dimensions, by employing Newton's second law of motion, one will get the following componental equations for the acceleration due to gravitational pull on a particular body $i$.
-\begin{equation}
-\label{eq:componentalnewton}
-\frac{d^2x}{dt^2} = \frac{F_{G,x}}{M_i}, \quad
-\frac{d^2y}{dt^2} = \frac{F_{G,y}}{M_i}, \quad
-\frac{d^2z}{dt^2} = \frac{F_{G,z}}{M_i},
-\end{equation}
-each of which can be integrated in order to find the position of the bodies at any given time and for a particular inital velocitiy and position.
-Moreover, a neat vectorial expression for the Newtonian gravity (equation \ref{eq:newtoniangravity}) in three dimensions is
-\begin{equation}
-\vb{F}_G = \frac{GM_1M_2}{r^3}\vb{r}
-\end{equation}
-where $\vb{r}$ gives the posistion of the body in a cartesian coordinate system.
-
-\subsection{Relativity}
-
-The theory of general relativity (GR), put forth by Albert Einstein in 1915, is the current description of gravity in modern physics.
-On curiosity of the solar system that cannot be explained is the perihelion precession of Mercury's orbit.
-For every Mercury year the perihelion of the planets orbit moves sligthly.
-The observed value of this effect is $43$ arc-seconds per century.
-Einstein showed himself that GR could explain this anomaly.
-
-Closed elliptical orbits are a special feature of the factor ($1/r^2$) in Newton's law of gravitation.
-Any change to this factor will result in a change in the orbit.
-For a small such correction, each orbit will be almost the same, but as time progresses the orientation of the elliptical orbit will itself rotate.
-In this study we will not compute a space-time manifold in order to make this correction, but rather add a general relativistic error term to the Newtonian gravitational force (equation \ref{eq:newtoniangravity}.
-\begin{equation}
-\label{eq:relativenewton}
-F_G = \frac{GM_{\odot}M_{Mercury}}{r^2}\left[1+\frac{3l^2}{r^2c^2} \right],
-\end{equation}
-where $M_{MERCURY}$ is the mass of Mercury, $r$ is the distance between Mercury and the Sun, $l=\abs{\vb{r}\times\vb{v}}$ is the orbital angular momentum per unit mass, and $c$ is the speed of light in vacuum.
-
-The precession of Mercury's orbit is not easy to spot simply by plotting the planets orbit, but the angle of the perihelion\footnote{The point in the orbit closest to the sun.} can be calculated easily enough by the following equation
-\begin{equation}
-\label{eq:perangle}
-\theta_p = \arctan \left( \frac{y_p}{x_p}\right)
-\end{equation}
-
-\section{Units}
-In order to make the system easy to work with and the calculations easier as well, a change of units is warranted.
-In this study we will therefore express the mass of any body as a fraction of solar masses, that is $M_{\odot} = 2\times10^{30}kg$.
-This means that the Earth\footnote{In SI-units, $M_{Earth}\approx6\times10^{24}kg$}, for instance will have a mass of $M_{Earth}= 3\times10^{-6}$.
-
-Additionally, the units for distance will be astronmical units $AU$.
-This is the mean distance between the earth and the sun.
-For a simple system and a particular coordinate system with the sun at origin, the initial position for the Earth could simply be $\vb{r}_ {Earth}=(1,0,0)$.
-
-The unit for time will be Earth years, $yr$. This means that velocity will be in units $AU/yr$.
-
-Changing the variables will have consequences for the gravitationan constant $G$ and velocities $v$ as well.
-This can be deduced easily enough by picking a sample system consisting of the earth and the sun.
-Inserting in equation \ref{eq:newtoniangravity} gives
-\begin{equation}
-\label{eq:earthandsun}
-F_G = \frac{GM_{\odot}M_{Earth}}{r^2}
-\end{equation}
-Furthermore, assuming the orbit of the earth is perfectly circular, will give the following relation
-\begin{equation}
-\label{eq:circularearth}
-F_G = \frac{M_{Earth}v^2}{r}
-\end{equation}
-Combining equations \ref{eq:earthandsun} and \ref{eq:circularearth} yields the following
-\begin{equation}
-v^2r=gM_{\odot}=4\pi^2AU^3yr^{-2}
-\end{equation}
-Because the mass of the sun as a fraction of solar masses is $M_{\odot}=1$  and the distance between the earth and the sun is $r=1AU$ we land at
-\begin{equation}
-G=4\pi^2AU^3yr^{-2} \quad v_{Earth}=2\pi AUyr^{-1}
-\end{equation}
-
-\section{Algorithms and implementation}
-
-\subsection{Object orientation}
-Object orientation allows for more general code to be written and for easy reuse.
-Moreover, object orientation makes sense for humans, who tends to classify objects we see and interact with in everyday life.
-In this study we make use of all the advantages of object oriented programming by implementing several classes.
-A class diagram illustration of the program is included in appendix \ref{app:classdiagram}.
-Bear in mind that this diagram does not represent the functionality of the program to a full extent, as many methods and a few classes are left out.
-Rather, it helps one to understand how the class hierarchy is built up and how the program as whole is supposed to function.
-
-The \lstinline|System| class contains all the information about the initial conditions of the system, which integration method should be used and also some helpful methods, like one that writes data to file.
-
-The \lstinline|InitialCondition| superclass contains a method for setting up particles.
-Depending on what system one wants to model, we have implemented three subclasses of \lstinline|IntialCondition|: \lstinline|TwoBody|, \lstinline|ThreeBody| and \lstinline|MultiBody|.
-The name of these classes speek mostly for themselves.
-
-The intersting objects, contained within the \lstinline|InitialCondition| subclasses are instances of the \lstinline|Particle| class.
-The class could have another name, like "planet" or "body" because instances of it will represent the largest bodies in the solar system; the sun and (eventually) all nine planets\footnote{Including Pluto for historical reasons.}.
-Every particle instance takes a position vector, a velcity vector and a mass as inputs and has only these attributes\footnote{The position and velocity vectors are defined by its own class \lstinline|Vec3|.}.
-The extra abstraction to "particle" if therefore fitting, and the class could be employed elsewhere, for instance in a molecular dynamics simulation.
-
-The \lstinline|Integrator| superclass har two subclasses \lstinline|EulerCromer| and \lstinline|VelocityVerlet| each of which must owerwrite the \lstinline|integrateOneStep| method.
-These are the most important part of the program, because it is where the essential part of the two algorithms are implemented.
-The \lstinline|integrateOneStep| method is called iteratively for a specified number of steps from a class instance of the \lstinline|system| class, hence the name of the method.
-Within the \lstinline|integrateOneStep| method, the integration is performed for every particle in the system.
-A further description of the integrator algorithms follows.
-
-\subsection{Euler-Cromer}
-
-Analytically the Euler-Cromer method, or the semi-implicit Euler method, can be expressed in one dimension by the following recursive relations
-\begin{align}
-v_{n+1} &= v_n + a_n dt \\
-x_{n+1} &= x_n + v_{n+1} dt,
-\end{align}
-where $a_n$ is the acceleration, $v_n$ is the velocity and $x_n$ is the position, after a certain number of steps $n$.
-$dt$ is the time step and $t_n = t_0 + ndt$ is the time after $n$ steps.
-One sees that the algorithm is relatively cheap, with $4n$ FLOPS.
-In C++ code the \lstinline|integrateOneStep| method of the \lstinline|EulerCromer| class looks like this
-\begin{lstlisting}
-void EulerCromer::integrateOneStep(std::vector<Particle*> particles) {
-
-	m_system->computeForces();
-
-	for (int i=0; i<particles.size(); i++) {
-		Particle *p = particles.at(i);
-
-		// Acceleration vector
-		vec3 a = (p->getForce()) / (p->getMass());
-
-		// Velocity update
-		p->getVelocity() += m_dt*a;
-
-		// Position update
-		p->getPosition() += m_dt*p->getVelocity();
-	}
-}
-\end{lstlisting}
-Notice that the forces used to compute the acceleration are computed within the \lstinline|System| class and that the method uses three-dimensional vectors for more realistic results.
-
-The Euler Cromer method is a first-order integrator which means that it commits a global error of the order of $dt$. This means that by decreasing the time step one will get a more accurate result.
-
-\subsection{Velocity Verlet}
-The Verlet method is an integration method designed to integrate Newton's equation of motion.
-The method provides good numerical stability and more precise results compared with a regular method like the Euler-Cromer method.
-The reason for this is that it is symplectic, which means that it conserves state-space volume\footnote{One would probably read an entire book to understand this, so we will leave it at that.}.
-
-The velocity Verlet method can also be represented by a (one-dimensional) recursive relation
-\begin{align}
-x_{n+1} &= x_n + v_ndt + \frac{1}{2}a_ndt^2 \\
-v_{n+1} &= v_n + \frac{a_n + a_{n+1}}{2}dt.
-\end{align}
-These equations tell us that the algorithm is slightly more expensive compared to the Euler-Cromer scheme at $11n$ FLOPS. Here follows an implementation of the algorithm in C++.
-
-\begin{lstlisting}
-void VelocityVerlet::integrateOneStep(std::vector<Particle*> particles) {
-
-	m_system->computeForces();
-
-	for (int i=0; i<particles.size(); i++) {
-		Particle *p = particles.at(i);
-
-		// Acceleration vector
-		vec3 a = (p->getForce()) / (p->getMass());
-
-		// Position update
-		p->getPosition() += p->getVelocity()*m_dt + 0.5*a*m_dt*m_dt;
-
-		m_system->computeForces();
-
-		// New accelration vector
-		vec3 anew = (p->getForce()) / (p->getMass());
-
-		// Velocity update
-		p->getVelocity() += (a + anew)*m_dt/2;
-	}
-}
-\end{lstlisting}
-One sees that it is necessary to compute forces and update acceleration once more in order to update the velocity, which means more operations must be conducted.
-Time will show if the extra cost of this algorithm is worth the  increase in precision it promises.
-
-\section{Data}
-As we wish to determine the orbits of our solar system as accurately as possible, we use date from HORIZON Web-Interface, provided by the Jet Propulsion Laboratory (NASA) at the California institute of technology: \url{http://ssd.jpl.nasa.gov/horizons.cgi}.
-Here of data from several celestial bodies can be downloaded.
-It is simple to find both position and velocity for all particles in our bodies at high precision.
-Units of length can be set to $AU$ and velocities to $AU$ per day.
-The velocities must be converted for our simulation.
-We have faith that these data are accurate.
-
-In an arbitrary system, the entire system will usually gain momentum and drift off in some direction over time.
-One must usually correct for such drift be substracting total linear momenum.
-Because the velocity data from NASA is very good, this is unnecessary for our system.
-Total linear momemtum for the solar system is already zero from our frame of reference.
+\section{Algorithm}
 
@@
 
-The object-oriented model for the solar system that has been built is very flexible, as the reader will soon come to know.
-We start by constructing some simple models with two and three bodies, then the entire solar system, and lastly, simulate the perihelion precession of Mercury, as discussed in the theory section.
-
-\subsection{Two-Body system: Earth and Sun}
-
-A two-body system is a meager representation of the solar system, but is the first important stepping stone towards a more interesting and realistic model.
-The two-body system has two instances of the \lstinline|Particle| class, representing the sun and the Earth.
-The orbits are found by solutions from both the Euler-Cromer method and the velocity Verlet method.
-
-In figure \ref{fig:earthcompare} shows a plot of a solution by Euler-Cromer to the left and velocity Verlet to the right.
-The same time step $dt=10^{-2}$ and the numper of points $N=10^6$ is employed in both simulations.
-The sun is the yellow dot in the middle of the plot and the Earth is the blue dot orbiting along the red path.
-The reason the past is wide is because the simulation has run for such a long time that the Earth has travelled several orbits\footnote{One hundred points per revolution and 100000 points should give 1000 revolutions in total.}.
-The difference between the two methods is quite clear.
-The velocity Verlet method har a much narrower ``orbit band'' and is therefore more precise.
-
-\begin{figure}
-	\centering
-	\textbf{Comparison of Euler-Cromer and Velocity Verlet}
-	\includegraphics[width=0.99\textwidth]{../figures/earthcompare.png}
-	\caption{Figure showing the difference in precision between the Euler-Cromer method (left) and the velocity Verlet method (right).
-The time step is $dt=10^{-2}$ and the number of points is $N=10^6$\label{fig:earthcompare}}
-\end{figure}
-
-The model is easily adjustable and one can modify it to show many interesting phenomena.
-In figure \ref{fig:earthescape} is a plot showing the Earth escaping its orbit around the sun.
-The escape velocity was found by trial and error; $v=\sqrt{8}\pi AU/yr$.
-
-\begin{figure}
-	\centering
-	\textbf{Earth Escape Velocity}
-	\includegraphics[width=0.9\textwidth]{../figures/earthescape.png}
-	\caption{Illustration of Earth escaping orbit at $v=\sqrt{8}\pi AU/yr$\label{fig:earthescape}}
-\end{figure}
-
-\begin{table}
-	\centering
-	\caption{Time comparison of Euler-Cromer and velocity Verlet applied to a two-body system. \label{tab:eulervsverlet}}
-	\begin{tabular}{ccc} \hline
-	$N$ & Euler-Cromer [t] & Velocity Verlet [t] \\ \hline
-	$10^{3}$ & 0.023 & 0.036   \\
-	$10^{4}$ & 0.20  & 0.17   \\
-	$10^{5}$ & 1.92  & 1.83   \\
-	$10^{6}$ & 18.42 & 17.88 \\ \hline
-	\end{tabular}
-\end{table}
-
-Table \ref{tab:eulervsverlet} shows a comparison of how much time the two algorithm requires to handle a problem of increasing size.
-The two algorithms were applied to the same two-body problem.
-As one can see, it appears that the velocity Verlet method is a bit quicker, but these results are probably not statistically significant.
-More on this in the discussion below.
-
-\subsection{Three-Body system: Earth, Sun and Jupiter}
-
-A natural progression from a two-body system is a three-body system.
-We have implemented a model consisting of particles representing Jupiter, Earth, and the Sun.
-The top left subfigure of figure \ref{fig:jupiter} in two dimensions.
-The velocity Verlet method is employed here.
-
-\begin{figure}
-	\centering
-	\textbf{Earth, Sun and Jupiter}
-	\includegraphics[width=0.99\textwidth]{../figures/threebody.png}
-	\caption{Model outcome for a Jupiter with different masses.
-Top left: Normal mass.
-Top right: Jupiter mass $10\times$ normal.
-Bottom: Jupiter mass $1000\times$ normal, in 2D (left) and 3D (right)\label{fig:jupiter}}
-\end{figure}
-
-The rest of the subfigures in figure \ref{fig:jupiter} show what would happen if Jupiter had larger mass than it has.
-The top right subplot shows this scenario if Jupiter's mass was $10\times$ larger than normal.
-One can see that this would drag the Earth along into the Sun, which stays mostly put, and likely skew the entire solar system.
-
-The two bottom figures show what would happen i Jupiter would have had $1000\times$ normal mass.
-The plot speaks for itself; the entire system is pulled apart.
-This situation is shows in 2D to the left and 3D to the right.
-
-\subsection{Multi-Body system: all planets}
-
-\begin{figure}
-	\centering
-	\textbf{Multi-body model of the Solar System}
-	\includegraphics[width=0.9\textwidth]{../figures/multibody.png}
-	\caption{All planets of our solar system (including Pluto) \label{fig:multibody}}
-\end{figure}
-
-Finally for the pinnacle of this study, a model of the entire solar system!\footnote{Excluding moons and other satelites, but including Pluto.} Figure \ref{fig:multibody} shows a multi-body simulation representing the solar system, integrated using the velocity Verlet algorithm.
-One can see that the algorithm handles the problem quite well.
-We were even able to make an anmation of this system in MATLAB, which was a delightful exercise.
-
-\subsection{The perihelion precession of Mercury}
-
-Figure \ref{fig:perangle} shows the angel of the perihelion of Mercury calculated by equation \ref{eq:perangle} in the theory section.
-This plot is over a period of 30 Earth years and with a time step of $dt = 10^{-6}yrs$.
-One can see a clear trend in the angle of the perihelion as predicted.
-We get a slope of the perhelion angle versus number of orbist of approximately $10^{-6}$, for a century on earth Mercury makes $415.2091$ orbits, which gives us a perhelion precession of $\approx1.5''/100yrs$.
-This much lower than the expected precession of $43''/100yrs$.
-
-\begin{figure}
-	\centering
-	\includegraphics[width=0.8\textwidth]{../figures/periangle30yrs.png}
-	\caption{The precession of Mercury's orbit\label{fig:perangle}}
-\end{figure}
-
 \section{Discussion}
-
-\subsection{Object Orientation}
-This project has been a nice exercise in object oriented thinking.
-We have built a larger and larger system by adding more class instances and building subclasses such as different integrators.
-Object oriented programming is close to how we as humans usually thinks about all things in the world.
-A planet is something that has mass, a position, velocity and some other factors.
-However, when one learns how to program learns object orientation at a late stage of the learning process, when one has been focusing on a more procedural programming technique.
-
-Most widely used programming languages today are object-oriented to a greater or lesser extent: Java, C++, Python, PHP, Ruby, Smalltalk and Common Lisp, to name a few.
-Critique of object orientation usually emphasizes that design and modelling comes at the expense of other important aspects like the actual computation and algorithms.
-The creator of Erlang, Joe Armstrong, is quoted as saying, \say{The problem with object-oriented languages is they've got all this implicit environment that they carry around with them.
-You wanted a banana but what you got was a gorilla holding the banana and the entire jungle}.
-
-As fairly level-headed Norwegian patriots\footnote{A friendly reminder to the reader: the creators of the first object oriented languas, Simula, were Norwegian: Ole-Johan Dahl, Kristen Nygaard.} we cannot end the discussion about object orientation there, but rather by pointing out the advantages.
-Object Oriented Programming provides a clear modular structure for programs and makes it easy to maintain and modify existing code thereby reducing lowering programming ``cost''.
-
-\subsection{Integration Algorithm}
-
-We have seen that the velocity Verlet algorithm is much more precise than the Euler-Cromer method (figure \ref{tab:eulervsverlet}).
-In addition it appears that it may also be quicker (table \ref{tab:eulervsverlet}), which is a surprise since the Verlet algorithm seems to include more FLOPS than the Euler-Cromer method.
-This leads us to believe that the velocity Verlet may not be that much faster, at least we can conclude that the results are not statistically significant since we only have one observation per method per value for $N$.
-An improvement on this study would have been to conduct such a test.
-We propose larger sampling of code run time and, for simplicity, a paired Student's t-test as a start to answer this question\cite{student}.
-
-The discussion of an Euler method versus a Verlet method is a rather big one.
-Both methods are quite quick cheap which is what you want, and they are usually the two methods of choice if one does not need entirely exact results, but results that are ``good enough''.
-A field where this discussion prominent is in physics engine design for video games.
-With more computing power one would probably go for a more sophisticated method like a foruth degree Runge-Kutta method\cite{physicsengine}.
-
-It is worth metioning something about the symplecticity of the velocity Verlet method.
-As stated previously, a symplectic method will be good at simulating systems with energy conservation.
-We believe that the solar system has conserved energy, but our model of it may not be.
-Looking further into this matter we have found that both energy and angular momentum is conserved for the two-body Earth-Sun system.
-This would warrant further use of the velocity Verlet scheme.
-See appendix \ref{app:energy} for a sample readout of energies and angular momenta.
-
-\subsection{Mercury}
-
-We tried to compute the perihelion precessrion of Mercury.
-We found it to be $\approx1.5''/100yrs$ when it should have been much more, id est $43''/100yrs$.
-This error can stem from several sources.
-Most obvious is that we have done something wrong and have been unable to find where.
-Antoher likely source of uncertainty is that the method is that the integration method is not precise enough.
-We had problems decreasing the step size further, and extending the simulation to a full century because the job was too much for our hardware.
-One last probelm that must be mentioned is that Mercury's precession cannot be explained by the simple relativistic correction we are employing.
 
@@
 
-\begin{thebibliography}{9}
-
-\bibitem{erlang}
-	Armstrong, Joe (2009),
-	\emph{Coders at Work: Reflection on the Craft of Programming},
-	Peter Seibel, ed.
-
-\bibitem{student}
-	Hazewinkel, Michiel, ed. (2001), 
-	Student test,
-	\emph{Encyclopedia of Mathematics},
-	Springer.
-
-\bibitem{physicsengine}
-	Boesch, Florian (2010, August 28),
-	\emph{Integration by Example - Euler vs Verlet vs Runge-Kutta},
-	Retrieved October 24 2016 from \url{http://codeflow.org/entries/2010/aug/28/integration-by-example-euler-vs-verlet-vs-runge-kutta/},
-
-\end{thebibliography}
-
-\pagebreak
-
-\begin{appendix}
-
-\section{Class Diagram}
-\label{app:classdiagram}
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=0.9\textwidth]{../figures/classdiagram.png}
-\end{figure}
-
-\section{Energy Read-out}
-\label{app:energy}
-\begin{lstlisting}[keywordstyle=\ttfamily, identifierstyle=\ttfamily]
-Energies printed in C++:
-
-Energies for Earth-Sun
-EULER-CROMER
-Angular momentum: 39.4784*m
-Step: 1000000  E = -2e-05   Ek =1.9739e-05  Ep =-3.9478e-05
-VERLET
-Angular momentum: 39.4784*m
-Step: 1000000  E = -2e-05   Ek =1.9739e-05  Ep =-3.9478e-05
-
-Energies for three-body system
-Step: 1000000  E = -0.004   Ek = 0.0040197  Ep =-0.0077056
-For 10m_j
-Step: 1000000  E =  -0.04   Ek =  0.041006  Ep = -0.077361
-For 1000m_j
-Step: 100000   E =     -4   Ek =     3.314  Ep =    -6.944
-
-Energies for multi-body system dt=1e-5 N=1e5
-Step: 100000   E = -0.004   Ek = 0.0040429  Ep =-0.0084804
-\end{lstlisting}
-
-\end{appendix}
-
 \end{document}

### 2016-11-06 Project4/article/article.tex
Theory OK. 2-by-2 system solved

@@
 \usepackage[dvipsnames]{xcolor}
+\usepackage{tikz}
 
@@
 
-\author[Winther-Larsen]{Sebastian G. Winther-Larsen}
+\author[Winther-Larsen]{Sebastian G. Winther-Larsen \\
+\href{https://github.com/gregwinther/FYS3150/}{\texttt{github.com/gregwinther}}}
 
@@
 \section{Theory}
+The model we will employ in this study of phase transitions at finite temperature for magnetic systems is the Ising model, named after Ernst Ising who solved a one-dimensional variant of the model\cite{Ising}.
+The two-dimensional square lattice model, employed herein, was given an analytic description much later by Lars Onsager\cite{Onsager}.
+
+\subsection{The general Ising model}
+Let $\Lambda$ be a set of lattice positions, each with adjacent positions, forming a $d$-dimensional lattice.
+For each lattice site, $k \in \Lambda$, there is a discrete variable $\sigma_k \in \{+1,-1\}$, representing the spin of the site, $\uparrow$ and $\downarrow$, respectively.
+A spin configuration $\sigma$ is a specific spin configuration of the lattice.
+
+For two adjacent sites, $i,j \in \Lambda$, one has an interaction $J_{ij}$.
+A site $j \in \Lambda$ will also have an external magnetic field $h_j$ interacting with it.
+The energy of a specific system is given by the following Hamiltonian
+\begin{equation}
+\label{eq:IsingHamilton}
+E = H(\sigma)= -\sum_{\ev{ik}}J_{ij}\sigma_i \sigma_j - \mu \sum_j h_j \sigma_j,
+\end{equation}
+where the first sum is over pairs of adjacent spins. The notation $\ev{ij}$ indicates that sites $i$ and $j$ are nearest neighbours. In the second sum, $\mu$ is the magnetic moment.
+
+The probability of a certain configurations is given by the Boltzmann distribution
+\begin{equation}
+\label{eq:Boltzmann}
+P_\beta(\sigma) = \frac{e^{-\beta H(\sigma)}}{Z_\beta}
+\end{equation}
+where $Z_\beta$ is the partition function acting as a normalisation constant, and $\beta=(k_BT)^{-1}$.
+This would mean that by increasing the temperature $T$, finding the system in one particular configuration decreases.
+It must be so, because with a relatively higher temperature, one would expect previously unfavourable configuration to become more feasible.
+
+There are two important expected values that are important in order to characterise a magnetic system. The \emph{mean energy} of the system is
+\begin{equation}
+\label{eq:meanenergy}
+\ev{E} = \sum_{i=1}^ME_iP_\beta(\sigma) = \frac{1}{Z}\sum_{i=1}^ME_ie^{-\beta E_i},
+\end{equation}
+and the \emph{mean magnetisation} is
+\begin{equation}
+\label{eq:meanmagnet}
+\ev{\mathcal{M}} = \sum_{i=1}^M\mathcal{M}_iP_\beta(\sigma) = \frac{1}{Z}\sum_{i=1}^M\mathcal{M}_ie^{\beta E_i},
+\end{equation}
+where $\mathcal{M}_i = \sum_{j \in \Lambda}\sigma_j$ for all configurations $\sigma$, and $M$ denotes the number of possible configurations.
+Another quantity of interest is \emph{magnetic susceptibility} $\chi$ which tells us how much an extensive parameter changes when an intensive parameter increases.
+It is given by
+\begin{equation}
+\label{eq:susceptibility}
+\chi = \frac{1}{k_BT}(\ev{\mathcal{M}^2} - \ev{\mathcal{M}}^2).
+\end{equation}
+The \emph{heat capacity}, at constant volume, is given by
+\begin{equation}
+\label{eq:heatcap}
+C_V = \frac{1}{k_BT^2}(\ev{E^2}-\ev{E}^2)
+\end{equation}
+
+The minus sign on each term of the Hamiltonian $H(\sigma)$ in equation \ref{eq:IsingHamilton} is conventional.
+By this sign convention, the Ising model can be classified according to the sign of the interaction.
+If, for all pairs $i,j$:
+\begin{itemize}
+\item $J_{ij} > 0$, the interaction is ferromagnetic,
+\item $J_{ij} < 0$, the interaction is anti-ferromagnetic,
+\item $J_{ij} = 0$, the spins are non-interacting.
+\end{itemize}
+In a ferromagnetic Ising model, spins desire to be aligned: the configurations in which adjacent spins are of the same sign have higher probability.
+In an anti-ferromagnetic model, adjacent spins tend to have opposite signs.
+
+\subsection{Simplified ferromagnetic Ising model}
+The general Ising model, as described above, will not be used in this study, but a much simpler version of it.
+Firstly, we will examine a system with no external magnetic field, as was originally solved analytically by Onsager\cite{Onsager}.
+Because the second sum in the Hamiltonian in equation \ref{eq:IsingHamilton} is zero, and we are left with
+\begin{equation}
+\label{eq:IsingHamSimple1}
+E = H(\sigma) = -\sum_{\ev{ik}}J_{ij}\sigma_i\sigma_j.
+\end{equation} 
+Secondly, we assume that the coupling constant $J_{ij}$, that describe the interaction of a spin with its neighbour, to be constant.
+That is $J_{ij} = J \forall i,j \in \Lambda$ and equation \ref{eq:IsingHamSimple1} is simplified further to
+\begin{equation}
+\label{eq:IsingHamSimple2}
+E = H(\sigma) = -J \sum_{\ev{ik}}\sigma_i\sigma_j
+\end{equation}
+
+In this study, we will assume that we have a ferromagnetic ordering, id est  $J>0$.
+This means that neighbouring spins are aligned, because it would lead to lower energy.
+It is easy to see why it must be so, as $\sigma_i\sigma_j=1$ whenever spin $i$ and $j$ have the same sign.
+
+\subsection{Example: $2\times2$ Ising model}
+It would be beneficial to test the waters of the ocean of the vast ocean that is the Ising model, by a two-dimensional model with lattice dimension $L=2$ and periodic boundary conditions.
+This model has $s=2^4=16.$ different configurations $\sigma$.
+The energy of a given configuration would be
+\begin{equation*}
+E_i = -J \sum_{\ev{kl}}^4 \sigma_k \sigma_l.
+\end{equation*}
+
+Figure \ref{fig:2by2lattice} shows an arbitrary configuration of a $2\times2$ spin lattice. The energy for this configuration is
+\begin{equation*}
+E_i = -J((+1)(-1)+(+1)(-1)+(+1)(-1)+(+1)(-1)) = 8J
+\end{equation*}
+which happens to be the highest energy possible for the system.
+When all spins are parallel we see that $E_i=-8J$ which is the lowest energy possible.
+If only one spin would differ from the others, the energy would be $E_i = 0$ and so on.
+Of the $16$ possible configurations, several will have degeneracies ($\Omega(E_i)$), which corresponds to the number of configurations with the same energy.
+Moreover, the magnetisation of a particular configuration is simply the sum of the spins and is easy to calculate.
+All the possible configurations of this system can be found in table \ref{tab:2by2lattice}.
+
+\begin{figure}[h]
+	\centering
+	\begin{tikzpicture}
+	\filldraw 	(0,0) circle(2pt) node[align=left, below]{$\uparrow$}-- 
+				(0,1) circle(2pt) node[align=left, above]{$\uparrow$}-- 
+				(1,1) circle(2pt) node[align=right, above]{$\downarrow$}-- 
+				(1,0) circle(2pt) node[align=right, below]{$\downarrow$}-- (0,0);
+	\end{tikzpicture}
+	\caption{Sample $2\times2$ spin lattice.}
+	\label{fig:2by2lattice}
+\end{figure}
+
+\begin{table}
+	\centering
+	\caption{All possible configurations of a $2\times2$ Ising model}
+	\begin{tabular}{cccc} \hline
+	No of $\uparrow$ & $\Omega(E_i)$ & $E_i$ & $\mathcal{M}_i$ \\ \hline
+	$4$ & $1$ & $-8J$ & $4$  \\
+	$3$ & $4$ & $0$   & $2$ \\
+	$2$ & $4$ & $0$   & $0$  \\
+	$2$ & $2$ & $8J$  & $0$  \\
+	$1$ & $4$ & $0$   & $-2$ \\
+	$0$ & $1$ & $-8J$ & $-4$ \\ \hline
+	\end{tabular}
+	\label{tab:2by2lattice}
+\end{table}
+
+Now to compute the physical quantities as discussed above, expected value for energy $\ev{E}$, expected value for magnetisation $\ev{\mathcal{M}}$, expected value for specific heat $\ev{C_V}$, and susceptibility $\chi$.
+For the $2\times2$ Ising model these quantities have closed form expressions.
+The partition function for the system is given by
+\begin{equation}
+\label{2by2partition}
+Z = \sum_{i=1}^16 e^{-\beta E_i} = e^{\beta8J} + 12 + 2e^{-\beta8J} + e^{\beta8J} = 4\cosh(\beta8J) + 12
+\end{equation}
+The expected energy is
+\begin{equation}
+\label{2by2expectedE}
+\ev{E} = -\frac{\partial}{\partial\beta} \ln Z = -\frac{\partial}{\partial \beta}\ln(4\cosh(\beta J) + 12)
+ = -8J \frac{\sinh(8\beta J)}{\cosh(8\beta J) + 3}.
+\end{equation}
+The mean magnetisation for this system is easiest to compute with equation \ref{eq:meanmagnet}, merely adding all possible states and dividing by the partition function.
+\begin{equation}
+\label{eq:2by2meanmagnet}
+\ev{\mathcal{M}}= \frac{1}{Z}(-4e^{8\beta J} - 8e^0 + 8e^0 +8e^{8\beta J}) = 0
+\end{equation}
+the expected absolute magnetisation, on the other hand, becomes
+\begin{equation}
+\label{eq:2by2absmeanmagnet}
+\ev{\abs{\mathcal{M}}} = \frac{1}{Z}(4e^{8\beta J} + 8e^0 + 8e^0 + 4e^{8\beta J}) = \frac{4 +2 e^{8\beta J}}{\cosh(8\beta J) +3}.
+\end{equation}
+The expected value for specific heat is
+\begin{equation}
+\label{eq:2by2specificheat}
+\ev{C_V} = \frac{1}{k_bT^2}\frac{\partial^2}{\partial\beta^2}
+\end{equation}
+inserting equation \ref{2by2expectedE} gives
+\begin{align*}
+\ev{C_V} 	&= -\frac{1}{k_BT^2}\frac{\partial}{\partial\beta}\left(-8J\frac{\sinh(8\beta J)}{\cosh(8\beta J)+3} \right) \\
+			&= \frac{1}{k_BT^2}\left(\frac{64J^2\cosh(8\beta J)}{\cosh(8\beta J)+3}-\frac{64J^2\sinh^2(8\beta J)}{(\cosh(8\beta J)+3)^2}\right) \\
+			&= \frac{1}{k_BT^2}\frac{64J^2}{\cosh(8\beta J) + 3}\left(\cosh(8\beta J) - \frac{\sinh^2(8\beta J)}{\cosh(8\beta J) +3} \right)
+\end{align*}
+
+The susceptibility $\chi$ of a thermodynamic system is easy to compute if one knows what the variance of magnetisation, $\sigma_{\mathcal{M}}^2)$, is. Rewriting equation \ref{eq:susceptibility} gives
+\begin{equation}
+\label{eq:susceptibility2}
+\chi = \frac{1}{k_BT}\sigma_{\mathcal{M}}^2.
+\end{equation}
+Using equations \ref{eq:2by2meanmagnet} and \ref{eq:2by2absmeanmagnet} one can deduce that the variance of the magnetisation must be
+\begin{equation}
+\label{eq:2by2magnetvar}
+\sigma_{\mathcal{M}}^2 = \ev{\mathcal{M}^2} - \ev{\mathcal{M}}^2 = \frac{32}{Z}(e^{8\beta J} +1) - 0 = \frac{8(e^{8\beta J} + 1)}{\cosh(8\beta J) + 3}.
+\end{equation}
+Inserting \ref{eq:2by2magnetvar} into \ref{eq:susceptibility2} yields the susceptibility for the system
+\begin{equation}
+\label{eq:2by2susceptibility}
+\chi = \frac{8(e^{8\beta J}+1)}{k_BT(\cosh(8\beta T) + 3)}
+\end{equation}
 
@@
 
+\begin{thebibliography}{9}
+
+\bibitem{Ising} Ising, E., Beitrag zur Theorie des Ferromagnetismus,
+	\emph{Z. Phys.,} 31, pp. 253-258 (1925).
+
+\bibitem{Onsager} Onsager, L., Crystal statistics. I. A two-dimensional model with an order-disorder transition,
+	\emph{Physical Review,} Series II, 65 (3-4), pp. 117-149 (1944).
+
+\end{thebibliography}
+
 \end{document}

### 2016-11-06 Project4/src/ising.cpp
Ising class methods, PBC, initialising, CML arguments

@@ -1,12 +1,58 @@
 #include "ising.h"
 #include <iomanip>
 #include <iostream>
+#include <time.h>
 
-Ising::Ising() {
-	this->some_number = 2;
+Ising::Ising(int lattice_dimension) {
+	this->lattice_dimension = lattice_dimension;
+
+	lattice = arma::mat(lattice_dimension, lattice_dimension);
+
+
+	energy = 0;
+	magnetisation = 0;
 	
 }
 
-void Ising::print_some_number() {
-	cout << this->some_number << endl;
+void Ising::initialise_system(double temp) {
+	temperature = temp;
+
+	// reset variables
+	energy = 0;
+	magnetisation = 0;
+
+	// Time as seed for random number generator
+	srand(time(NULL));
+
+	for (int i = 0; i < lattice_dimension; i++) {
+		for (int j = 0; j < lattice_dimension; j++) {
+
+			// Start in random configuration
+			lattice(i, j) = (((double) rand() / (RAND_MAX)) > 0.5) ? 1: -1;
+
+			magnetisation += lattice(i,j);
+
+			std::cout << lattice(i, j) << std::endl;
+
+		}
+	}
+
+	std::cout << "Magnetisation = " << magnetisation << std::endl;
+
+	// Initial energy - Mulitply every sites spin direction with neighbhours
+	for (int i = 0; i < lattice_dimension; ++i) {
+		for (int j = 0; j < lattice_dimension; ++j) {
+			energy -= lattice(i, j) 
+					* (lattice(periodic_boundary_translation(i, lattice_dimension, -1), j)
+					+  lattice(i, periodic_boundary_translation(j, lattice_dimension, -1)));
+		}
+	}
+
+	std::cout << "Energy = " << energy << std::endl;
+
+}
+
+// Periodic boundary conditions: "wraps around"
+int Ising::periodic_boundary_translation(int x, int dimension, int translation) {
+	return (x + dimension + translation) % dimension;
 }

### 2016-11-06 Project4/src/ising.h

@@ -1,12 +1,28 @@
-#pragma once
+//#pragma once
 
-//#include <armadillo>
+#include <armadillo>
 
 class Ising {
 	private:
-		int some_number;
+		int lattice_dimension;
+
+		arma::mat lattice;
+
+		double temperature;
+
+		double expected_energy;
+		double expected_magnetisation;
+		double specific_heat;
+		double susceptibility;
+
+		double energy;
+		double magnetisation;
+
+		int periodic_boundary_translation(int x, int dimension, int translation);
 
 	public:
-		Ising() {}
-		void print_some_number() {}
+		Ising(int lattice_dimension);
+		
+		void initialise_system(double temp);
+		void simulate(int cycles);
 };

### 2016-11-06 Project4/src/main.cpp

@@ -1,8 +1,11 @@
 #include "ising.h"
 
 int main(int argc, char *argv[]) {
-	Ising myClass;
-	myClass = Ising::Ising();
-	myClass->print_some_number();
+	
+	int dim = atoi(argv[1]);
+	int temp = atoi(argv[2]);
+	Ising myClass = Ising(dim);
+	myClass.initialise_system(temp);
+
 	return 0;
 }

### 2016-11-16 Project4/scripts/analytic.py
Done for now

@@ -1,0 +1,22 @@
+import numpy as np 
+from matplotlib import pyplot as plt 
+
+N = 4
+temp = 1.0
+
+exp_E     = -8*np.sinh(8/temp)/(3 + np.cosh(8/temp))
+exp_M     = 0
+exp_M_abs = 2*(np.exp(8/temp) + 2)/(np.cosh(8/temp) + 3) 
+C_v       = (64./(temp*temp))*(np.cosh(8/temp)/(3 + np.cosh(8/temp))\
+			 - (np.sinh(8/temp)/(3+np.cosh(8/temp)))**2)
+X         = (8./temp) * (np.exp(8./temp) + 1)/(3+np.cosh(8./temp))
+X_abs     = (4./(temp*(np.cosh(8./temp) + 3)))*(2*(np.exp(8./temp) + 1)\
+			 - ((np.exp(8./temp)+2)**2)/(np.cosh(8./temp)+3))
+
+
+print(exp_E/N)
+print(exp_M/N)
+print(exp_M_abs/N)
+print(C_v/N)
+print(X/N)
+print(X_abs/N)

### 2016-11-16 Project4/scripts/plot4d.py

@@ -1,0 +1,85 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import matplotlib.mlab as mlab
+from matplotlib import rc
+from scipy import special
+c= 60 #Number of cores!
+T,E,Cv,M,X,Mabs,Nacc = np.loadtxt("../data/expectations1.dat",usecols=(0,1,2,3,4,5,6), unpack=True)
+Ts,Es,Cvs,Ms,Xs,Mabss,Naccs = np.loadtxt("../data/expectations24_2.dat",usecols=(0,1,2,3,4,5,6), unpack=True)
+#cycles = len(T)
+#time = range(cycles)
+weights = np.ones_like(E[20000:])/len(E[20000:])
+
+sigma = np.mean(Cv)*np.mean(T)**2
+sigmaT1 = np.var(E)
+sigmaT24 = np.var(Es)
+print sigma, sigmaT1, sigmaT24
+# ---------------- New color scheme -----------------
+# These are the "Tableau 20" colors as RGB.    
+tableau20 = [(31, 119, 180), (174, 199, 232), (255, 127, 14), (255, 187, 120),    
+             (44, 160, 44), (152, 223, 138), (214, 39, 40), (255, 152, 150),    
+             (148, 103, 189), (197, 176, 213), (140, 86, 75), (196, 156, 148),    
+             (227, 119, 194), (247, 182, 210), (127, 127, 127), (199, 199, 199),    
+             (188, 189, 34), (219, 219, 141), (23, 190, 207), (158, 218, 229)]    
+  
+# Scale the RGB values to the [0, 1] range, which is the format matplotlib accepts.    
+for i in range(len(tableau20)):    
+    r, g, b = tableau20[i]    
+    tableau20[i] = (r / 255., g / 255., b / 255.) 
+#-------------------------------------------------------
+fig = plt.figure()
+#Equilibrium around 3e5/c=5000
+plt.subplot(211)
+plt.hist(E[20000:],2,color=tableau20[2],weights=weights)#,range=[-2.0, -1.8])
+plt.title('Energy probability after EQ T=1',fontsize=12)
+plt.ylabel('Probability',fontsize=10)
+#plt.xlabel('Energy',fontsize=10)
+#--------------------Configuration------------------------------
+# Remove the plot frame lines. They are unnecessary chartjunk.    
+ax = plt.subplot(211)  
+ax.yaxis.grid()  
+ax.spines["top"].set_visible(False)    
[…]

### 2016-11-16 Project4/scripts/plt2by2.py

@@ -1,0 +1,14 @@
+import numpy as np 
+from matplotlib import pyplot as plt
+
+
+filename = "expectations60.dat"
+
+
+Ts,Es,Cvs,Ms,Xs,Mabss = np.loadtxt(filename, usecols=(0,1,2,3,4,5), unpack=True)
+
+plt.plot(Ts, Xs)
+plt.xlabel("Temp")
+plt.ylabel("E")
+plt.title("L=60")
+plt.show()

### 2016-11-16 Project4/src/ising.cpp

@@ -2,21 +2,24 @@
 #include <iomanip>
 #include <iostream>
 #include <time.h>
-#include <mpi.h>
 #include <string.h>
-
+#include <fstream>
+#include <stdlib.h>
+#include "lib.h"
+#include <mpi.h>
 
 Ising::Ising(int lattice_dimension, int rank, int world_size) {
 
     this->lattice_dimension = lattice_dimension;
     this->no_of_spins = lattice_dimension*lattice_dimension;
-    this->expectations_filename = "expectations";
+    this->expectations_filename = "expectations.dat";
     this->rank = rank;
     this->world_size = world_size;
 
-    lattice = arma::mat(lattice_dimension, lattice_dimension);
+    //lattice = arma::mat(lattice_dimension, lattice_dimension);
+    this->lattice = (double**) matrix(lattice_dimension, lattice_dimension, sizeof(double));
 
-    delta_energy = arma::vec(17);
+    double delta_energy[17];
 
     energy = 0;
     magnetisation = 0;
@@ -29,9 +32,12 @@ void Ising::initialise_system(double temp) {
     // Reset variables
     energy = 0;
     magnetisation = 0;
+    no_of_accepted_states = 0;
 
     // Possible energies
-    for (int i = -8; i <= 8; i += 4) delta_energy(i + 8) = exp(-i/temperature);
+    for (int i = -8; i <= 8; i++) delta_energy[i + 8] = 0;
+    for (int i = -8; i <= 8; i += 4) delta_energy[i + 8] = exp(-i/temperature);
+
 
     // Time as seed for random number generator
     srand(time(NULL));
@@ -40,9 +46,10 @@ void Ising::initialise_system(double temp) {
         for (int j = 0; j < lattice_dimension; j++) {
 
             // Start in random configuration
-            lattice(i, j) = (((double) rand() / (RAND_MAX)) > 0.5) ? 1: -1;
+            lattice[i][j] = (((double) rand() / (RAND_MAX)) > 0.5) ? 1: -1;
 
-            magnetisation += lattice(i,j);
+            // Magnetisation
+            magnetisation += lattice[i][j];
 
         }
     }
@@ -50,14 +57,14 @@ void Ising::initialise_system(double temp) {
     // Initial energy - Mulitply every sites spin direction with neighbhours
     for (int i = 0; i < lattice_dimension; ++i) {
         for (int j = 0; j < lattice_dimension; ++j) {
[…]

### 2016-11-16 Project4/src/ising.h

@@ -1,5 +1,4 @@
-#pragma once
-#include <armadillo>
+#include <string>
 
 class Ising {
     private:
@@ -11,10 +10,9 @@ class Ising {
         int rank;
         int world_size;
 
-		arma::mat lattice;
+        double** lattice;
 
-        arma::vec delta_energy = arma::zeros<arma::mat>(17);
-        //arma::vec* expected_values;// = arma::zeros<arma::mat>(5);
+        double delta_energy [17];
         std::string expectations_filename;
 
 		double temperature;
@@ -32,14 +30,14 @@ class Ising {
 
 		int periodic_boundary_translation(int x, int dimension, int translation);
         double get_energy_of_site(int x, int y);
-        void output(int current_cycle, arma::vec &expected_values);
+        void output(int current_cycle, double (&expected_values)[5]);
 
     public:
 
         Ising(int lattice_dimension, int rank, int world_size);
 		
 		void initialise_system(double temp);
-        void simulate(int cycles, arma::vec &expected_values);
+        void simulate(int cycles, double (&expected_values)[5]);
         void write_to_terminal();
 
 };

### 2016-11-16 Project4/src/lib.h

@@ -1,0 +1,57 @@
+/*
+ * The definition module
+ *                      lib.h
+ * for the library function common for all C programs.
+ */
+
+// Standard ANSI-C++ include files
+#pragma clang diagnostic push
+#pragma clang diagnostic ignored "-Wdeprecated-register"
+#pragma clang diagnostic pop
+
+#include <iostream>
+#include <new>
+#include <cstdio>
+#include <cstdlib>
+#include <cmath>
+#include <cstring>
+
+using namespace std;
+
+
+#define   NULL_PTR   (void *) 0
+#define   ZERO       1.0E-10
+#define   UL         unsigned long
+
+/* a macro used in function pythag() */
+
+static float sqrarg;
+#define SQR(a) ((sqrarg = (a)) == 0.0 ? 0.0 : sqrarg * sqrarg)
+
+
+/* Macro definitions for integer arguments only */
+
+#define   SIGN(a,b) ((b)<0 ? -fabs(a) : fabs(a))
+
+// Function declarations
+
+void  **matrix(int, int, int);
+void free_matrix(void **);
+void rk4(double *, double *, int, double, double, double  *,
+         void (*derivs)(double, double *, double *));
+void ludcmp(double **, int, int *, double*);
+void lubksb(double **, int, int *, double *);
+void tqli(double *, double *, int, double **);
+void tred2(double **, int, double *, double *);
+double pythag(double, double);
+void gauleg(double, double, double *, double *, int);
+void jacobi(double** a, double* d, double** v, int n, int& nrot);
+double rectangle_rule(double, double, int, double (*func)(double));
+double trapezoidal_rule(double, double, int, double (*func)(double));
+void spline(double *, double *, int, double, double, double *);
+void splint(double *, double *, double *, int, double, double *);
+void polint(double *, double *, int, double, double *, double *);
+double ran0(long *);
+double ran1(long *);
+double ran2(long *);
+double ran3(long *);

### 2016-11-16 Project4/src/main.cpp

@@ -1,5 +1,9 @@
 #include "ising.h"
 #include <mpi.h>
+#include <iostream>
+#include <fstream>
+#include <stdlib.h>
+#include <stdio.h>
 
 // input arguments: [dim] [temp_init] [temp_final] [temp_step] [cycles]
 int main(int argc, char *argv[]) {
@@ -7,8 +11,8 @@ int main(int argc, char *argv[]) {
     int world_size;
     int world_rank;
 
-    arma::vec expected_values = arma::zeros<arma::mat>(5); //arma::zeros<arma::mat>(5); //= arma::zeros<arma::mat>(5);
-    arma::vec tot_expected_values = arma::zeros<arma::mat>(5);
+    double expected_values [5]; //arma::zeros<arma::mat>(5); //= arma::zeros<arma::mat>(5);
+
 
     MPI_Init(&argc, &argv);
     MPI_Comm_size(MPI_COMM_WORLD, &world_size);
@@ -20,10 +24,6 @@ int main(int argc, char *argv[]) {
     double temp_step  = atof(argv[4]);
     int cycles = atoi(argv[5]);
 
-    // dont need this!
-    //arma::vec temp_steps = arma::linspace<arma::vec>(temp_init, temp_final, stuff_to_do);
-    //std::cout << temp_steps << std::endl;
-
     MPI_Bcast(&cycles, 1, MPI_INT, 0, MPI_COMM_WORLD);
     MPI_Bcast(&temp_init, 1, MPI_DOUBLE, 0, MPI_COMM_WORLD);
     MPI_Bcast(&temp_final, 1, MPI_DOUBLE, 0, MPI_COMM_WORLD);
@@ -34,9 +34,14 @@ int main(int argc, char *argv[]) {
     timeStart = MPI_Wtime();
     Ising myFriendTheIsingModel = Ising(dim, world_rank, world_size);
     for (float temp=temp_init; temp <= temp_final; temp += temp_step) {
+        for (int i = 0; i < 5; i ++) {
+            expected_values[i] = 0;
+        }
         myFriendTheIsingModel.initialise_system(temp);
         myFriendTheIsingModel.simulate(cycles, expected_values);
+        if (world_rank == 0) myFriendTheIsingModel.write_to_terminal();
     }
+
     timeStop = MPI_Wtime();
 
     totalTime = timeStop - timeStart;

### 2016-11-16 Project4/article/article.tex
Ising class methods, PBC, initialising, CML arguments

@@
 
-\title[Ising model in 2d]{The Ising model in two dimensions \\ 
+\title[Ising Model]{The Ising model and the Metropolis Algorithm \\ 
 	\hrulefill\small{ FYS3150: Computational Physics }\hrulefill}
@@
 \begin{abstract}
-Ising model in 2D. Metropolis algorithm etc. Shit goes down.
+The aim of this project is to determine the Curie temperature $T_c$, of a system undergoing phase transitions.
+When the temperature is raised above this critical level a ferromagnetic material will change into a paramagnetic material.
+The aim is to solve the problem employing the Metropolis-Hastings algorithm and Monte Carlo methods.
 \end{abstract}
@@
 \section{Introduction}
+This study is both an exercise in the use of Monte Carlo methods and an analysis of the very popular Ising model.
+Monte Carlo methods are applicable in several situations where many other methods are too computationally heavy.
+They are a broad class of methods that rely on random sampling to obtain numerical results.
+The Ising model is a mathematical model of ferromagnetism in statistical mechanics.
+The Ising model and Monte Carlo methods go hand in hand as we shall see in this study.
+
+I will start by giving a very brief description and definition of the two-dimensional Ising model.
+This general model will be simplified by assuming zero extern field which leads to a simple energy equation.
+Then I will present the Metropolis-Hastings algorithm and how it is implemented using Monte Carlo methods.
+Lastly, several simulations are presented, and I will try to uncover the Curie temperature where the ferromagnetic model transitions into a paramagnetic one.
 
@@
 
+Bear in mind that $\beta = \frac{1}{k_BT}$, and that all the quantities computed are functions of $T$. The results computed here can be used as comparison for numerical computations.
+
 \section{Algorithm}
+An initial spin configuration will be
 
-\section{Results}
+\subsection{Markov chains}
+A Markov chain gives a good description for a system that moves towards a steady state given an initial configuration.
+A Markov process is a random walk with a a set of probabilities for making a set of moves.
+Such a process is usually represented by a stochastic matrix containing probabilities of transitioning from one state to another.
+In physics Markov chains can by used to study diffusion because it gives a framework for the rules of Brownian motion - the behaviour exhibited by small parts of any system when exposed to random fluctuations of the medium.
+
+In order to simulate how the system in this study evolves towards a steady state, Markov chains are used repeatedly in Monte Carlo simulations.
+From Markov chains one gets tow conditions needed for reaching a steady state, detailed balance and ergodicity.
+These properties governs our choice of algorithm, the Metropolis-Hastings algorithm.
+
+\subsection{The Metropolis-Hastings algorithm}
+In this project we need random samples from a probability distribution $P_\beta (\sigma)$ associated with the physical system we wish to model.
+Regrettably, direct sampling can be incredibly difficult, as we need to compute the partition function in order to express the probability distribution in its entirety.
+The Metropolis-Hastings algorithm is circumvents this problem because it only requires a function $f$ proportional to the distribution density.
+
+The Metropolis-Hastings algorithm can be condensed down to the following steps.
+\begin{enumerate}
+\item The system is initialised by an initial state which can be randomly generated. The energy $E_b$ of this configuration is computed.
+\item The initial configuration is changed by flipping the spin of an arbitrary site. One then computes the energy of this new trial state $E_t$.
+\item The difference in energy between the two states $\Delta E = E_t - E_b$ is calculated. For the two-dimensional Ising model there are only five different possilbe values for $\Delta E$.
+\item If $ \Delta E < 0$ the new configuration is accepted, meaning that the energy is lower than it was, and the system is progressing towards an energy minimum.
+\item If $ \Delta E > 0$ one computes $w = e^(\beta \Delta E)$ and compares $w$ to a random number $r$. If $r < w$ the state is accepted, else the inital state is kept.
+\item The expectation values are updated. And the steps are repeated until a sufficiently good steady state is reached.
+\end{enumerate}
+
+As stated above, the possible different energy changes for a two-dimensional Ising model are finite.
+The possible changes are $\Delta E = 8J, 4J, 0, -4J, -8J$.
+An array containing these values is constructed which contains the possible values before doing the Metropolis sampling.
+One can compare the possible energy changes against elements of this array instead of computing $w$ for every other iteration.
+This should vastly increase computing speed.
+
+The algorithm determines whether a proposed move is implemented based on a transition probability and an acceptance probability.
+The strength of the algorithm is that the transition probability need not be known.
+
+\section{Implementation}
+The Ising model is implemented in C++ in the class \lstinline|Ising.cpp| contains all the methods necessary for a simulation of the model.
+The full program with can be found by following the link at the front page of this article.
+
+\subsection{Ising class}
+An object-oriented approach was picked in order to increase flexibility and tidiness.
+The only data structure that exists outside the Ising class is the array of expected values, for easy comparison between several class instances.
+As the program is built right now, it can be run with command line arguments for spin lattice size, temperature domain and number of Monte Carlo cycles, which called $M$ times by \lstinline|simulate| adds up to a total complexity of $\mathcal{O}(ML^2)$.
+
+Of the various methods incorporated in the \lstinline|Ising| class, \lstinline|simulate| and \lstinline|metropolis| are the most important ones.
+The method \lstinline|simulate| calls the \lstinline|metropolis| for every Monte Carlo cycle.
+Denoting $L$ as the spin lattice dimension and $M$ the number of Monte Carlo cycles, each call of \lstinline|metropolis| constitutes a complexity of $\mathcal{O}(L^2)$.
+
+\section{Simulations}
+
+\subsection{$2\times2$ Ising model with $T=1$}
+Figure \ref{tab:2by2analMC} shows how increasing number of Monte Carlo cycles measures up to the analytic (exact) solution for a $2\times2$ system.
+One can see that it is necessary to have towards $1000000$ cycles in order to get the best estimates.
+This is an important results when making further computations.
+Moreover, a diagram of the magnetisation and energy of this system is shown in figure
+
+\begin{table}
+	\centering
+	\caption{Comparison between Monte Carlo simulations at different length and the analytical solution for a $2\times2$ system.} \label{tab:2by2analMC}
+	\begin{tabular}{lccccc} \hline
+	$M$ & $\ev{E}$ & $C_V$ & $\chi$ & $\ev{\mathcal{M}}$ \\ \hline
+	100         &  $-2$ & $0$ & $0$ & $1$ \\
+	1000       & $-1.998$ & $0.015984$ & $0.03996$ & $0.6235$ \\
+	10000     & $-1.9964$ & $0.028748$ & $0.003594$ & $0.44705$ \\
+	100000   & $-1.99626$ & $0.029864$ & $0.003694$ & $0.040625$ \\
+	1000000 & $-1.995968$ & $0.032191$ & $0.0039439$ & $0.059681$ \\ \hline
+	Analytical & $ -1.99598 $ & $0.03208$ & $0.00401$ &  $0$ \\ \hline
+	\end{tabular}
+\end{table}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.9\textwidth]{../figures/EandM_2_T1_bothorient.pdf}
+	\caption{Energy and Magnetisation as a function of Monte Carlo cycles for the $2\times2$ system. \label{fig:2by2simple}}
+\end{figure}
+
+\subsection{$20\times20$ Ising model}
+
+We are interested in figuring out how many Monte Carlo cycles are necessary in order to establish an steady state.
+This must be done in order to know where it is alright to start calculation expected values.
+If we start the computations before a steady state is reached in the system, the results can be skewed.
+How the system approaches a steady state can already be seen in figure \ref{fig:2by2simple}.
+In this figure though, a mere hundred Monte Carlo cycles were conducted, and the system is indeed incredibly small.
+Therefore, we consider both an initial random configuration for a larger $20 \times 20$ spin system, both for temperatures $T=1$ and $T=2.4$.
+The result of this simulation can be found in figure \ref{fig:temp1and2.4}.
+This simulations has been allowed to run for a one million Monte Carlo cycles.
+One can clearly see that both systems starts at some configuration which is some distance away from the equilibrium and then quickly falls (rises) asymptotically towards a steady state.
+Interestingly, for the higher temperature, magnetisation behaves differently than for the lower temperature.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.9\textwidth]{../figures/20EandMT1and24.pdf}
+	\caption{Mean Energy and Mean magnetisation for a $20 \times 20$ spin lattice with temperatures $T=1$ and $T=2.4$. \label{fig:temp1and2.4}}
+\end{figure}
+
+It would also be interesting to see how many spin changes of the Metropolis-Hastings algorithm are accepted at this higher temperature.
+A plot for accepted configuration changes as a function of cycles for the two temperatures $T=1$ and $T=2.4$ is shown in figure \ref{fig:acceptedstates}.
+One can clearly see that for the higher-temperature system the number of accepted states linearly proportional to the number of cycles in the simulation, while for the lower-temperature system the number of accepted states stay the same after the steady state is reached.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.9\textwidth]{../figures/20NaccofNT1and24.pdf}
+	\caption{Figure showing that the number of accepted states for a higher-temperature system ($T=2.4$) rises linearly, while the accepted states for for a lower-temperature system ($T=1$) stays the constant at equilibrium.\label{fig:acceptedstates}}
+\end{figure}
+
+The probabilities for finding the system in a given configuration will also change for a higher temperature.
+A simulation of the probability density functions of the $20\times20$ system for temperatures $T=1$ and $T=2.4$ is given in figure \ref{fig:pde}.
+One can see that the probability for finding the system at a higher energy is moved larger at a higher temperature, as one would expect.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.9\textwidth]{../figures/EhistProbBothT.pdf}
+	\caption{Probability densities for a lower temperature $T=1$ and a higher temperature $T=2.4$. \label{fig:pde}}
+\end{figure}
+
+\subsection{Larger spin lattices}
+
+Figure \ref{fig:largersystems} depicts the expected energy $\ev{E}$, average absolute magnetisation $\ev{\abs{\mathcal{M}}}$, specific heat $C_V$ and susceptibility $\chi$ as functions of temperature $T \in [2.0,2.4]$ for large spin lattices $L=20$, $L=40$, $L=60$ and $L=80$.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=\textwidth]{../figures/largersystems.pdf}
+	\caption{The mean energy, magnetisation, specific heat and susceptibility for larger spin lattices \label{fig:largersystems}}
+\end{figure}
 
 \section{Discussion}
+I have reason to believe that the graphs in figure \ref{fig:largersystems} are wrong, especially the specific heat and the susceptibility.
+In theory, as the size of the spin lattice approaches a very large number, one should find the Curie $T_c$ temperature, according to the following equation.
+\begin{equation}
+\label{eq:curie}
+T_c(L\to\infty) = T_c(L)-aL^{-\frac{1}{\nu}}
+\end{equation}
+where the exact result gives $\nu = 1$ and $a$ is a constant.
+
+After long hours, working through the night, in front of the computer I have regrettably been unable to uncover the problem in my program.
+According to Onsager \cite{onsager} the critical temperature is $kT_C/J = 2/\ln(1+\sqrt{2}) \approx 2.269$ with $\nu = 1$.
+This temperature would be should be clear in all the plots in figure \ref{fig:largersystems}.
+First, the energy plot show be splitting slightly at $T_c$.
+Second, the magnetisation should be falling more sharply at $T_c$, with a sharper bend for larger $L$.
+Third, the specific heat plot should sloping upwards before reaching a maximum at $T_c$ and then sloping downwards.
+Fourth, the susceptibility plot would show traits similar to the plot for specific heat, with a clear bump at $T_c$.
+
+The plot for expected energy $\ev{E}$ and $\ev{\abs{\mathcal{M}}}$ clearly exhibits some of the these described traits, while the plot for susceptibility $\chi$ is quite close to what one would expect, but not entirely spot on.
+The specific heat $C_V$ on the other hand is as horrible as it could be.
+
+I have conducted computations for spin lattices up to $L=140$ and one million Monte Carlo cycles during this study\footnote{The computations were done on the 64 core computing cluster at the Institute of Astophysics.
+Thanks to Trygve L.
+Svalheim for giving me access.}, but as I improved upon the code during the night leading up to deadline for this project I have chosen not to include the results herein as they are less accurate than the ones presented in figure \ref{fig:largersystems}.
+If interested, data files and a figure can be found by following the URL to github on the front page of this document.
 
 \section{Conclusion}
+In this project I employed the Ising model and the Metropolis algorithm to model a phase transition from ferromagnetic to paramagnetic material.
+I tried to obtain the Curie temperature, $T_c$ at which such a phase transition happens, but was only partly successful in this endeavour.
+Despite this I feel that I have demonstrated to a great extent how the Ising model functions and the benefit of Monto Carlo methods.
 

### 2016-11-22 Project4/article/article.tex
P5: Added file write function. P4: minor changes to article

@@
 \section{Algorithm}
-An initial spin configuration will be
+The algorithm for solving the Ising model consists of several parts and interesting aspects.
+The path to a steady state for the Ising mode can be seen as a Markov chain, which justifies the use of the Metropolis-Hastings algorithm, which again can be seen as one cycle in a Monte Carlo solver.
 
@@
 In order to simulate how the system in this study evolves towards a steady state, Markov chains are used repeatedly in Monte Carlo simulations.
-From Markov chains one gets tow conditions needed for reaching a steady state, detailed balance and ergodicity.
+From Markov chains one gets two conditions needed for reaching a steady state, detailed balance and ergodicity.
 These properties governs our choice of algorithm, the Metropolis-Hastings algorithm.

### 2016-11-22 Project5/scripts/plothisttest.py

@@ -1,0 +1,7 @@
+from matplotlib import pyplot as plt 
+import numpy as np 
+
+data = np.loadtxt("../data/stuff2.dat")
+
+plt.hist(data, bins = 100)
+plt.show()

### 2016-11-22 Project5/src/functions.cpp

@@ -2,22 +2,45 @@
 #include <time.h>
 #include <iostream>
 #include <armadillo>
+#include <string.h>
+#include <iomanip>
 
-void transact(arma::vec (&agents), int no_of_transactions, int no_of_agents) {
+void transact(arma::vec (&agents), int no_of_transactions, int no_of_agents, double savings_rate) {
+
+    double lambda = savings_rate;
 
     for (int i = 0; i < no_of_transactions; i++) {
+
         // Picking random dudes
         int agent1_rank = (int) rand() % no_of_agents;
         int agent2_rank = (int) rand() % no_of_agents;
         // They should not be the same dude
         while (agent1_rank == agent2_rank) agent2_rank = (int) rand() % no_of_agents;
+
         double epsilon = (double) rand() / RAND_MAX;
-        double new_agent1 = epsilon * (agents(agent1_rank) + agents(agent2_rank));
-        double new_agent2 = (1 - epsilon) * (agents(agent1_rank) + agents(agent2_rank));
+        double new_agent1 = lambda * agents(agent1_rank)
+                + epsilon * (1 - lambda) * (agents(agent1_rank) + agents(agent2_rank));
+        double new_agent2 = lambda * agents(agent1_rank) +
+                (1 - epsilon) * (1 - lambda) * (agents(agent1_rank) + agents(agent2_rank));
         agents(agent1_rank) = new_agent1;
         agents(agent2_rank) = new_agent2;
     }
 
     // Re-seeding random number generator
     srand(time(NULL));
+}
+
+void write_to_file(arma::vec data, int vector_length, std::string filename) {
+
+    using namespace std;
+
+    ofstream ofile;
+    ofile.open(filename, ios::app);
+    ofile << setiosflags(ios::showpoint | ios::uppercase);
+    for (int i = 0; i < vector_length; i++) {
+
+        ofile << setw(15) << setprecision(8) << data[i] << endl;
+    }
+    ofile.close();
+
 }

### 2016-11-22 Project5/src/functions.h

@@ -1,4 +1,6 @@
 #include <armadillo>
 
 
-void transact(arma::vec &agents, int no_of_transactions, int no_of_agents);
+void transact(arma::vec &agents, int no_of_transactions, int no_of_agents, double savings_rate);
+
+void write_to_file(arma::vec data, int vector_length, std::string filename);

### 2016-11-22 Project5/src/main.cpp

@@ -1,26 +1,36 @@
 #include <iostream>
 #include "functions.h"
-# include <time.h>
+#include <time.h>
+#include <string.h>
 
 using namespace std;
 
 
 int main(int argc, char *argv[]) {
 
-    int N = 10; // No of agents
+    int N = 1000; // No of agents
     int no_of_transactions = 1e7;
     double initial_wealth = 100;
+    double savings_rate = 0.25;
+    string filename = "stuff.dat";
 
     arma::vec agents(N);
     agents.fill(initial_wealth);
 
-    cout << "No of agents = " << N << ". Sum = " << arma::sum(agents) << endl;
-    cout << "Initial wealth distribution: " << endl;
-    cout << agents << endl;
-    transact(agents, no_of_transactions, N);
-    cout << "Wealth distribution after " << no_of_transactions << " transactions"
-         << ". Sum = " << arma::sum(agents) << endl;
-    cout << agents << endl;
+    // Timer variables
+    clock_t start, end;
+
+    // Start timer
+    start = clock();
+
+    transact(agents, no_of_transactions, N, savings_rate);
+    write_to_file(agents, N, filename);
+
+    // Stop timer
+    end = clock();
+
+    cout << "Time elapsed: " << (double) (end - start) / CLOCKS_PER_SEC
+         << " seconds." << endl;
 
     return 0;
 }

### 2016-12-07 Project5/article/article.tex
Started theory section

\documentclass[10pt, a4paper]{amsart}

\usepackage[]{graphicx}
\usepackage[]{hyperref}
\usepackage[]{physics}
\usepackage[]{listings}
\usepackage[T1]{fontenc}

\title[Simulation of monetary transactions]{Monte Carlo simulation of monetary transactions: \\
\normalsize{A simple model for wealth distribution} \\
  \hrulefill\small{ FYS3150: Computational Physics }\hrulefill}

\author[Svalheim \& Winther-Larsen]{Trygve Leithe Svalheim \\
   Sebastian G. Winther-Larsen \\
  \href{https://github.com/gregwinther/FYS3150/}{\texttt{github.com/gregwinther}}}

\begin{document}

\begin{titlepage}
\begin{abstract}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. In id neque elementum, accumsan ligula at, lobortis tortor. Duis elementum pellentesque purus, sit amet euismod diam facilisis volutpat. Nulla facilisi. Mauris quis felis ante. Aliquam ac velit sit amet velit porta condimentum iaculis ut quam. Phasellus pretium libero nec lectus placerat, ac consectetur sem faucibus. Maecenas dictum porta finibus.
\end{abstract}
\maketitle
\tableofcontents
\end{titlepage}

%% -----------------
%% INTRODUCTION
\section{Introduction}

%% -----------------
%% THEORY
\section{Theoretical Background}
\subsection{The simplest model for an economy}
Arguably, the most famous equation in macroeconomics is the ``autarky identity"
\begin{equation}
\label{eq:YCGI}
Y = C + G + I,
\end{equation}
where $Y$ is income, $C$ is consumption, $G$ is government spending and $I$ is investment. The best-know, but not necessarily the best, measure of income $Y$ is the Gross Domestic Product (GDP)\footnote{GDP = GNP (Gross National Product) in autarky.}. Consumption $C$ is the monetary value of all goods and services purchased in the private sector, while government spending $G$ is the consumption of the government. Finally, investment $I$ is the sum of private and public saving.

Equation \ref{eq:YCGI} is autarkic because all interactions with other economies are excluded from the expression. There are no terms representing exports and captial inflow, for instance. We are dealing with a \emph{closed} economy, alternatively the entire world as a whole. Moreover, let's assume that the economy we are studying is a peaceful anarchy, without a governing authority of any sort, in effect setting $G = 0$. To begin with, we will also forgo the agents the ability to save such that the worth of every individual, or agent, in the economy must be spent at once. Equation \ref{eq:YCGI} is reduced to $Y = C$, everything one agent spends is the income of another.

\subsection{Monetary transactions}
To simulate monetary transactions in out model economy we expand employ the framework introduced in Patriarca et al.\cite{Patriarca}. We assume there are $N$ agents that exchange money in pairs ($i,j$). We assume also that all agents start with the same amount of money $m_0>0$. For every period an arbitrary pair of agents are picked at randomly and let them conduct business, id est a transaction takes place between them. Money is conserved during the transaction such that
\begin{equation}
m_i + m_j = m_i' + m_j',
\end{equation}
where the right-hand side represents the amount of money agents $i$ and $j$ have after the transaction. The exchange is done via a random reassignment factor $\epsilon$ such that
\begin{align}
m_i' &= \epsilon (m_i + m_j) \\
m_j' &= (1 - \epsilon) (m_i + m_j)
\end{align}
Th

%% -----------------
%% ALGORITHM
\section{Algorithm}

%% -----------------
%% RESULTS
\section{Results}

%% ----------------
%% DISCUSSION
\section{Discussion}

%% ---------------
%% CONCLUSION
\section{Summary Remarks}

\begin{thebibliography}{10}

\bibitem{Patriarca} Patriarca, M., Chakraborti, A., \& Kaski, K. (2004). 
	Gibbs versus non-Gibbs distributions in money dynamics. 
	\emph{Physica A: Statistical Mechanics and its Applications},
	340(1), 334-339.

\end{thebibliography}

\end{document}

### 2016-12-07 Project5/src/functions.cpp
Some tests with different params

@@ -53,7 +53,7 @@ void trade(int N, int no_of_transactions, arma::vec (&agents), double lambda, do
 
             // Pick new agents ...
             agent_i = (int) rand() % N;
-            //agent_j = (int) rand() % N;
+            agent_j = (int) rand() % N;
 
             // ... find wealth of these ...
             m_i = agents(agent_i);

### 2016-12-07 Project5/src/main.cpp

@@ -14,7 +14,7 @@ int main() {
     int transactions    =  1e7;         // No. of transactions
     int simulations     =  1e3;         // No. of simulations
     double lambda       =  0.5;         // Propensity to save
-    double alpha        =  0.0;         // Wealth proximity variable
+    double alpha        =  2.0;         // Wealth proximity variable
     double gamma        =  0.0;         // Past interaction variable
 
     arma::vec agents(N);    // Agent vector

### 2016-12-07 Project5/article/article.tex
A bit of savings

@@
 \begin{equation}
+\label{eq:wealthconserved}
 m_i + m_j = m_i' + m_j',
 \end{equation}
-where the right-hand side represents the amount of money agents $i$ and $j$ have after the transaction. The exchange is done via a random reassignment factor $\epsilon$ such that
+where the left hand side is the agent $i$ and $j$'s updated wealth and the right-hand side represents the amount of money agents $i$ and $j$ had before the transaction.
+The exchange is done via a random reassignment factor $\epsilon$, such that
 \begin{align}
@@
 \end{align}
-Th
+No agent will ever have negative wealth, that is $m \geq 0$.
+Moreover, because of the conservation law in equation \ref{eq:wealthconserved}, the system eventually reaches an equilibrium state given by a Gibbs distribution
+\begin{equation}
+\label{eq:gibbs}
+w_m = \beta e^{-\beta m}, \quad \beta = \frac{1}{\ev{m}},
+\end{equation}
+where $\ev{m}$ is the expected wealth, for which the arithmetic mean is an unbiased estimator.
+This implies that after an equilibrium has been reached the majority of agents is left with lower wealth than they had initially and
+the number of rich agents exponentially decrease.
+
+\subsection{Transactions and savings}
+We are can now expand upon the model by introducing a savings rate $\lambda$.
+The savings rate is defined as a fraction of an agent's wealth that does not partake in a transaction for every period.
+One can gather from the macroeconomic identity in equation \ref{eq:YCGI} that income must still be the same and the transaction law in equation \ref{eq:wealthconserved} still holds.
+Now one
 

### 2016-12-08 Project5/article/article.tex
Theory done

@@
 \usepackage[]{listings}
-\usepackage[T1]{fontenc}
+\usepackage[utf8]{inputenc}
 
@@
 \subsection{The simplest model for an economy}
-Arguably, the most famous equation in macroeconomics is the ``autarky identity"
+Arguably, the most famous equation in macroeconomics is the ``autarky identity"\footnote{Any introductory text on macroeconomics will give a thorough elaboration, e.g. Gärtner.}
 \begin{equation}
@@
 \end{equation}
-where the left hand side is the agent $i$ and $j$'s updated wealth and the right-hand side represents the amount of money agents $i$ and $j$ had before the transaction.
+where the right-hand side is the agent $i$ and $j$'s updated wealth and the left-hand side represents the amount of money agents $i$ and $j$ had before the transaction.
 The exchange is done via a random reassignment factor $\epsilon$, such that
@@
 One can gather from the macroeconomic identity in equation \ref{eq:YCGI} that income must still be the same and the transaction law in equation \ref{eq:wealthconserved} still holds.
-Now one
+The updated wealth of agents $i$ and $j$ after a transaction becomes
+\begin{align}
+m_i' &= \lambda m_i + \epsilon (1 - \lambda) (m_i + m_j) \\
+m_j' &= \lambda m_j + (1- \epsilon) (1 - \lambda) (m_i + m_j)
+\end{align}
+one can rewrite these expressions to
+\begin{align*}
+m_i' &= m_i + \delta m \\
+m_j' &= m_j - \delta m
+\end{align*}
+where
+\begin{equation}
+\delta m = (1 - \lambda)(\epsilon m_j - (1 - \epsilon)m_i)
+\end{equation}
+
+\subsection{Economic inequality and social friction}
+``The first historical series of income distribution statistics became available with the publication in 1953 fo Kuznet's monumental Shares of Upper Income Groups in Income and Savings.
+Kuznet's series dealt with only one country (the United Stetes) over a period of thirty-five years (1913-1948)" (Pikkety \& Ganser, 2014, Kindle locations 276-280\cite{Piketty}).
+Ever since that time there has been wealth and income inequality to a greater or lesser degree.
+According to the PolitiFact the top 400 richest Americans in 2011 "[had] more wealth than half of all Americans combined"\cite{Moore}.
+
+One of the possible reasons for continued wealth and income inequality that has risen in popularity the last few years is the process of wealth concentration.
+This is a process by which, under certain conditions, newly created wealth concentrates in the possession of already-wealthy individuals or entities.
+Those who already have wealth have the means to invest in newly created sources and structures of wealth.
+Piketty argues that the fundamental force for wealth divergence is the usually greater return of capital than economic growth (Piketty \& Ganser, 2014 p.
+284 Table 12.2\cite{Piketty}).
+
+To incorporate the social rigidities described above, we will make two further additions to the model.
+These additions build mostly on the work of Goswami and Sen\cite{GoswamiSen}.
+Firstly, we will make it more likely for two agents to conduct a transaction if they have similar wealth.
+Second, we will make it more likely for two agents to make a transaction if they have made a transaction before.
+We define a probability
+\begin{equation}
+\label{eq:rigidprob}
+p_{ij} \propto \abs{m_i - m_j}^{-\alpha}(c_{ij}+1)^{\gamma},
+\end{equation}
+where the first factor $\abs{m_i-m_j}^{-\alpha}$ relates to wealth similarity and the second factor $(c_{ij}+1)^{\gamma}$ relates to previous transactions.
+A relatively large difference between $m_i$ and $m_j$ should translate to a lower probability.
+The strength of this effect is governed by the exponent $\alpha$, a larger value decreases the probability of a trade further.
+The variable $c_{ij}$ is simply the number of times agent $i$ and $j$ have conducted transactions in the past.
+The number $1$ is added in order to ensure that if they have not interacted earlier they can still interact.
+The strength of this effect is governed by the exponent $\gamma$.
 
@@
 
+\bibitem{Piketty} Piketty, T., \& Ganser, L. J. (2014).
+	\emph{Capital in the twenty-first century}.
+
+\bibitem{Moore} Moore, M., (March 7, 2011).
+	The Forbes 400 vs. everybody else.
+	\emph{\href{https://web.archive.org/web/20110309211959/http://www.michaelmoore.com/words/must-read/forbes-400-vs-everybody-else}{michaelmoore.com}}.
+	Retrieved December 8. 2016.
+
+\bibitem{GoswamiSen} Goswami, S., \& Sen, P. (2014).
+	Agent based models for wealth distribution with preference in interaction.
+	\emph{Physica A: Statistical Mechanics and its Applications},
+	415, 514-524.
+
 \end{thebibliography}

### 2016-12-08 Project5/article/article.tex
Ploughing through results section

@@
 \usepackage[]{graphicx}
+\usepackage{subcaption}
 \usepackage[]{hyperref}
@@
 \usepackage[utf8]{inputenc}
+\usepackage[dvipsnames]{xcolor}
+
+\definecolor{mygray}{gray}{0.9}
+
+\lstset{
+	frame = single,
+	language = C++,
+	showstringspaces = false,
+	tabsize = 2,
+	otherkeywords = {self},
+	keywordstyle = \color{Maroon},
+	identifierstyle=\color{olive},
+ 	stringstyle=\color{orange},
+ 	backgroundcolor=\color{mygray},
+ 	breaklines = true
+}
 
@@
 \subsection{Monetary transactions}
-To simulate monetary transactions in out model economy we expand employ the framework introduced in Patriarca et al.\cite{Patriarca}.
+To simulate monetary transactions in our model economy we expand employ the framework introduced in Patriarca et al.\cite{Patriarca}.
 We assume there are $N$ agents that exchange money in pairs ($i,j$).
@@
 
+It is easy to see that the logarithm of the Gibbs measure in equation \ref{eq:gibbs} yields a linear equation
+\begin{equation}
+\label{eq:lingibbs}
+\ln(w_m)=\ln(\beta) - \beta m
+\end{equation}
+
+\subsection{Pareto distributions}
+
 \subsection{Transactions and savings}
-We are can now expand upon the model by introducing a savings rate $\lambda$.
+We are can expand upon the model by introducing a savings rate $\lambda$.
 The savings rate is defined as a fraction of an agent's wealth that does not partake in a transaction for every period.
@@
 %% ALGORITHM
-\section{Algorithm}
+\section{Algorithm \& Implementation}
+All programming code used in this project is written in C++ and is publicly available at github by following the link on the first page.
+The program consists mainly of two files, \lstinline|main.cpp| and \lstinline|functions.cpp|.
+\lstinline|main.cpp| contains the set-up of the program and is where the number of agents, initial wealth, number of transactions, number of simulations and all parameters ($\lambda$, $\alpha$ and $\gamma$) are specified.
+The main part of the program however, is in \lstinline|functions.cpp| which in addition to the \lstinline|trade(...)|\footnote{The functions have several arguments: \lstinline|trade(int N, int no_of_transactions, arma::vec (&agents), double lambda, double alpha, double gamma)|.
+To include all the arguments in the text would have decreased readability and they are replaced with dots.} function contains the function \lstinline|output(...)|\footnote{\lstinline|output(...)| = \lstinline|output(int N, arma::vec agents, std::string filename)|.} which writes data to file.
+
+\subsection{The \lstinline|trade(...)| function}
+This function runs a specified number of Monte Carlo cycles, where the transactions take place.
+For each iteration two random agents are picked, then some logic follows which will determine if the two agents are allowed to go into business.
+This is a very important part of the program and is therefore listed here:
+\begin{lstlisting}
+while ((pow(fabs(m_i - m_j), -alpha)*
+		pow(previous_transactions+1, gamma) < random_number)
+		|| (agent_i == agent_j)) {
+
+	// Pick new agents ...
+	agent_i = (int) rand() % N;
+	agent_j = (int) rand() % N;
+
+	// ... find wealth of these ...
+	m_i = agents(agent_i);
+	m_j = agents(agent_j);
+
+	// ... update previous transactions ...
+	previous_transactions = C(agent_i, agent_j);
+
+	// ... and update the random comparison number
+	random_number = (double) rand() / RAND_MAX;
+\end{lstlisting}
+One can see that the the argument of the while loop in the program listing above contains equation \ref{eq:rigidprob} which is compared against a random number and an equality test that will return \lstinline|True| if agents $i$ and $j$ are the same agent.
+This part of the code makes sure that a new pair of randomly picked agents until we are satisfied that the agents are not the same and that they are close enough in wealth and in previous relations to go forward with a transaction.
+After everything is all-right and we have a good pair of agents a transaction takes place:
+\begin{lstlisting}
+// Random value of transaction b/w 0 and 1
+double epsilon = (double) rand() / RAND_MAX;
+
+// Transaction takes place, but some is saved (lambda)
+double delta_m = (1 - lambda) * 
+	(epsilon*m_j - (1 - epsilon)*m_i);
+agents(agent_i) += delta_m;
+agents(agent_j) -= delta_m;
+
+// Register that a transaction has taken place
+C(agent_i, agent_j) += 1;
+C(agent_j, agent_i) += 1;
+\end{lstlisting}
+The program listing above should be straight-forward. Notice that the transaction is registered in matrix \lstinline|C| by incrementing element $i,j$ and $j,i$ by $1$.
+
+The last part of the \lstinline|trade(...)| function is not particularly important for the quality of the data that a simulation will provide, but vastly increased the speed of the program.
+This last functionality breaks the for loop within the \lstinline|trade(...)| function when a steady state is reached.
+Listed in its entirety here:
+\begin{lstlisting}
+double var = arma::var(agents);
+cumVarBlock += var;
+cumVar2Block += var*var;
+
+// Enter here at the correct place: end of block
+if (i % blockSize == 0) {
+	// Variance of a block "averaged" over the block
+	double avgVar  =  cumVarBlock / blockSize;
+	double avgVar2 = cumVar2Block / blockSize;
+
+	// Variance of a block of variance
+	double varVarBlock = avgVar2 - avgVar*avgVar;
+
+	// Check if variance is low enough
+	if ((fabs(avgVarBlockOld - avgVar) / fabs(avgVarBlockOld)     < 0.2) &&
+		(fabs(varVarBlock - varVarBlockOld) / fabs(varVarBlockOld) < 0.5)) {
+
+		// Telling a user that an equilibrium has been reached
+		std::cout << "Equilibrium reached at transaction no. " << i << std::endl;
+
+		// Done! For now.
+		break;
+
+		} else {
+			avgVarBlockOld = avgVar;
+			varVarBlockOld = varVarBlock;
+		}
+	// Reset cumulative variance
+	cumVarBlock  = 0;
+	cumVar2Block = 0;
+}
+\end{lstlisting}
+This code section does some calculations on the results after a certain amount of transactions have taken place, an amount predetermined by \lstinline|blockSize|.
+The average variance of the block is computed as well as the average square variance of the block size.
+Then the "variance of the variance" is computed.
+These measures are compared against the measures from the previous block.
+The relative size of the average variance must be less than $20\%$ and the relative size of the variance of the variance must be less than $50\%$ of the measures from the previous block.
+If both these logical statements evaluate to \lstinline|True|, the for loop of the \lstinline|trade(...)| function is broken, terminating the current simulation.
+We found that for most simulations an equilibrium is reached quite quickly, usually between $40000$ and $120000$ transactions have taken place between $N=1000$ agents.
+This speeds up the time for a simulation to finish tremendously, compared with the initially, and rather naively, picked $10^7$ iterations.
+In order to find correct values \lstinline|blockSize| as well as when the variance measures are ``low enough'' the behaviour of the variance measures were meticulously studied through several trial runs of the function.
+
+\subsection{Example \lstinline|main()|}
+The parameters of the simulation is set up in the \lstinline|main()| function. When all parameters are chosen the following for loop is almost everything that is needed
+\begin{lstlisting}
+// Start simulations
+for (int i = 0; i < simulations; i++) {
+
+	// Print progress
+	std::cout << "Simulation no. " << i << std::endl;
+
+	// Assign intitial wealth to all agents
+	agents.fill(m0);
+
+	// Trade!
+	trade(N, transactions, agents, lambda, alpha, gamma);
+
+	// Sort and add equilibrium wealth to total
+	totagents += arma::sort(agents);
+\end{lstlisting}
+One can see that what must be specified is the number of simulations (\lstinline|simulations|), initial wealth (\lstinline|m0|), number of agents (\lstinline|N|), number of transactions (\lstinline|transactions|), an array containing the wealth of every agent (\lstinline|agents|), savings rate (\lstinline|lambda|), parameter governing weight of wealth similarity (\lstinline|alpha|) and parameter governing weight of past transactions (\lstinline|gamma|).
 
@@
 \section{Results}
+
+Figure \ref{fig:distribution} shows plot of different wealth distributions after $1000$ simulations for $N=500$ agents.
+We see that that without savings, the distribution shows a decreasing number of agents as wealth increases.
+As the savings rate $\lambda$ is allowed to increase, the distribution moves to the right and eventually peaks at the initial wealth $m_0 = 100$ when $\lambda = 0.9$.
+This plot, were it not for the colors, could be an exact copy of figure 1 in Patriarca et al.\cite{Patriarca}.
+Without savings the distribution appears as a inverse exponential function ($e^{-x}$) which is the same class of function as the Gibbs distribution (equation \ref{eq:gibbs}).
+However, these traits disappear from the wealth distributions once savings are allowed.
+
+A further study of the wealth distributions from figure \ref{fig:distribution} is visualized in figure \ref{fig:logplot}.
+Figure  \ref{fig:logdistribution} shows the same graphs as in figure \ref{fig:distribution}, but on with logarithmic scales on both axes.
+Figure \ref{fig:straightline} the wealth values evaluated by equation \ref{eq:gibbs} and plotted with logarithmic $y$-axis.
+Observe that all all the wealth distributions give straight lines.
+This makes sense, as the logarithm of the Gibbs distribution yields a first order expression (equation \ref{eq:lingibbs}).
+
+Figure \ref{fig:paramfig1} shows two different Pareto distrobutions plotted against a distribution at $\lambda = 0$.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.9\textwidth]{../figures/5ac/5c_N500_varSavings.pdf}
+	\caption{Wealth distribution for $N=500$ agents after $1000$ simulations at different savings rates $\lambda$}
+	\label{fig:distribution}
+\end{figure}
+
+\begin{figure}
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=\textwidth]{../figures/5ac/5cLOGLOG_N500_varSavings.pdf}
+  \caption{}
+  \label{fig:logdistribution}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=\linewidth]{../figures/5ac/omega.pdf}
+  \caption{}
+  \label{fig:straightline}
+\end{subfigure}
+\caption{Logarithmic plot wealth distribution from figure \ref{fig:distribution} (right), and }
+\label{fig:logplot}
+\end{figure}
+
+\begin{figure}
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=\linewidth]{../figures/5ac/5c_parameterization0.pdf}
+  \caption{}
+  \label{fig:paramfig1}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=\linewidth]{../figures/5ac/5c_parameterization09.pdf}
+  \caption{}
+  \label{fig:paramfig2}
+\end{subfigure}
+\caption{Parametrization}
+\label{fig:paramfig}
+\end{figure}
 
@@
 
-\bibitem{Piketty} Piketty, T., \& Ganser, L. J. (2014).
-	\emph{Capital in the twenty-first century}.
+\bibitem{Pareto} Pareto, V. (1897).
+	\emph{Cours d'Économique politique},
+	Lausanne: Ed. Rouge.
+
+\bibitem{Piketty} Piketty, T., \& Goldhammer, A. (2014).
+	\emph{Capital in the twenty-first century},
+	Cambridge: Harvard University Press.
 

### 2016-12-08 Project5/scripts/plothist.py

@@ -91,7 +91,8 @@ m = np.linspace(min(m1),max(m1),N1)
 lamb = 0.
 n = 1+ (3*lamb)/(1-lamb)
 x = m*n/np.mean(m)
-P = x**(n-1)*np.exp(-x)/(gamma(n))
+P = n**n*x**(n-1)*np.exp(-n*x)/(gamma(n))
+#P = x**(n-1)*np.exp(-x)/(gamma(n))
 plt.loglog(m,m**-3.,label=r'Pareto $\alpha = 2.0$',color=tableau20[4])
 plt.loglog(m,m**-2.8,label=r'Pareto $\alpha = 1.8$',color=tableau20[2])
 plt.loglog(x,P,color=tableau20[0], label='$\lambda = 0$ P(x)')

### 2016-12-08 Project5/src/functions.cpp

@@ -44,7 +44,7 @@ void trade(int N, int no_of_transactions, arma::vec (&agents), double lambda, do
         /*
          * The logical statement in the following while loops will check if the agents are
          * "good enough": Have they dealed in the past and are close in the "food chain".
-         * Furthermore: checks if agents are the same. Until a new good match of agents are
+         * Furthermore: checks if agents are the same. Until a good match of agents are
          * found, new random agents are picked.
          */
 

### 2016-12-08 Project5/src/main.cpp

@@ -29,7 +29,6 @@ int main() {
     start = clock();
 
     // Start simulations
-
     for (int i = 0; i < simulations; i++) {
 
         // Print progress

### 2016-12-09 Project5/article/article.tex
Results 2/3

@@
 \usepackage{subcaption}
+\usepackage[toc,page]{appendix}
 \usepackage[]{hyperref}
@@
 \subsection{Pareto distributions}
+How the income and wealth of a population has been studied for a long time.
+Already in 1897, Vilfredo Pareto, sought to describe bot wealth and income distribution by power law probability distributions\cite{Pareto}.
+Pareto showed that the higher end of the distribution of money follows a distribution
+\begin{equation}
+\label{eq:paretodist}
+w_m \propto m^{-1-\alpha}, \quad \alpha \in [1,2].
+\end{equation}
+Pareto's ``80-20 rule", that $20\%$ of the population controls $80\%$ of the wealth, corresponds to a certain value of $\alpha$.
+Today, a whole class of distributions similar to the one in equation \ref{eq:paretodist} are named after Pareto\cite{Arnold}.
 
@@
 
+%% Problem a, b, c.
+
 Figure \ref{fig:distribution} shows plot of different wealth distributions after $1000$ simulations for $N=500$ agents.
@@
 
-Figure \ref{fig:paramfig1} shows two different Pareto distrobutions plotted against a distribution at $\lambda = 0$.
+Figure \ref{fig:paramfig1} shows two different Pareto distributions with $\alpha=1.8$ and $\alpha = 2$ (equation \ref{eq:paretodist}) plotted against a distribution at $\lambda = 0$.
+We see that they fit adequately with the higher end of the distribution, but not perfectly.
+The simple Pareto distribution becomes a straight line because of the logarithmic axes in the plot.
+The same figure also contains a parametrization of the wealth \footnote{This is done in the way Partriarca et al.
+(2004)\cite{Patriarca} does.
+A recipe is shown in appendix \ref{app:partriarca}}.
+Figure \ref{fig:paramfig2} also shows such a parametrization, but for a higher savings rate $\lambda=0.9$.
+The parametrisations are indicated by a label $P(x)$.
 
@@
 \end{subfigure}
-\caption{Logarithmic plot wealth distribution from figure \ref{fig:distribution} (right), and }
+\caption{Logarithmic plot wealth distribution from figure \ref{fig:distribution} (right), and the logarithm of the Gibbs measure (equation \ref{eq:gibbs}) of every data point (right).}
 \label{fig:logplot}
@@
 \end{subfigure}
-\caption{Parametrization}
+\caption{Parametrization of two different wealth distributions, $\lambda=0$ and $\lambda=0.9$.
+Method for parametrization is found in appendix \ref{app:partriarca}.
+For $\lambda=0$ two Pareto power law distributions are also shown.}
 \label{fig:paramfig}
+\end{figure}
+
+%% Problem d.
+\begin{figure}
+\begin{subfigure}{.49\textwidth}
+	\centering
+	\includegraphics[width=\linewidth]{../figures/5d/5d_0-var-0.pdf}
+	\caption{}
+	\label{fig:neighbour1}
+\end{subfigure}
+\begin{subfigure}{.49\textwidth}
+	\centering
+	\includegraphics[width=\linewidth]{../figures/5d/5d_05-var-0.pdf}
+	\caption{}
+	\label{fig:neighbour2}
+\end{subfigure}
+\caption{Effect of neighbour-only interaction}
+\end{figure}
+
+\begin{figure}
+\begin{subfigure}{.49\textwidth}
+	\centering
+	\includegraphics[width=\linewidth]{../figures/5d/5d_0-var-0-PARETO.pdf}
+	\caption{}
+	\label{fig:neighbourpareto1}
+\end{subfigure}
+\begin{subfigure}{.49\textwidth}
+	\centering
+	\includegraphics[width=\linewidth]{../figures/5d/5d_0-var-0-POWERLOG.pdf}
+	\caption{}
+	\label{fig:neighourpareto2}
+\end{subfigure}
+\caption{Effect of neighbour-only interaction}
 \end{figure}
@@
 
+\bibitem{Arnold} Arnold, B. C. (1983). 
+	\emph{Pareto Distributions}.
+	 International Co-operative Publishing House
+
 \bibitem{Piketty} Piketty, T., \& Goldhammer, A. (2014).
@@
 
+\begin{appendices}
+
+\section{Parametrization from Patriarcal et al.}
+\label{app:partriarca}
+A corresponding exact solution of an income distributions for a generic value of $\lambda$ with $0<\lambda<1$ is provided in Patriarca et al.\cite{Patriarca}. Fir one employs the reduced variable
+\begin{equation}
+x = \frac{m}{\ev{m}}
+\end{equation}
+the agent money in units of average money $\ev{m}$ and the parameter
+\begin{equation}
+n(\lambda) = 1 + \frac{3\lambda}{1-\lambda}.
+\end{equation}
+The money distributions, for arbitrary values of $\lambda$, are well fitted by the function
+\begin{equation}
+P_n(x) = a_nx^{n-1}e^{-nx}
+\end{equation}
+where $a_n$ is a normalization factor shown to be
+\begin{equation}
+a_n = \frac{n^n}{\Gamma(n)}
+\end{equation}
+\end{appendices}
+
 \end{document}

### 2016-12-09 Project5/article/article.tex
Results done (for now)

@@
   \href{https://github.com/gregwinther/FYS3150/}{\texttt{github.com/gregwinther}}}
+
+\date{\today}
 
@@
 
-A further study of the wealth distributions from figure \ref{fig:distribution} is visualized in figure \ref{fig:logplot}.
-Figure  \ref{fig:logdistribution} shows the same graphs as in figure \ref{fig:distribution}, but on with logarithmic scales on both axes.
-Figure \ref{fig:straightline} the wealth values evaluated by equation \ref{eq:gibbs} and plotted with logarithmic $y$-axis.
-Observe that all all the wealth distributions give straight lines.
-This makes sense, as the logarithm of the Gibbs distribution yields a first order expression (equation \ref{eq:lingibbs}).
-
-Figure \ref{fig:paramfig1} shows two different Pareto distributions with $\alpha=1.8$ and $\alpha = 2$ (equation \ref{eq:paretodist}) plotted against a distribution at $\lambda = 0$.
-We see that they fit adequately with the higher end of the distribution, but not perfectly.
-The simple Pareto distribution becomes a straight line because of the logarithmic axes in the plot.
-The same figure also contains a parametrization of the wealth \footnote{This is done in the way Partriarca et al.
-(2004)\cite{Patriarca} does.
-A recipe is shown in appendix \ref{app:partriarca}}.
-Figure \ref{fig:paramfig2} also shows such a parametrization, but for a higher savings rate $\lambda=0.9$.
-The parametrisations are indicated by a label $P(x)$.
-
 \begin{figure}
@@
 \end{figure}
+
+A further study of the wealth distributions from figure \ref{fig:distribution} is visualized in figure \ref{fig:logplot}.
+Figure  \ref{fig:logdistribution} shows the same graphs as in figure \ref{fig:distribution}, but on with logarithmic scales on both axes.
+Figure \ref{fig:straightline} the wealth values evaluated by equation \ref{eq:gibbs} and plotted with logarithmic $y$-axis.
+Observe that all all the wealth distributions give straight lines.
+This makes sense, as the logarithm of the Gibbs distribution yields a first order expression (equation \ref{eq:lingibbs}).
 
@@
 \end{subfigure}
-\caption{Logarithmic plot wealth distribution from figure \ref{fig:distribution} (right), and the logarithm of the Gibbs measure (equation \ref{eq:gibbs}) of every data point (right).}
+\caption{Logarithmic plot of wealth distribution from figure \ref{fig:distribution} \ref{fig:logdistribution}), and the logarithm of the Gibbs measure (equation \ref{eq:gibbs}) of every data point (\ref{fig:straightline}).}
 \label{fig:logplot}
 \end{figure}
+
+Figure \ref{fig:paramfig1} shows two different Pareto distributions with $\alpha=1.8$ and $\alpha = 2$ (equation \ref{eq:paretodist}) plotted against a distribution at $\lambda = 0$.
+We see that they fit adequately with the higher end of the distribution, but not perfectly.
+The simple Pareto distribution becomes a straight line because of the logarithmic axes in the plot.
+The same figure also contains a parametrization of the wealth \footnote{This is done in the way Partriarca et al.
+(2004)\cite{Patriarca} does.
+A recipe is shown in appendix \ref{app:partriarca}}.
+Figure \ref{fig:paramfig2} also shows such a parametrization, but for a higher savings rate $\lambda=0.9$.
+The parametrisations are indicated by a label $P(x)$.
 
@@
 %% Problem d.
+
+The effect of ``nearest neighbhour'' interactions are summarized in figures \ref{fig:neighbour} and \ref{fig:neighbourpareto}.
+Figure \ref{fig:neighbour} shows logarithmic plots of simulations for different values of $\alpha$ from equation \ref{eq:rigidprob}.
+The parameter for previous transactions $gamma$ is set to zero for now.
+The only difference between the subfigures in figure \ref{fig:neighbour} is that figure \ref{fig:neighbour1} is for $\lambda=0$ and figure \ref{fig:neighbour2} is for $\lambda=0.5$, colloquially with and without saving respectively.
+Figure \ref{fig:neighbour} shows similar results to that of Goswarmi and Sen (2014)\cite{GoswamiSen}.
+
 \begin{figure}
@@
 \end{subfigure}
-\caption{Effect of neighbour-only interaction}
+\caption{The effect of different weights of propensity to trade with nearest neighbours, $\alpha$.
+Figure \ref{fig:neighbour1} is excludes saving ($\lambda = 0$), figure \ref{fig:neighbour2} includes saving ($\lambda = 0.5$).}
+\label{fig:neighbour}
 \end{figure}
+
+Figure \ref{fig:neighbourpareto} shows the simulations for nearest neighbour interaction fitted to a Pareto distribution.
+The Pareto distribution should fit only for higher values of the wealth distribution, id est the tails of the distribution.
+Figure \ref{fig:neighbourpareto1} show that the power law fittings (Pareto distributions) appear to have similar shapes to the results from the simulations.
+One can see, however, that the data from the simulations tend drop quicker than their Pareto power law counterparts.
+Figure \ref{fig:neighourpareto2} shows a logarithmic version of figure \ref{fig:neighbourpareto1}.
+Here the Pareto fittings are straight lines, as one would assume, but the data from the simulations apparently inhibit more complexity than the a simple power law relationship between wealth and frequency.
 
@@
 \end{subfigure}
-\caption{Effect of neighbour-only interaction}
+\caption{Power law fittings of neighbourhood effect wealth distributions. Figure \ref{fig:neighbourpareto1} is a linear plot, while figure \ref{fig:neighourpareto2} is double-logarithmic.}
+\label{fig:neighbourpareto}
+\end{figure}
+
+%% Problem e
+
+The last effect to consider is the former transaction effect, in combination with savings and nearest neighbour interactions.
+This effect is illustrated in figure \ref{fig:transbefore}.
+This figure is an attempt to replicate figure 5 in Goswami and Sen (2014)\cite{GoswamiSen}.
+Both subfigures show the wealth distribution for different values of $\gamma$ (propensity to trade with agents you have traded with before), but figure \ref{fig:transbefore1} weighs neighbour interactions less strongly than figure \ref{fig:transbefore2}, with $\alpha=1.0$ and $\alpha=2.0$ respectively.
+
+\begin{figure}
+\begin{subfigure}{.49\textwidth}
+	\centering
+	\includegraphics[width=\linewidth]{../figures/5e/5e_05-10-var.pdf}
+	\caption{}
+	\label{fig:transbefore1}
+\end{subfigure}
+\begin{subfigure}{.49\textwidth}
+	\centering
+	\includegraphics[width=\linewidth]{../figures/5e/5e_05-20-var.pdf}
+	\caption{}
+	\label{fig:transbefore2}
+\end{subfigure}
+\caption{Wealth distributions when weighing of previous transactions, in addition, has been taken into account.
+Figure \ref{fig:transbefore1} weighs wealth similarity less strongly ($\alpha=1$) than figure \ref{fig:transbefore2} ($\alpha=2.0$).}
+\label{fig:transbefore}
+\end{figure}
+
+Lastly, figure \ref{fig:transbeforefit} show a power law fitting of the wealth distributions first presented in figure \ref{fig:transbefore}.
+The figure both includes ($\lambda=0.5$) and excludes savings ($\lambda=0$).
+The plot is logarithmic and the Pareto power distribution is therefore presented as a straight line.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.6\textwidth]{../figures/5e/5e_var-2-var-POWERlog_FIXXED.png}
+	\caption{Power law fitting of wealth distribution including savings, nearest neighbhour- and previous transaction effects.
+The plot is logarithim and both includes ($\alpha=0$) and excludes ($\alpha=0.5$) savings.}
+	\label{fig:transbeforefit}
 \end{figure}
@@
 
-\section{Parametrization from Patriarcal et al.}
+\section{Parametrization from Patriarca et al.}
 \label{app:partriarca}

### 2016-12-09 Project5/article/article.tex
Discussion ok

@@
 \subsection{Economic inequality and social friction}
-``The first historical series of income distribution statistics became available with the publication in 1953 fo Kuznet's monumental Shares of Upper Income Groups in Income and Savings.
-Kuznet's series dealt with only one country (the United Stetes) over a period of thirty-five years (1913-1948)" (Pikkety \& Ganser, 2014, Kindle locations 276-280\cite{Piketty}).
+\say{The first historical series of income distribution statistics became available with the publication in 1953 of Kuznet's monumental Shares of Upper Income Groups in Income and Savings.
+Kuznet's series dealt with only one country (the United Stetes) over a period of thirty-five years (1913-1948)} (Piketty \& Goldhammer, 2014, Kindle locations 276-280\cite{Piketty}).
 Ever since that time there has been wealth and income inequality to a greater or lesser degree.
@@
 	\centering
-	\includegraphics[width=0.6\textwidth]{../figures/5e/5e_var-2-var-POWERlog_FIXXED.png}
+	\includegraphics[width=0.7\textwidth]{../figures/5e/5e_var-2-var-POWERlog_FIXXED.png}
 	\caption{Power law fitting of wealth distribution including savings, nearest neighbhour- and previous transaction effects.
@@
 \section{Discussion}
+Most part of this study was a replications of the work done by Patriaraca et al.
+(2004)\cite{Patriarca} and Goswami and Sen (2014)\cite{GoswamiSen} and we have more than sufficiently managed to do so, but the path to that point has regrettably involved a bit of guesswork.
+We found that both articles were somewhat lacking in procedural and algorithmic descriptions, which we find not to be consistent with sound science and the scientific method\footnote{We went as far as to e-mail some of the co-authors of the articles, asking (very nicely!) for descriptions of algorithms, software and methods.
+At the time of writing we are still without answer.}.
+\say{[Just] as other information should be available to those who want to learn and understand, program source code is the only means for programmers to learn the art from their predecessors.
+It would be unthinkable for playwrights not to allow other playwrights to read their plays (...)} (Erik Naggum)\cite{Rendering}.
+That being said, we have confidence that our own results are very much comparable to both of of the studies that this study is based upon.
+
+\subsection{The ``savings rate'' and other quirks}
+In this study we call the factor $\lambda \in [0,0.9]$ the savings rate.
+It makes sense to employ it to the model described herein, as one would not assume that anyone would spend (potentially) their entire net worth on a single transaction.
+On the other hand, a savings rate would usually be defined as the fraction of income that an individual saves over a certain amount of time, e.g.
+a year.
+The term ``savings rate'' as it is used in this model can therefore lead to some confusion, mostly because there is no concept of how time progresses throughout the simulation.
+The only thing that happens is a series of transactions, and after a steady state is reached we have a snapshot of the monetary distribution, but no feel for the dynamics of the economy.
+
+\subsection{Future developments}
+
+We found, after delving into the data, that the Parato distributions than follow the higher tails of the simulation distribution data best are best described by the following rule
+\begin{equation}
+\label{eq:ourpareto}
+w_m \propto m^{-1-\lambda-\alpha-\gamma}
+\end{equation}
+where, of course, $\lambda$ is the savings rate, $\alpha$ is the importance of wealth similarity and $\gamma$ is the importance of previous transactions.
+This is a very neat result, but we have only confirmed that the Pareto distributions fits the most-wealth in a visual manner.
+It would be good to confirm the explanatory power by some sort of regression analysis or a goodness-of-fit test.
+This was regrettably beyond the scope of our study.
+
+A very important part of the social sciences of which economics is a part is that they have applications for the real world and help in a way to explain the behaviour of human beings.
+It would therefore be nice to compare the results of these simulations to real wealth distributions.
+As Piketty points out\cite{Piketty}, data on wealth and income distributions have been available in the US since 1953.
+One study\cite{Kjersti}, in the Norwegian business magazine Kaptial uses the magazine's study of the 400 richest Norwegians to show that their wealth are indeed Pareto distributed.
+
+Norway is a special case because it is one of very few OECD\footnote{Organisation for Economic Co-operation and Development.
+Colloquially defined as a \say{club for rich countries}.} member countries that with a wealth tax.
+These include Switzerland, Hungary, Spain and France\cite{OECD}.
+In Norway, the wealth tax has a tremendous impact on high net worth individuals, making up $57.1\%$ of the total tax burden of the 400 richest individuals\cite{NHHmaster}.
+It would therefore be beneficial to include tax effects in the model in some way, as different tax regimes may have an impact of the monetary distribution.
+
+We started the the theory section by stating a macroeconomic identity (equation \ref{eq:YCGI}).
+In it we included government spending, which necessarily is very important to an economy.
+Depending on the relative size of the government, the spendings of the government could impact wealth distribution through welfare.
+Even though poor individuals may not have a high personal wealth it may not be necessary for them, because of the security that a welfare state provides.
+This would in turn affect savings rate in an economy.
+A huge ``social security net'' could lead to lower savings and vice versa.
+
+In a modern, open economy one inevitably must trade with other economies.
+The macroeconomic identity in equation \ref{eq:YCGI} can be expanded to net exports $X-M$, where $X$ is exports and $M$ is imports
+\begin{equation}
+Y = C + G + I + (X-M).
+\end{equation}
+It would be interesting to see what opening the economy up for trade would do for wealth distribution.
+According to the Stolper-Samuelson theorem, opening an economy to trade should make unskilled workers producing traded goods in a high-skilled economy worse off as trade increases\cite{StolperSamuelson}.
+Advocates of protectionism usually use such theories to underline their argumentation.
+Building a model consisting of several ``countries'' with increasing amount of trade could be beneficial in this debate, especially as we been seen an increase in protectionist political rhetoric in recent times.
 
@@
 	\emph{Physica A: Statistical Mechanics and its Applications},
-	340(1), 334-339.
+	340(1), pp. 334-339.
 
@@
 	\emph{Physica A: Statistical Mechanics and its Applications},
-	415, 514-524.
+	415, pp. 514-524.
+
+\bibitem{Rendering} Pharr, M., \& Humphreys, G. (2010).
+	\emph{Physically Based Rendering}, 
+	Amsterdam: Elsevier.
+
+\bibitem{Kjersti} Aas, K. (2010).
+	Stor formue - flaks eller dyktighet?
+	\emph{Kapital} 18, p. 30.
+
+\bibitem{OECD} OECD (2016). 
+	\emph{OECD.stat} extract.
+	\url{stats.oecd.org}. 
+	Retrieved December 9. 2016.
+
+\bibitem{NHHmaster} Lieng, E., \& Sveen, P. B. (2016).
+	\emph{kattenivået til Norges rikeste : en analyse av skattebyrden i 2014 og formuesskattens regelendringer mellom 2006 og 2014}.
+	Master Thesis. 
+	\url{http://hdl.handle.net/11250/2403838}.
+	Retrieved December 9. 2016.
+
+\bibitem{StolperSamuelson}  Stolper, W. F., \& Samuelson, P. A. (1941).
+	Protection and real wages.
+	\emph{Economic Studies}. Oxford Journals.
+	9 (1), pp. 58-73.
 

### 2016-12-09 Project5/article/article.tex
First draft

@@
 \begin{abstract}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-In id neque elementum, accumsan ligula at, lobortis tortor.
-Duis elementum pellentesque purus, sit amet euismod diam facilisis volutpat.
-Nulla facilisi.
-Mauris quis felis ante.
-Aliquam ac velit sit amet velit porta condimentum iaculis ut quam.
-Phasellus pretium libero nec lectus placerat, ac consectetur sem faucibus.
-Maecenas dictum porta finibus.
+We build a simple agent-interaction money to model wealth distribution by taking inspiration from the field of econophysics.
+The model can account for effects of saving ($\lambda$) and agent preference to trade only with agents of similar wealth ($\alpha$) and that they have dealt with before ($\gamma$).
+We find that the upper tails of the wealth distributions can be adequately explained by simple Pareto distribution: $w_m \propto m^{-1-\lambda-\alpha-\gamma}$.
+We suggest some improvements to the model; inclusion of time propagation dynamics and a system of several economies connected through varying degrees of trade.
 \end{abstract}
@@
 \section{Introduction}
+This is a first venture into the field of econophysics for us.
+We use concepts from statistical mechanics to build a model consisting of agents allowed to conduct monetary transactions with one another.
+This study is based on the previous work of Patriarca et al.
+(2004)\cite{Patriarca} and Goswami and Sen (2014)\cite{GoswamiSen}.
+We start by giving a very brief introduction to the simplest of macroeconomic models, justifying the use of a simplification to an agent-interaction model.
+Thereafter we show how the agent model works.
+It is built up piece by piece and takes into account the effects of saving, and a preference for an agent to conduct a transaction with another agent of similar wealth and one the agent has dealt with before.
+We then give a thorough description of how the model can be implemented using Monte Carlo methods in C++ \footnote{All program source code is available online.
+Follow the github url on the title page}.
+The program is relatively straightforward to implement, except perhaps for the algorithm that breaks the transaction loop if a steady state is reached prematurely.
+After that we swiftly present all results, accounting for all the different effects.
+What follows is a somewhat lengthy discussion of the model's shortcomings and suggestions for future developments.
+We end this article with some summary remarks.
 
@@
 	\includegraphics[width=0.9\textwidth]{../figures/5ac/5c_N500_varSavings.pdf}
-	\caption{Wealth distribution for $N=500$ agents after $1000$ simulations at different savings rates $\lambda$}
+	\caption{Wealth distribution for $N=500$ agents after $1000$ simulations at different savings rates $\lambda \in \{0,0.25,0.5,0.9\}$.}
 	\label{fig:distribution}
@@
 \end{subfigure}
-\caption{Logarithmic plot of wealth distribution from figure \ref{fig:distribution} \ref{fig:logdistribution}), and the logarithm of the Gibbs measure (equation \ref{eq:gibbs}) of every data point (\ref{fig:straightline}).}
+\caption{Logarithmic plot of wealth distribution from figure \ref{fig:distribution} is shown in figure \ref{fig:logdistribution}.
+Gibbs measure (equation \ref{eq:gibbs}) of every data point from the simulations plotted with logarithmic $y$-axis in figure \ref{fig:straightline}.}
 \label{fig:logplot}
@@
 \end{subfigure}
-\caption{Parametrization of two different wealth distributions, $\lambda=0$ and $\lambda=0.9$.
+\caption{Parametrization of two different wealth distributions, $\lambda=0$ in figure \ref{fig:paramfig1} and $\lambda=0.9$ in figure \ref{fig:paramfig2}.
 Method for parametrization is found in appendix \ref{app:partriarca}.
-For $\lambda=0$ two Pareto power law distributions are also shown.}
+For $\lambda=0$ two Pareto power law fits are also shown in figure \ref{fig:paramfig1}.}
 \label{fig:paramfig}
@@
 \end{subfigure}
-\caption{The effect of different weights of propensity to trade with nearest neighbours, $\alpha$.
-Figure \ref{fig:neighbour1} is excludes saving ($\lambda = 0$), figure \ref{fig:neighbour2} includes saving ($\lambda = 0.5$).}
+\caption{The effect of different propensity to trade with nearest neighbours, $\alpha\in\{0.5,1.0,1.5,2.0\}$.
+Figure \ref{fig:neighbour1} excludes saving, $\lambda = 0$, figure \ref{fig:neighbour2} includes saving, $\lambda = 0.5$.}
 \label{fig:neighbour}
@@
 \end{subfigure}
-\caption{Power law fittings of neighbourhood effect wealth distributions. Figure \ref{fig:neighbourpareto1} is a linear plot, while figure \ref{fig:neighourpareto2} is double-logarithmic.}
+\caption{Pareto power law fits of wealth distributions including propensity to trade with nearest neighbour, $\alpha\in\{0.5,1.0,1.5,2.0\}$.
+Figure \ref{fig:neighbourpareto1} is a linear plot, while figure \ref{fig:neighourpareto2} is double-logarithmic.}
 \label{fig:neighbourpareto}
@@
 \end{subfigure}
-\caption{Wealth distributions when weighing of previous transactions, in addition, has been taken into account.
-Figure \ref{fig:transbefore1} weighs wealth similarity less strongly ($\alpha=1$) than figure \ref{fig:transbefore2} ($\alpha=2.0$).}
+\caption{Wealth distributions when weighing of previous transactions has been taken into account, $\gamma\in\{0.0, 1.0, 2.0, 3.0, 4.0\}$.
+Figure \ref{fig:transbefore1} wealth similarity is weighed less strongly ($\alpha=1$) than in figure \ref{fig:transbefore2} ($\alpha=2.0$).}
 \label{fig:transbefore}
@@
 
-\subsection{The ``savings rate'' and other quirks}
+\subsection{The ``savings rate'' and time dynamics}
 In this study we call the factor $\lambda \in [0,0.9]$ the savings rate.
@@
 The term ``savings rate'' as it is used in this model can therefore lead to some confusion, mostly because there is no concept of how time progresses throughout the simulation.
-The only thing that happens is a series of transactions, and after a steady state is reached we have a snapshot of the monetary distribution, but no feel for the dynamics of the economy.
+The only thing that happens is a series of transactions, and after a steady state is reached we have a snapshot of the monetary distribution, but no feel for the dynamics of the economy and how the situation may propagate through time.
+Such time dynamics functionality could be useful to see the time it takes for an economic of monetary policy to take effect.
 
@@
 \section{Summary Remarks}
+In this study we have used two previous academic articles\cite{Patriarca}\cite{GoswamiSen} as a basis to build a simple model that mimics monetary distribution in a simple economy consisting of agents that are allowed to trade with one another.
+The model is implemented in C++ employing Monte Carlo methods.
+The model is gradually expanded upon to include savings and social rigidities like wealth similarity bias and former transaction bias.
+We found that the wealth distribution seems to follow a Gibbs distribution for almost all simulations, with a very few number of high net worth individuals.
+The only situations where the simulations resulted in a wealth distribution with a relative amount of equality is with a very high savings rate($\lambda=0.9$).
+We also found the upper tail of the wealth distribution to be close to a Pareto distribution as has been the case since 1897\cite{Pareto}.
+This is confirmed visually, but remains to be statistically tested.
+Lastly, we suggest some possible future additions to the model, most importantly having several economies at once in a simulation with different amounts of trade, and including a some sort of time dynamics in the model.
 
@@
 	340(1), pp. 334-339.
+
+\bibitem{GoswamiSen} Goswami, S., \& Sen, P. (2014).
+	Agent based models for wealth distribution with preference in interaction.
+	\emph{Physica A: Statistical Mechanics and its Applications},
+	415, pp. 514-524.
 
@@
 	Retrieved December 8. 2016.
-
-\bibitem{GoswamiSen} Goswami, S., \& Sen, P. (2014).
-	Agent based models for wealth distribution with preference in interaction.
-	\emph{Physica A: Statistical Mechanics and its Applications},
-	415, pp. 514-524.
 
@@
 
+\vfill
+\pagebreak
+
 \begin{appendices}

### 2016-12-09 Project5/article/article.tex
Second draft

@@
 \begin{abstract}
-We build a simple agent-interaction money to model wealth distribution by taking inspiration from the field of econophysics.
-The model can account for effects of saving ($\lambda$) and agent preference to trade only with agents of similar wealth ($\alpha$) and that they have dealt with before ($\gamma$).
-We find that the upper tails of the wealth distributions can be adequately explained by simple Pareto distribution: $w_m \propto m^{-1-\lambda-\alpha-\gamma}$.
-We suggest some improvements to the model; inclusion of time propagation dynamics and a system of several economies connected through varying degrees of trade.
+We build a simple agent-interaction model in order to simulate wealth distribution by taking inspiration from the field of econophysics.
+The model can account for effects of saving ($\lambda$) and agent preference to trade only with agents of similar wealth ($\alpha$) and agents they have dealt with before ($\gamma$).
+We find that the upper tails of the wealth distributions can only be adequately explained by simple Pareto distribution: $w_m \propto m^{-1-\lambda-\alpha-\gamma}$.
+We suggest some prospective future additions to the model; inclusion of time propagation dynamics, a system of several economies connected through varying degrees of trade, and functionality that takes welfare and social benefits into account.
 \end{abstract}
@@
 Thereafter we show how the agent model works.
-It is built up piece by piece and takes into account the effects of saving, and a preference for an agent to conduct a transaction with another agent of similar wealth and one the agent has dealt with before.
+It is built up piece by piece and takes into account the effects of saving, and the preference for an agent to conduct a transaction with another agent of similar wealth and one the agent has dealt with before.
 We then give a thorough description of how the model can be implemented using Monte Carlo methods in C++ \footnote{All program source code is available online.
@@
 The program is relatively straightforward to implement, except perhaps for the algorithm that breaks the transaction loop if a steady state is reached prematurely.
+This functionality speeds up the simulations tremendously.
 After that we swiftly present all results, accounting for all the different effects.
@@
 where $Y$ is income, $C$ is consumption, $G$ is government spending and $I$ is investment.
-The best-know, but not necessarily the best, measure of income $Y$ is the Gross Domestic Product (GDP)\footnote{GDP = GNP (Gross National Product) in autarky.}.
+The best-known, but not necessarily the best, measure of income $Y$ is the Gross Domestic Product (GDP)\footnote{GDP = GNP (Gross National Product) in autarky.}.
 Consumption $C$ is the monetary value of all goods and services purchased in the private sector, while government spending $G$ is the consumption of the government.
@@
 \subsection{Monetary transactions}
-To simulate monetary transactions in our model economy we expand employ the framework introduced in Patriarca et al.\cite{Patriarca}.
+To simulate monetary transactions in our model economy we employ the framework introduced in Patriarca et al.\cite{Patriarca}.
 We assume there are $N$ agents that exchange money in pairs ($i,j$).
 We assume also that all agents start with the same amount of money $m_0>0$.
-For every period an arbitrary pair of agents are picked at randomly and let them conduct business, id est a transaction takes place between them.
+For every period an arbitrary pair of agents are picked at randomly and we let them conduct business, i.e.
+a transaction takes place between them.
 Money is conserved during the transaction such that
@@
 \label{eq:wealthconserved}
-m_i + m_j = m_i' + m_j',
+m_i' + m_j'= m_i + m_j,
 \end{equation}
-where the right-hand side is the agent $i$ and $j$'s updated wealth and the left-hand side represents the amount of money agents $i$ and $j$ had before the transaction.
+where the left-hand side is the agent $i$ and $j$'s updated wealth and the right-hand side represents the amount of money agents $i$ and $j$ had before the transaction.
 The exchange is done via a random reassignment factor $\epsilon$, such that
@@
 \subsection{Pareto distributions}
-How the income and wealth of a population has been studied for a long time.
-Already in 1897, Vilfredo Pareto, sought to describe bot wealth and income distribution by power law probability distributions\cite{Pareto}.
+How the income and wealth of a population is distributed has been studied for a long time.
+Already in 1897, Vilfredo Pareto, sought to describe both wealth and income distribution by power law probability distributions\cite{Pareto}.
 Pareto showed that the higher end of the distribution of money follows a distribution
@@
 \label{eq:paretodist}
-w_m \propto m^{-1-\alpha}, \quad \alpha \in [1,2].
+w_m \propto m^{-1-\nu}, \quad \nu \in [1,2].
 \end{equation}
-Pareto's ``80-20 rule", that $20\%$ of the population controls $80\%$ of the wealth, corresponds to a certain value of $\alpha$.
+Pareto's ``80-20 rule", that $20\%$ of the population controls $80\%$ of the wealth, corresponds to a certain value of $\nu$.
 Today, a whole class of distributions similar to the one in equation \ref{eq:paretodist} are named after Pareto\cite{Arnold}.
@@
 We are can expand upon the model by introducing a savings rate $\lambda$.
-The savings rate is defined as a fraction of an agent's wealth that does not partake in a transaction for every period.
-One can gather from the macroeconomic identity in equation \ref{eq:YCGI} that income must still be the same and the transaction law in equation \ref{eq:wealthconserved} still holds.
+The savings rate is defined as a fraction of an agent's wealth that does not partake in a transaction for each cycle.
+One can gather from the macroeconomic identity in equation \ref{eq:YCGI} that income must still be the same and the transaction conservation law in equation \ref{eq:wealthconserved} still holds.
 The updated wealth of agents $i$ and $j$ after a transaction becomes
@@
 \section{Algorithm \& Implementation}
-All programming code used in this project is written in C++ and is publicly available at github by following the link on the first page.
+All programming code used in this project is written in C++ and is publicly available at github by following the link on the title page.
 The program consists mainly of two files, \lstinline|main.cpp| and \lstinline|functions.cpp|.
-\lstinline|main.cpp| contains the set-up of the program and is where the number of agents, initial wealth, number of transactions, number of simulations and all parameters ($\lambda$, $\alpha$ and $\gamma$) are specified.
+\lstinline|main.cpp| contains the set-up of the program and is where the number of agents, initial wealth, number of transactions, number of simulations and all parameters (like $\lambda$, $\alpha$ and $\gamma$) are specified.
 The main part of the program however, is in \lstinline|functions.cpp| which in addition to the \lstinline|trade(...)|\footnote{The functions have several arguments: \lstinline|trade(int N, int no_of_transactions, arma::vec (&agents), double lambda, double alpha, double gamma)|.
@@
 \subsection{The \lstinline|trade(...)| function}
-This function runs a specified number of Monte Carlo cycles, where the transactions take place.
+This function runs a specified number of Monte Carlo cycles, where the actual transactions take place.
 For each iteration two random agents are picked, then some logic follows which will determine if the two agents are allowed to go into business.
@@
 One can see that the the argument of the while loop in the program listing above contains equation \ref{eq:rigidprob} which is compared against a random number and an equality test that will return \lstinline|True| if agents $i$ and $j$ are the same agent.
-This part of the code makes sure that a new pair of randomly picked agents until we are satisfied that the agents are not the same and that they are close enough in wealth and in previous relations to go forward with a transaction.
+This part of the code makes sure that a new pair of agents are randomly picked until we are satisfied that the agents are not the same and that they are close enough in wealth and have good enough relations to go forward with a transaction.
 After everything is all-right and we have a good pair of agents a transaction takes place:
@@
 // Transaction takes place, but some is saved (lambda)
-double delta_m = (1 - lambda) * 
+double delta_m = (1 - lambda)* 
 	(epsilon*m_j - (1 - epsilon)*m_i);
@@
 
-The last part of the \lstinline|trade(...)| function is not particularly important for the quality of the data that a simulation will provide, but vastly increased the speed of the program.
+The last part of the \lstinline|trade(...)| function is not particularly important for the quality of the data that a simulation will provide, but vastly increase the speed of the program.
 This last functionality breaks the for loop within the \lstinline|trade(...)| function when a steady state is reached.
@@
 	cumVar2Block = 0;
-}
 \end{lstlisting}
@@
 If both these logical statements evaluate to \lstinline|True|, the for loop of the \lstinline|trade(...)| function is broken, terminating the current simulation.
-We found that for most simulations an equilibrium is reached quite quickly, usually between $40000$ and $120000$ transactions have taken place between $N=1000$ agents.
+We found that for most simulations an equilibrium is reached quite quickly, usually when between $40000$ and $120000$ transactions have taken place between $N=1000$ agents.
 This speeds up the time for a simulation to finish tremendously, compared with the initially, and rather naively, picked $10^7$ iterations.
-In order to find correct values \lstinline|blockSize| as well as when the variance measures are ``low enough'' the behaviour of the variance measures were meticulously studied through several trial runs of the function.
+In order to find correct values for \lstinline|blockSize| as well as when the variance measures are ``low enough'' the behaviour of the variance measures were meticulously studied through several trial runs of the function.
 
@@
 This plot, were it not for the colors, could be an exact copy of figure 1 in Patriarca et al.\cite{Patriarca}.
-Without savings the distribution appears as a inverse exponential function ($e^{-x}$) which is the same class of function as the Gibbs distribution (equation \ref{eq:gibbs}).
+Without savings the distribution appears as a inverse exponential function which is the same class of function as the Gibbs distribution (equation \ref{eq:gibbs}), and the Pareto distribution (equation \ref{eq:paretodist}).
 However, these traits disappear from the wealth distributions once savings are allowed.
@@
 Observe that all all the wealth distributions give straight lines.
-This makes sense, as the logarithm of the Gibbs distribution yields a first order expression (equation \ref{eq:lingibbs}).
+This makes sense, as the logarithm of the Gibbs distribution yields a first degree polynomial (equation \ref{eq:lingibbs}).
 
@@
 
-Figure \ref{fig:paramfig1} shows two different Pareto distributions with $\alpha=1.8$ and $\alpha = 2$ (equation \ref{eq:paretodist}) plotted against a distribution at $\lambda = 0$.
+Figure \ref{fig:paramfig1} shows two different Pareto distributions with $\nu=1.8$ and $\nu = 2$ (equation \ref{eq:paretodist}) in the same plot as a simulated wealth distribution where $\lambda = 0$ (no savings).
 We see that they fit adequately with the higher end of the distribution, but not perfectly.
 The simple Pareto distribution becomes a straight line because of the logarithmic axes in the plot.
-The same figure also contains a parametrization of the wealth \footnote{This is done in the way Partriarca et al.
+The same figure also contains a parametrization of the wealth distribution \footnote{This is done in the way Partriarca et al.
 (2004)\cite{Patriarca} does.
@@
 The parametrisations are indicated by a label $P(x)$.
+We see that the parametrisation provides a very good fit.
+The Pareto distribution only works for the right tail of the distribution and only to a certain extent.
 
@@
 Figure \ref{fig:neighbour} shows logarithmic plots of simulations for different values of $\alpha$ from equation \ref{eq:rigidprob}.
-The parameter for previous transactions $gamma$ is set to zero for now.
-The only difference between the subfigures in figure \ref{fig:neighbour} is that figure \ref{fig:neighbour1} is for $\lambda=0$ and figure \ref{fig:neighbour2} is for $\lambda=0.5$, colloquially with and without saving respectively.
-Figure \ref{fig:neighbour} shows similar results to that of Goswarmi and Sen (2014)\cite{GoswamiSen}.
+The parameter for previous transactions $\gamma$ is set to zero for now.
+The only difference between the subfigures in figure \ref{fig:neighbour} is that in figure \ref{fig:neighbour1} $\lambda=0$ and in figure \ref{fig:neighbour2} $\lambda=0.5$, colloquially with and without saving respectively.
+Figure \ref{fig:neighbour} shows similar results to figure 5 in Goswarmi and Sen (2014)\cite{GoswamiSen}.
 
@@
 Figure \ref{fig:neighbourpareto} shows the simulations for nearest neighbour interaction fitted to a Pareto distribution.
-The Pareto distribution should fit only for higher values of the wealth distribution, id est the tails of the distribution.
+The Pareto distribution should fit only for higher values of wealth, i.e.
+the tails of the distribution.
 Figure \ref{fig:neighbourpareto1} show that the power law fittings (Pareto distributions) appear to have similar shapes to the results from the simulations.
-One can see, however, that the data from the simulations tend drop quicker than their Pareto power law counterparts.
+One can see, however, that the data from the simulations tend to drop quicker than their Pareto power law counterparts.
 Figure \ref{fig:neighourpareto2} shows a logarithmic version of figure \ref{fig:neighbourpareto1}.
-Here the Pareto fittings are straight lines, as one would assume, but the data from the simulations apparently inhibit more complexity than the a simple power law relationship between wealth and frequency.
+Here the Pareto fittings are straight lines, as one would assume, but the data from the simulations apparently inhibit more complexity than the a simple power law relationship between wealth and frequency, but the fit is less horrible than it could have been.
 
@@
 Both subfigures show the wealth distribution for different values of $\gamma$ (propensity to trade with agents you have traded with before), but figure \ref{fig:transbefore1} weighs neighbour interactions less strongly than figure \ref{fig:transbefore2}, with $\alpha=1.0$ and $\alpha=2.0$ respectively.
+Again, the Pareto distribution fit is okay, but not tremendously good.
+One can see that it fits best for the very tip of the tail.
 
@@
 \subsection{The ``savings rate'' and time dynamics}
-In this study we call the factor $\lambda \in [0,0.9]$ the savings rate.
+In this study we call the factor $\lambda \in [0,1]$ the savings rate.
 It makes sense to employ it to the model described herein, as one would not assume that anyone would spend (potentially) their entire net worth on a single transaction.
-On the other hand, a savings rate would usually be defined as the fraction of income that an individual saves over a certain amount of time, e.g.
-a year.
+On the other hand, a savings rate would usually be defined as the fraction of income that an individual saves over a certain amount of time, for example over a year.
 The term ``savings rate'' as it is used in this model can therefore lead to some confusion, mostly because there is no concept of how time progresses throughout the simulation.
-The only thing that happens is a series of transactions, and after a steady state is reached we have a snapshot of the monetary distribution, but no feel for the dynamics of the economy and how the situation may propagate through time.
-Such time dynamics functionality could be useful to see the time it takes for an economic of monetary policy to take effect.
+The only thing that happens is a series of transactions which could have happened during any time span, and after a steady state is reached we have a snapshot of the monetary distribution, but no feel for the dynamics of the economy and how the situation may propagate through time.
+Such time dynamics functionality could be useful to see the time it takes for an economic of monetary policy to take effect, for instance.
 
@@
 
-We found, after delving into the data, that the Parato distributions than follow the higher tails of the simulation distribution data best are best described by the following rule
+We found, after delving into the data, that the Parato distributions that fit the higher tails of the simulation distribution data best can be described by the following rule
 \begin{equation}
@@
 where, of course, $\lambda$ is the savings rate, $\alpha$ is the importance of wealth similarity and $\gamma$ is the importance of previous transactions.
-This is a very neat result, but we have only confirmed that the Pareto distributions fits the most-wealth in a visual manner.
+This is a very neat result, but we have only confirmed that the Pareto distributions fits, and only adequately at the, in a visual manner.
 It would be good to confirm the explanatory power by some sort of regression analysis or a goodness-of-fit test.
@@
 
-A very important part of the social sciences of which economics is a part is that they have applications for the real world and help in a way to explain the behaviour of human beings.
+A very important part of the social sciences, of which economics, is to try to have applications for the real world and help in a way to explain the behaviour of human beings.
 It would therefore be nice to compare the results of these simulations to real wealth distributions.
 As Piketty points out\cite{Piketty}, data on wealth and income distributions have been available in the US since 1953.
-One study\cite{Kjersti}, in the Norwegian business magazine Kaptial uses the magazine's study of the 400 richest Norwegians to show that their wealth are indeed Pareto distributed.
+One study\cite{Kjersti}, in the Norwegian business magazine Kaptial uses the magazine's study of the 400 richest Norwegians to show that their wealth are indeed Pareto distributed, but what about the rest?
+We have simulated the distribution of the entire population here!
 
 Norway is a special case because it is one of very few OECD\footnote{Organisation for Economic Co-operation and Development.
-Colloquially defined as a \say{club for rich countries}.} member countries that with a wealth tax.
+Colloquially defined as a \say{club for rich countries}.} member countries that still have a wealth tax.
 These include Switzerland, Hungary, Spain and France\cite{OECD}.
 In Norway, the wealth tax has a tremendous impact on high net worth individuals, making up $57.1\%$ of the total tax burden of the 400 richest individuals\cite{NHHmaster}.
-It would therefore be beneficial to include tax effects in the model in some way, as different tax regimes may have an impact of the monetary distribution.
+It would therefore be beneficial to include tax effects in the model in some way, as different tax regimes will almost definitely have an impact of the monetary distribution.
 
@@
 Advocates of protectionism usually use such theories to underline their argumentation.
-Building a model consisting of several ``countries'' with increasing amount of trade could be beneficial in this debate, especially as we been seen an increase in protectionist political rhetoric in recent times.
+Building a model consisting of several ``countries'' with increasing amount of trade could be beneficial in this debate, especially as we have seen an increase in protectionist political rhetoric in recent times.
 
@@
 \section{Summary Remarks}
-In this study we have used two previous academic articles\cite{Patriarca}\cite{GoswamiSen} as a basis to build a simple model that mimics monetary distribution in a simple economy consisting of agents that are allowed to trade with one another.
+In this study we have used two academic articles in the field of econophysics\cite{Patriarca}\cite{GoswamiSen} as a basis to build a simple model that mimics monetary distribution in a simple economy consisting of agents that are allowed to make monetary transactions with one another.
 The model is implemented in C++ employing Monte Carlo methods.
 The model is gradually expanded upon to include savings and social rigidities like wealth similarity bias and former transaction bias.
-We found that the wealth distribution seems to follow a Gibbs distribution for almost all simulations, with a very few number of high net worth individuals.
+We found that the wealth distribution seems to follow something similar to a Gibbs distribution for almost all simulations, with a very few number of high net worth individuals.
 The only situations where the simulations resulted in a wealth distribution with a relative amount of equality is with a very high savings rate($\lambda=0.9$).
-We also found the upper tail of the wealth distribution to be close to a Pareto distribution as has been the case since 1897\cite{Pareto}.
-This is confirmed visually, but remains to be statistically tested.
-Lastly, we suggest some possible future additions to the model, most importantly having several economies at once in a simulation with different amounts of trade, and including a some sort of time dynamics in the model.
+We also found the upper tail of the wealth distribution to be close to, but not perfectly fitting, a Pareto distribution as has been the case since 1897\cite{Pareto}.
+This is studied visually, but remains to be statistically tested.
+Lastly, we suggest some possible future additions to the model, most importantly having several economies at once in a simulation with different amounts of trade, including a some sort of time dynamics in the model, and adding functionality that mimics a government that can provide welfare and social benefits.
 

### 2018-09-13 Project2/article/article.tex
typos

@@
 By once again introducing a dimensionless variable $\rho=r/\alpha$ with $\alpha=\hbar^2/(m\beta e^2)$.
-By, in addition, including the relative angular frequency $\omega_r^2=(\frac{2\omega}{2\hbar}\alpha^2)^2$ and an energy parameter $\lambda=\frac{m\alpha^2}{\hbar^2}E_r$, we arrive at an equation comparable to \ref{eq:schr3}
+By, in addition, including the relative angular frequency $\omega_r^2=(\frac{2\omega}{2\hbar}\alpha^2)^2$ and an energy parameter $\lambda=\frac{m\alpha^2}{\hbar^2}E_r$, we arrive at an equation comparable to \ref{eq:schr2}
 \begin{equation}
--\frac{d\psi(\rho)}{d\rho^2}+\left(\omega^2_r\rho^2 + \frac{1}{\rho} \right) \psi(\rho)= \lambda\psi(\rho)
+-\frac{d^2\psi(\rho)}{d\rho^2}+\left(\omega^2_r\rho^2 + \frac{1}{\rho} \right) \psi(\rho)= \lambda\psi(\rho)
 \end{equation}
@@
 This proof is important to the algorithm which is described in the next section.
-As a result of Abel and Ruffini's theorem, regarding the impossibility of a solution to equations degree five and higher, a discretized problema s described herein is unsolvable by conventional, analytical means, and an iterative process is thus necessary.
+As a result of Abel and Ruffini's theorem, regarding the impossibility of a solution to equations degree five and higher, a discretized problem as described herein is unsolvable by conventional, analytical means, and an iterative process is thus necessary.
 
@@
 Notice that this matrix is symmetric as well.
-We want to roate in such a manner that the offdiagonal elemtes of the new matrix are zero.
+We want to roate in such a manner that the off-diagonal elements of the new matrix are zero.
 Hence we need to solve $(a_{22}-a_{11})cs + a_{12}(c^2-s^2)=0$, which gives
@@
 \end{equation}
-which has two roots. Of these two roots we pickk the smallest in order to ensure maximum  numerical stability. To put it another way, we rotate a tiny angle rather than a large one at each iteration.
+which has two roots. Of these two roots we pick the smallest in order to ensure maximum  numerical stability. To put it another way, we rotate a tiny angle rather than a large one at each iteration.
 
@@
 
-The \texttt{jacobiRotation} function works by calculating equation \ref{eq:rotation} straightforward and updating all matrix elements. Lastly the eigenvectors are saved to the matrix R.
+The \texttt{jacobiRotation} function works by calculating equations \ref{eq:rotation} straightforward and updating all matrix elements. Lastly the eigenvectors are saved to the matrix R.
 

### 2018-09-13 Project2/julia/simple_jacobi.jl
Jacobi's method in Julia
	I just wrote the worst eigenvalue solver in Julia.
	Results are quite wrong.

@@ -1,0 +1,55 @@
+# I am at it again!
+#
+# This is a Julia program where I try to diagonalize a matrix
+# using Jacobi's method.
+#
+# Greg Winther
+
+import Base
+
+ITERATION_LIMIT = 10
+
+# We have a matrix
+A = Array{Float64}(
+    [10 -1 2 0;
+     -1 11 -1 3;
+     2 -1 01 -1;
+     0 3 -1 8])
+   
+# RHS vector
+b = Array{Float64}(
+    [6 25 -11 15])
+
+x = zeros(size(b))
+
+for iteration = 1:ITERATION_LIMIT
+
+    global x
+    
+    println("Current solution: ", x)
+    xNew = zeros(size(x))
+
+    # println("Iteration = ", iteration)
+
+    for i = 1:size(A)[1]
+        s1 = *(transpose(A[i, 1:i-1]), x[1:i-1])
+        # println(A[i, 1:i], "*", x[1:i], "=", s1)
+        s2 = *(transpose(A[i, i+1:end]), x[i+1:end])
+        # println(A[i, i+1:end], "*", x[i+1:end], "=", s2)
+        xNew[i] = (b[i] - s1 - s2) / A[i, i]
+    end
+       
+    if isapprox(x, xNew, atol=1e-10)
+        break
+    end
+
+    x = xNew
+    #println()
+
+end
+
+println("Solution: ")
+println(x)
+error = *(A, transpose(x)) - transpose(b)
+println("Error: ")
+println(error)

### 2018-10-13 Project1/article/article.tex
Typo
Removed numbering of dots (not really equations)

@@
 \begin{array}{cccc|c}
-1 & c_1/b_1 & 0 & 0 & b_1/b_1 \\
+1 & c_1/b_1 & 0 & 0 & f_1/b_1 \\
 0 & b_2-\frac{c_1}{b_1}a_2 & c_2 & 0  & f_2-\frac{f_1}{b_1}a_2 \\
@@
 l_{31}w_1 + l_{32}w_2 +w_3 &= f_3 \\
- & \vdots \\
+ & \vdots \nonumber \\
 l_{n-11}w_1 + l_{n-12}w_2 + l_{n-13}w_3 + \dots + w_{n-1}  &=
@@
 u_{33}v_3 + \dots+u_{3n-1}v_{n-1} + u_{3n}v_n &= w_3 \\
- & \vdots \\
+ & \vdots \nonumber \\
 u_{n-1n-1}v_{n-1} &=  w_{n-1} \\

### 2018-10-13 Project2/jacobiMethod.cpp
Find no of iterations for N=10. It was 153. wow.

@@ -199,6 +199,7 @@ void jacobiMethod(arma::mat &A, arma::mat &R, int n, int &minIndex) {
         jacobiRotation(A, R, k, l, n);
         max_nondiagonal = maxOffDiagonals(A, k, l, n);
         iterations++;
+        // cout << iterations << endl;
 
     }
 
